# Annotating figures and formatting axes

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

Every hvtiPlotR plot renders as a bare `ggplot` object: no labels, annotations,
or cropping are applied by the constructor. You add those by chaining `+`
layers. This chapter covers axis/title text (`labs()`), in-panel annotations
(`annotate()`), viewport cropping (`coord_cartesian()`), and draft footnotes
(`make_footnote()`).

We build one base KM plot and decorate it throughout.

```{r}
#| label: base
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)
```

## Labels with `labs()`

`labs()` sets axis labels, title, subtitle, legend title, and caption. Set them
on the rendered plot so they can be overridden per project.

```{r}
#| label: labs
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title   = "Overall Survival",
    x       = "Years after Operation",
    y       = "Freedom from Death (%)",
    caption = "Logit CI, α = 0.6827 (1 SD)"
  ) +
  theme_hv_poster()
```

## Annotations with `annotate()`

`annotate()` places text, segments, or arrows at fixed data coordinates —
a sample-size callout in a corner, or an arrow pointing to an event.

```{r}
#| label: annotate
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  annotate("text",    x = 1,  y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  annotate("segment", x = 10, xend = 10, y = 30, yend = 50,
           arrow = arrow(length = unit(0.2, "cm")), colour = "grey40") +
  annotate("text",    x = 10.3, y = 40,
           label = "Median survival", hjust = 0, size = 3, colour = "grey40") +
  theme_hv_poster()
```

## Cropping with `coord_cartesian()`

`coord_cartesian()` crops the viewport without dropping data, so any fit
computed on the full range is preserved.

```{r}
#| label: coord
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))

p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  coord_cartesian(xlim = c(1995, 2020), ylim = c(20, 70)) +
  theme_hv_poster()
```

## Draft footnotes with `make_footnote()`

`make_footnote()` stamps the **bottom-right corner** of the current graphics
device using grid — call it *after* `print()`ing a plot to flag a
work-in-progress figure. The ggplot object itself is untouched, so the final
`ggsave()` output is clean.

```
# During analysis                   # For publication
print(p)                             ggsave("fig1.pdf", p, ...)
make_footnote("R/analysis.R")        # <- no footnote call
```

Build any figure, then `print()` it and call `make_footnote()` with the source
path. The default stamp appends a timestamp.

```{r}
#| label: footnote_draft
#| fig-width: 7
#| fig-height: 5
p_draft <- plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Survival (%)") +
  theme_hv_poster()

print(p_draft)
make_footnote("annotation.qmd")
```

Pass any string as `text` and set `timestamp = FALSE` for a stable analyst tag.

```{r}
#| label: footnote_analyst
#| fig-width: 7
#| fig-height: 5
print(p_draft)
make_footnote(
  text      = paste("J. Ehrlinger |", Sys.Date()),
  timestamp = FALSE,
  prefix    = ""
)
```

For publication, write the ggplot object directly with `ggsave()` and simply
omit the `make_footnote()` call — the saved file carries no stamp.
