# Annotating figures and formatting axes

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

Every hvtiPlotR [@R-hvtiPlotR] plot renders as a bare `ggplot` object: no labels, annotations,
or cropping are applied by the constructor. You add those by chaining `+`
layers. This chapter covers axis/title text (`labs()`), in-panel annotations
(`annotate()`), viewport cropping (`coord_cartesian()`), and draft footnotes
(`make_footnote()`).

## When to use it

A bare plot shows the data correctly but says nothing about what the data mean.
The decorations in this chapter are how you turn a correct plot into one a
reader can interpret without your narration. You add labels so the axes name
real quantities, an annotation so a sample size or a landmark is on the panel
rather than in the caption, a crop so the eye lands on the region that carries
the story, and, while the work is still in progress, a draft footnote so nobody
mistakes a working figure for a final one.

Think of it as the last pass before a figure leaves your hands. The plot is
right; now make it legible. Each tool below is small, and you reach for them in
combination: a labelled, cropped curve with an `n =` callout is the everyday
case. We build one base Kaplan-Meier plot and decorate it throughout, so you can
see each layer added to the same starting point.

```{r}
#| label: base
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)
```

## Labels with `labs()`

`labs()` sets axis labels, title, subtitle, legend title, and caption. Set them
on the rendered plot so they can be overridden per project.

```{r}
#| label: labs
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title   = "Overall Survival",
    x       = "Years after Operation",
    y       = "Freedom from Death (%)",
    caption = "Logit CI, α = 0.6827 (1 SD)"
  ) +
  theme_hv_poster()
```

## Annotations with `annotate()`

`annotate()` places text, segments, or arrows at fixed data coordinates —
a sample-size callout in a corner, or an arrow pointing to an event.

```{r}
#| label: annotate
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  annotate("text",    x = 1,  y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  annotate("segment", x = 10, xend = 10, y = 30, yend = 50,
           arrow = arrow(length = unit(0.2, "cm")), colour = "grey40") +
  annotate("text",    x = 10.3, y = 40,
           label = "Median survival", hjust = 0, size = 3, colour = "grey40") +
  theme_hv_poster()
```

## Cropping with `coord_cartesian()`

`coord_cartesian()` crops the viewport without dropping data, so any fit
computed on the full range is preserved. This is the distinction that trips up
nearly everyone at some point. `coord_cartesian(xlim = ...)` is a camera: it
zooms in on a region while every point and every fitted line still knows about
the data outside the frame. `scale_x_continuous(limits = ...)` is a filter: it
removes the out-of-range data before anything is drawn, so a LOESS smooth or a
regression line is now fitted to a *subset* and can shift, sometimes
dramatically. When you only want to crop the view, reach for `coord_cartesian()`
every time.

```{r}
#| label: coord
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))

p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  coord_cartesian(xlim = c(1995, 2020), ylim = c(20, 70)) +
  theme_hv_poster()
```

## Draft footnotes with `make_footnote()`

`make_footnote()` stamps the **bottom-right corner** of the current graphics
device using grid. Call it *after* `print()`ing a plot to flag a
work-in-progress figure. The ggplot object itself is untouched, so the final
`ggsave()` output is clean.

```
# During analysis                   # For publication
print(p)                             ggsave("fig1.pdf", p, ...)
make_footnote("R/analysis.R")        # <- no footnote call
```

Build any figure, then `print()` it and call `make_footnote()` with the source
path. The default stamp appends a timestamp.

```{r}
#| label: footnote_draft
#| fig-width: 7
#| fig-height: 5
p_draft <- plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Survival (%)") +
  theme_hv_poster()

print(p_draft)
make_footnote("annotation.qmd")
```

Pass any string as `text` and set `timestamp = FALSE` for a stable analyst tag.

```{r}
#| label: footnote_analyst
#| fig-width: 7
#| fig-height: 5
print(p_draft)
make_footnote(
  text      = paste("J. Ehrlinger |", Sys.Date()),
  timestamp = FALSE,
  prefix    = ""
)
```

For publication, write the ggplot object directly with `ggsave()` and simply
omit the `make_footnote()` call. The saved file carries no stamp.

## Pitfalls

- **`coord_cartesian()` zooms, `scale_*(limits =)` drops.** This is the one to
  memorise. If you set limits on the scale to "zoom in," you have silently
  thrown away the data outside the window, and any smoother, ribbon, or summary
  refits to what is left. The curve you see is then a fit to a subset, not the
  region of a fit to everything. Use `coord_cartesian()` when you want the same
  fit, viewed closer. Use a scale limit only when you genuinely mean to exclude
  data, and say so.
- **A draft footnote left on a final figure.** `make_footnote()` stamps the
  device, not the ggplot object, which is the safe design: the stamp cannot
  survive into a `ggsave()` file unless you deliberately keep printing and
  stamping. The failure mode is human, not technical. You screenshot a working
  figure mid-analysis, the file path and timestamp ride along, and it ends up in
  a slide deck. Keep draft stamps to the analysis session and let the publication
  path be the plain `ggsave()`.
