# Covariate balance and follow-up

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## Covariate balance plots

The covariate balance plot is the standard quality-control figure for
propensity-score matching and IPTW analyses. Each covariate occupies a labelled
row; points show the standardized mean difference (SMD) for each comparison
group. A solid vertical line marks zero balance; dotted lines at ±10% give a
quick visual threshold for acceptable balance. `hv_balance()`, from hvtiPlotR
[@R-hvtiPlotR], prepares the data; `plot()` hands you a bare ggplot to style with `+`.

Input must be in **long format**: one row per covariate × group combination
with columns for the covariate name, the group label, and the numeric SMD.

### Sample data

`sample_covariate_balance_data()` returns 12 covariates in the long format
`hv_balance()` expects (`variable`, `group`, `std_diff`). Build the S3 object
once and reuse it across the styled variants below.

```{r}
#| label: cov_balance_data
dta_cb <- sample_covariate_balance_data(n_vars = 12)
head(dta_cb)

# Build the S3 object once; reuse for all plot variants below
cb <- hv_balance(dta_cb)
```

### Bare plot

The bare panel lays out one covariate per row with points at their SMD values,
but no colour, shape, axis limits, or theme yet.

```{r}
#| label: cov_balance_bare
plot(cb, alpha = 0.8)
```

### Adding colour, shape, and axis scales

Map `"Before match"` to red triangles and `"After match"` to blue squares. Set
the x-axis limits wide enough to include your largest pre-match SMD; symmetric
breaks make the ±10% threshold visually obvious.

```{r}
#| label: cov_balance_scales
plot(cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(
    limits = c(-45, 35),
    breaks = seq(-40, 30, 10)
  ) +
  labs(x = "Standardized difference (%)", y = "") +
  theme(legend.position = c(0.20, 0.95))
```

### Adding directional annotations and theme

`annotate()` places explanatory text on the panel to indicate which direction
of imbalance favours each group, using inside-panel coordinates and
`theme_hv_poster()`.

```{r}
#| label: cov_balance_annotated
n_vars <- length(unique(dta_cb$variable))

cb_final <- plot(cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(
    limits = c(-40, 30),
    breaks = seq(-40, 30, 10)
  ) +
  labs(x = "Standardized difference: SAVR - TF:TAVR (%)", y = "") +
  annotate("text", x = -30, y = 0,      label = "More likely TF-TAVR", size = 4.5) +
  annotate("text", x =  22, y = n_vars, label = "More likely SAVR",    size = 4.5) +
  theme(legend.position = c(0.20, 0.935)) +
  theme_hv_poster()

cb_final
```

### Controlling covariate order

Pass `var_levels` to the constructor to control the bottom-to-top display order
of covariates. The example below reverses the default order.

```{r}
#| label: cov_balance_order
cb_ord <- hv_balance(dta_cb, var_levels = rev(unique(dta_cb$variable)))

plot(cb_ord, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  labs(x = "Standardized difference (%)", y = "") +
  theme_hv_poster()
```

## Goodness-of-follow-up plots

The goodness-of-follow-up plot is a standard quality-control figure in
longitudinal outcome analyses. Each patient appears as a point at their
operation date (x-axis) and follow-up duration (y-axis), with a short vertical
tick below. A dashed diagonal marks the maximum potential follow-up given the
study start, study end, and follow-up closing date — points above the line have
longer follow-up than that window alone explains.

`hv_followup()` prepares the data; `plot()` hands you a bare ggplot. The `type`
argument selects the panel: `"followup"` (default, death/censoring scatter) or
`"event"` (competing non-fatal event panel).

### Sample data

`sample_goodness_followup_data()` generates 300 patients with operation dates,
follow-up durations, vital status, and a simulated non-fatal event column. The
constructor needs `study_start`, `study_end`, and `close_date` to draw the
maximum-follow-up diagonal correctly.

```{r}
#| label: gfup_data
gfup_dta <- sample_goodness_followup_data(n = 300, seed = 42)
head(gfup_dta)
```

### Death follow-up plot

The bare `plot(gf)` panel shows each patient as a point and tick without
scales, labels, or theme.

```{r}
#| label: gfup_basic
gf <- hv_followup(
  data        = gfup_dta,
  origin_year = 1990,
  study_start = as.Date("1990-01-01"),
  study_end   = as.Date("2019-12-31"),
  close_date  = as.Date("2021-08-06")
)

# Bare plot — no scales or labels yet
plot(gf)
```

Scale, label, and annotation layers compose with `+`. `scale_color_manual()`
and `scale_shape_manual()` map the binary alive/dead state; `annotate()` places
group-identifying text on the panel; `theme_hv_manuscript()` finishes it.

```{r}
#| label: gfup_styled
gfup_final <- plot(gf, alpha = 0.8) +
  scale_color_manual(
    values   = c("#377EB8", "#E41A1C"),
    labels   = c("Alive", "Dead"),
    na.value = "black",
    drop     = FALSE
  ) +
  scale_shape_manual(
    values = c(1, 4),
    labels = c("Alive", "Dead")
  ) +
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33, 3)) +
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  labs(
    x     = "Operation Date",
    y     = "Follow-up (years)",
    color = "Status",
    shape = "Status"
  ) +
  annotate("text", x = 1993, y = 31, label = "Alive at close",
           hjust = 0, size = 3.5) +
  annotate("text", x = 1993, y = 28, label = "Deceased",
           hjust = 0, size = 3.5, color = "#E41A1C") +
  theme(legend.position = "none") +
  theme_hv_manuscript()

gfup_final
```

### Non-fatal event panel

When the dataset includes a non-fatal competing event (e.g. relapse,
reoperation), pass `event_col`, `event_time_col`, and optionally
`death_for_event_col` to the constructor, then call `plot()` with
`type = "event"`.

```{r}
#| label: gfup_event_panel
gf2 <- hv_followup(
  gfup_dta,
  event_col           = "ev_event",
  event_time_col      = "iv_event",
  death_for_event_col = "deads",
  event_levels        = c("No event", "Relapse", "Death"),
  origin_year         = 1990,
  study_start         = as.Date("1990-01-01"),
  study_end           = as.Date("2019-12-31"),
  close_date          = as.Date("2021-08-06")
)

plot(gf2, type = "event", alpha = 0.8) +
  scale_color_manual(
    values = c("No event" = "blue", "Relapse" = "green3", "Death" = "red"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("No event" = 1L, "Relapse" = 2L, "Death" = 4L),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33, 3)) +
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  labs(
    x = "Operation Date",
    y = "Follow-up (years)",
    color = "Event", shape = "Event"
  ) +
  annotate("text", x = 1993, y = 31,
           label = "Systematic follow-up", hjust = 0, size = 3.5) +
  theme(legend.position = c(0.85, 0.15)) +
  theme_hv_manuscript()
```
