# Covariate balance and follow-up

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

This chapter covers the two figures we reach for to vouch for a study before we
report a single outcome from it. The covariate balance plot says the groups we
are comparing were actually comparable. The goodness-of-follow-up plot says we
kept track of the patients long enough to believe the outcome. Neither is the
headline figure of a paper, but a reviewer who does not see them will ask, and
both belong in the same quality-control habit.

## Covariate balance plots

### When to use it

Whenever you compare two groups that were not randomized (a propensity-matched
SAVR-versus-TAVR cohort, say, or an inverse-probability-weighted analysis) the
first question is whether the comparison is fair. If the treated patients were
sicker to begin with, any difference in outcome could just be that. The
covariate balance plot is how we show, covariate by covariate, that matching or
weighting pulled the two groups together.

Each covariate is a row. A point shows the standardized mean difference (SMD,
the gap between groups in standard-deviation units, scaled to a percent) for
each comparison stage, usually before and after matching. A solid line at zero
marks perfect balance; dotted guides at plus and minus ten percent give the eye
a threshold for "close enough." `hv_balance()` [@R-hvtiPlotR] prepares the data
and `plot()` hands back a bare ggplot to style with `+`.

### The data it needs

`hv_balance()` wants **long format**: one row per covariate-by-group
combination, with a column for the covariate name, one for the group label
(`"Before match"`, `"After match"`), and one for the numeric SMD.
`sample_covariate_balance_data()` returns 12 covariates already in that shape,
with columns `variable`, `group`, and `std_diff`.

```{r}
#| label: cov_balance_data
dta_cb <- sample_covariate_balance_data(n_vars = 12)
head(dta_cb)

# Build the S3 object once; reuse for all plot variants below
cb <- hv_balance(dta_cb)
```

If your numbers arrive in wide format (one column per stage, as a summary table
or spreadsheet export usually does), reshape to long first. `reshape()` from
base R does it in one call: name the two stage columns in `varying`, give the
stacked value column a name with `v.names`, and tell it which column identifies
the covariate.

```{r}
#| label: cov_balance_wide_reshape
dta_wide <- data.frame(
  variable      = c("Age", "Female sex", "Hypertension", "Diabetes", "COPD"),
  `Before match` = c(22.1, -15.3,  18.7, -9.4,  11.2),
  `After match`  = c( 3.5,   2.1,  -1.8,  4.0,  -2.3),
  check.names = FALSE
)

dta_long <- reshape(
  dta_wide,
  direction = "long",
  varying   = c("Before match", "After match"),
  v.names   = "std_diff",
  timevar   = "group",
  times     = c("Before match", "After match"),
  idvar     = "variable"
)
head(dta_long)
```

### Build it

Start from the bare panel to see what the constructor produced. One covariate
per row, points at their SMD values, and no colour, shape, axis limits, or theme
yet.

```{r}
#| label: cov_balance_bare
plot(cb, alpha = 0.8)
```

Now layer on the house style. We map `"Before match"` to red triangles and
`"After match"` to blue squares, the colour convention this team has used for
years, and widen the x-axis enough to hold the largest pre-match SMD. Symmetric
breaks around zero make the plus-or-minus-ten-percent threshold easy to eyeball.

```{r}
#| label: cov_balance_scales
plot(cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(
    limits = c(-45, 35),
    breaks = seq(-40, 30, 10)
  ) +
  labs(x = "Standardized difference (%)", y = "") +
  theme(legend.position = c(0.20, 0.95))
```

For a finished figure, place the legend and direction-of-imbalance labels inside
the panel and switch on the poster theme. `annotate()` puts text at chosen data
coordinates, here at the far left and far right rows, so the reader knows which
direction favours which group. `n_vars` reads the covariate count off the data
so the top annotation lands on the top row no matter how many covariates you
have.

```{r}
#| label: fig-balance
#| fig-cap: "Covariate balance plot showing standardized mean differences before and after matching, with direction-of-imbalance annotations"
n_vars <- length(unique(dta_cb$variable))

cb_final <- plot(cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(
    limits = c(-40, 30),
    breaks = seq(-40, 30, 10)
  ) +
  labs(x = "Standardized difference: SAVR - TF:TAVR (%)", y = "") +
  annotate("text", x = -30, y = 0,      label = "More likely TF-TAVR", size = 4.5) +
  annotate("text", x =  22, y = n_vars, label = "More likely SAVR",    size = 4.5) +
  theme(legend.position = c(0.20, 0.935)) +
  theme_hv_poster()

cb_final
```

### Read it

Read down the rows, comparing the two points in each:

- **The after-match points should collapse toward zero.** That is the whole
  claim of the figure. Blue squares sitting inside the dotted plus-or-minus-ten
  guides, with red triangles further out, is balance achieved by matching.
- **Watch any covariate that stays imbalanced.** A blue point still well past
  ten percent is a variable matching did not fix. You either adjust for it in
  the outcome model or say so in the limitations.
- **The annotations name the direction.** A positive SMD means one group has
  more of that covariate; the panel labels tell the reader which. Without them a
  reviewer cannot tell whether "More likely SAVR" sits left or right.

### Variations

#### Controlling covariate order

Pass `var_levels` to the constructor to set the bottom-to-top order of rows.
Supply any vector containing all the covariate names; the example reverses the
default. Order covariates the way a reader scans them, often most-imbalanced at
top, rather than by data-frame order.

```{r}
#| label: fig-balance-order
#| fig-cap: "Covariate balance plot with the row order reversed to control how a reader scans the covariates"
cb_ord <- hv_balance(dta_cb, var_levels = rev(unique(dta_cb$variable)))

plot(cb_ord, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  labs(x = "Standardized difference (%)", y = "") +
  theme_hv_poster()
```

## Goodness-of-follow-up plots

### When to use it

A survival or time-to-event analysis is only as trustworthy as its follow-up. If
patients vanished from the records early, an apparent absence of events may just
be an absence of looking. The goodness-of-follow-up plot is how we show a
reviewer that the cohort was tracked across the whole study window, not just at
the start.

Each patient is a point at their operation date (x-axis) and follow-up duration
(y-axis), with a short tick below. A dashed diagonal marks the maximum follow-up
the study window alone could explain: a patient operated on early could be
followed for many years, one operated on late for only a few. Points above the
diagonal have *more* follow-up than the window explains, which happens when
passive surveillance (registry linkage, say) extends beyond the active
cross-sectional contact. `hv_followup()` [@R-hvtiPlotR] prepares the data; the
`type` argument picks the panel: `"followup"` (the default death-or-censoring
scatter) or `"event"` (a competing non-fatal event).

### The data it needs

`hv_followup()` wants one row per patient with an operation date, a follow-up
duration, and a vital-status indicator. The constructor also needs three study
dates, `study_start`, `study_end`, and `close_date`, to place the
maximum-follow-up diagonal correctly. `sample_goodness_followup_data()`
generates 300 patients in that shape, including a simulated non-fatal event
column we use later.

```{r}
#| label: gfup_data
gfup_dta <- sample_goodness_followup_data(n = 300, seed = 42)
head(gfup_dta)
```

### Build it

Build the object with the three study dates, then look at the bare panel: each
patient as a point and tick, no scales or labels yet.

```{r}
#| label: gfup_basic
gf <- hv_followup(
  data        = gfup_dta,
  origin_year = 1990,
  study_start = as.Date("1990-01-01"),
  study_end   = as.Date("2019-12-31"),
  close_date  = as.Date("2021-08-06")
)

# Bare plot — no scales or labels yet
plot(gf)
```

Now finish it. `scale_color_manual()` and `scale_shape_manual()` map the binary
alive-or-dead state to colour and point shape; `coord_cartesian()` clips the
view to the study window; `annotate()` labels the two states on the panel; and
`theme_hv_manuscript()` gives the journal-sized version.

```{r}
#| label: fig-balance-followup
#| fig-cap: "Goodness-of-follow-up plot showing each patient's operation date against follow-up duration, coloured by vital status, with the maximum-follow-up diagonal"
gfup_final <- plot(gf, alpha = 0.8) +
  scale_color_manual(
    values   = c("#377EB8", "#E41A1C"),
    labels   = c("Alive", "Dead"),
    na.value = "black",
    drop     = FALSE
  ) +
  scale_shape_manual(
    values = c(1, 4),
    labels = c("Alive", "Dead")
  ) +
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33, 3)) +
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  labs(
    x     = "Operation Date",
    y     = "Follow-up (years)",
    color = "Status",
    shape = "Status"
  ) +
  annotate("text", x = 1993, y = 31, label = "Alive at close",
           hjust = 0, size = 3.5) +
  annotate("text", x = 1993, y = 28, label = "Deceased",
           hjust = 0, size = 3.5, color = "#E41A1C") +
  theme(legend.position = "none") +
  theme_hv_manuscript()

gfup_final
```

### Read it

The shape of the cloud tells the story:

- **A solid wedge under the diagonal is good news.** Points filling the triangle
  below the line mean patients across the whole span of operation dates were
  followed about as long as the window allows. Gaps or thin regions flag eras
  where follow-up was sparse.
- **Points above the diagonal are not errors.** They mark patients whose status
  is known beyond the active follow-up window, usually through passive linkage.
  A scatter of them above the line is expected; if *every* point sits exactly on
  the line, the close date is probably set wrong.
- **The colour split shows the events.** Red points are deaths. Where they
  cluster (early after surgery, or late at long follow-up) is a hint at the
  hazard pattern you will quantify in the survival analysis.

### Variations

#### Non-fatal event panel

When the dataset carries a non-fatal competing event (relapse, reoperation),
pass `event_col`, `event_time_col`, and optionally `death_for_event_col` to the
constructor, then call `plot()` with `type = "event"`. The panel then shows the
three-way event status (no event, the non-fatal event, death) instead of the
binary alive-or-dead split.

```{r}
#| label: fig-balance-followup-event
#| fig-cap: "Goodness-of-follow-up plot showing three-way event status (no event, non-fatal event, death) for each patient"
gf2 <- hv_followup(
  gfup_dta,
  event_col           = "ev_event",
  event_time_col      = "iv_event",
  death_for_event_col = "deads",
  event_levels        = c("No event", "Relapse", "Death"),
  origin_year         = 1990,
  study_start         = as.Date("1990-01-01"),
  study_end           = as.Date("2019-12-31"),
  close_date          = as.Date("2021-08-06")
)

plot(gf2, type = "event", alpha = 0.8) +
  scale_color_manual(
    values = c("No event" = "blue", "Relapse" = "green3", "Death" = "red"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("No event" = 1L, "Relapse" = 2L, "Death" = 4L),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33, 3)) +
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  labs(
    x = "Operation Date",
    y = "Follow-up (years)",
    color = "Event", shape = "Event"
  ) +
  annotate("text", x = 1993, y = 31,
           label = "Systematic follow-up", hjust = 0, size = 3.5) +
  theme(legend.position = c(0.85, 0.15)) +
  theme_hv_manuscript()
```

## Pitfalls

- **Balance plots need long format.** One row per covariate-by-group
  combination. Hand `hv_balance()` a wide table and you get one point per
  covariate instead of the before-and-after pair; reshape first.
- **Distinguish the groups.** If the before and after points sit on top of each
  other for every covariate, check that the `group` column actually has two
  distinct levels. Identical points usually mean a labelling slip, not perfect
  matching.
- **A single threshold is a guide, not a verdict.** The ten-percent lines are a
  convention. A covariate just past them may matter clinically or may not;
  judge it against what the variable is, not the line alone.
- **The follow-up diagonal depends on the dates.** `study_start`, `study_end`,
  and `close_date` set where the line falls. Get one wrong and every point looks
  mis-placed relative to it. Confirm the three dates against the protocol before
  reading the figure.
- **Above the line is not a data error.** Points above the diagonal reflect
  passive surveillance reaching past active contact. Do not "clean" them away;
  they are part of why the follow-up is good.
