# Bar Chart Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
library(patchwork)
```

## EDA barplots

`hv_eda()` drives exploratory data analysis of any variable against a reference
time axis. `eda_classify_var()` detects whether a column is continuous,
numeric-categorical, or character-categorical; `sample_eda_data()` generates a
mixed-type dataset. `plot()` returns a bare ggplot to dress with colour scales,
labels, and a theme.

```{r}
#| label: eda_data
dta_eda <- sample_eda_data(n = 300, seed = 42)
head(dta_eda)

# Auto-detected types for each column
sapply(dta_eda, eda_classify_var)
```

### Binary categorical: count barplot

Numeric 0/1 columns are classified as `"Cat_Num"`. `NA` values appear as an
explicit `"(Missing)"` fill level. The `y_label` argument sets the title and
fill-legend name in place of the raw column name.

```{r}
#| label: eda_binary_count
plot(hv_eda(dta_eda, x_col = "year", y_col = "male",
            y_label = "Sex")) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick", "(Missing)" = "grey80"),
    labels = c("0" = "Female", "1" = "Male", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  theme_hv_manuscript()
```

### Binary categorical: percentage barplot

`show_percent = TRUE` switches `geom_bar()` to `position = "fill"`.

```{r}
#| label: eda_binary_percent
plot(hv_eda(dta_eda, x_col = "year", y_col = "cabg",
            y_label = "Concomitant CABG", show_percent = TRUE)) +
  scale_fill_manual(
    values = c("0" = "grey70", "1" = "steelblue", "(Missing)" = "grey90"),
    labels = c("0" = "No CABG", "1" = "CABG", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Surgery Year", y = "Proportion") +
  theme_hv_manuscript()
```

### Ordinal and multi-level categorical

Columns with more than two numeric levels render as stacked count bars, one
level per fill colour. A reversed diverging palette (`"RdYlGn"`) signals grade
severity from green (low) through yellow to red (high).

```{r}
#| label: eda_ordinal
plot(hv_eda(dta_eda, x_col = "year", y_col = "nyha",
            y_label = "Preoperative NYHA Class")) +
  scale_fill_brewer(
    palette = "RdYlGn", direction = -1,
    labels  = c("1" = "I", "2" = "II", "3" = "III", "4" = "IV",
                "(Missing)" = "Missing"),
    name    = "NYHA"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  theme_hv_manuscript()
```

## Longitudinal participation counts

`hv_longitudinal()` reproduces a two-panel layout: a grouped bar chart of
patients and measurements at each follow-up window, paired with a numeric
summary table below. Call `plot(lc)` for the bar chart and
`plot(lc, type = "table")` for the table panel, then compose with `patchwork`.
Input must be pre-aggregated long-format data; `sample_longitudinal_counts_data()`
derives it from a simulated registry.

```{r}
#| label: lc_data
lc_dat <- sample_longitudinal_counts_data(n_patients = 300, seed = 42L)
lc_dat

# Build the S3 object once; use for both panels
lc <- hv_longitudinal(lc_dat)
```

### Bar chart

Layer fill colours onto the bare plot to distinguish Patients from
Measurements, then expand the y-axis to leave room above the tallest bar.

```{r}
#| label: lc_bar
p_lc_bar <- plot(lc) +
  ggplot2::scale_fill_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    name   = NULL
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq(0, 2000, 500),
    expand = c(0, 0)
  ) +
  ggplot2::coord_cartesian(ylim = c(0, 2200)) +
  ggplot2::labs(x = NULL, y = "Count (n)") +
  theme_hv_manuscript() +
  ggplot2::theme(legend.position = c(0.85, 0.85))

p_lc_bar
```

### Numeric table panel

`plot(lc, type = "table")` gives the numeric summary panel — the same counts
rendered as coloured text below the x-axis labels, intended to be composed
below the bar chart.

```{r}
#| label: lc_table
p_lc_tbl <- plot(lc, type = "table") +
  ggplot2::scale_colour_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    guide  = "none"
  ) +
  theme_hv_manuscript()

p_lc_tbl
```

### Two-panel layout with patchwork

Stack the bar chart above the table panel with `/` and use
`plot_layout(heights = c(3, 1))` to give the bar chart three times the vertical
space.

```{r}
#| label: lc_combined
#| fig-height: 5
p_lc_bar / p_lc_tbl +
  patchwork::plot_layout(heights = c(3, 1))
```
