# Bar Chart Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
library(patchwork)
```

## When to use it

A bar chart answers a counting question. How many patients had a concomitant
CABG this year? What fraction arrived in NYHA class III or IV? How did the size
of each annual cohort change over the study period? Whenever the thing you want
to show is a count or a proportion of a category, and you want it broken out
across a second axis (usually time), the bar chart is the plain, honest display
to reach for. Nobody misreads a bar.

There are two related jobs in this chapter, and the package splits them across
two functions. `hv_eda()` [@R-hvtiPlotR] is the exploratory workhorse you point
at a single variable to see how its categories fall out across a reference axis;
it is the same function that draws scatter plots for continuous columns, but
here we stay with the categorical case where it returns bars. `hv_longitudinal()`
is the reporting display: a grouped bar chart of how many patients and how many
measurements you have at each follow-up window, the figure you put at the front
of a longitudinal results section so the reader knows how much data backs the
later curves.

Both hand back a bare ggplot (or, for the longitudinal table, a panel you compose
with `patchwork`). You dress them with colour scales, labels, and a house theme
using the usual `+`.

## The data it needs

`hv_eda()` expects one row per observation with a reference axis in `x_col`
(here the surgery `year`) and the variable of interest in `y_col`. It does not
ask you to tell it whether that variable is categorical: `eda_classify_var()`
inspects the column and decides. Numeric 0/1 and small-integer columns come back
as `"Cat_Num"`, character columns as `"Cat_Char"`, and continuous columns as
`"Cont"`. The classifier is worth running on the whole frame before you build
anything, so you know which columns will draw as bars and which as a scatter.

```{r}
#| label: eda_data
dta_eda <- sample_eda_data(n = 300, seed = 42)
head(dta_eda)

# Auto-detected types for each column
sapply(dta_eda, eda_classify_var)
```

`hv_longitudinal()` is fussier: it expects **pre-aggregated** long-format data,
one row per time-window by series combination, with a count column. You do not
hand it raw patient records. `sample_longitudinal_counts_data()` derives that
shape from a simulated 300-patient registry so the recipe runs end to end.

## Build it

Start from the bare panel for `hv_longitudinal()` so you can see what the
constructor produced before any styling. Build the S3 object once and reuse it
across both panels below.

```{r}
#| label: lc_data
lc_dat <- sample_longitudinal_counts_data(n_patients = 300, seed = 42L)
lc_dat

# Build the S3 object once; use for both panels
lc <- hv_longitudinal(lc_dat)
```

```{r}
#| label: lc_bare
plot(lc)
```

That is the raw grouped bar chart: paired bars at each follow-up window, no fill
colours, no axis range, no theme. Now layer on the house style. We give the two
series distinct fills, expand the y-axis with `coord_cartesian()` so the tallest
bar has headroom, and tuck the legend into the upper-right corner where the bars
are short.

```{r}
#| label: lc_bar
p_lc_bar <- plot(lc) +
  ggplot2::scale_fill_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    name   = NULL
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq(0, 2000, 500),
    expand = c(0, 0)
  ) +
  ggplot2::coord_cartesian(ylim = c(0, 2200)) +
  ggplot2::labs(x = NULL, y = "Count (n)") +
  theme_hv_manuscript() +
  ggplot2::theme(legend.position = c(0.85, 0.85))

p_lc_bar
```

## Read it

A grouped count bar chart is read one window at a time. A few things to look
for:

- **Patients should always be at least Measurements.** Each patient contributes
  one or more measurements, never the other way round, so the Patients bar
  should never sit below the Measurements bar in the same window. If it does,
  the `series_col` mapping is pointing at the wrong column and the two series are
  swapped.
- **The shape of the decline.** Both bars fall as you move right across follow-up
  windows, because patients drop out of the registry over time. A gentle taper is
  expected; a cliff between two windows is a data-availability artefact worth
  chasing down before you trust the curves built on those windows.
- **The gap between the two bars.** A wide gap in a window means many patients
  there are contributing only a single measurement, which tells you how much you
  can lean on within-patient change in that window.

## Variations

The remaining variants come from `hv_eda()`, which draws the single-variable
exploratory bars.

### Binary categorical: count bars

Numeric 0/1 columns are classified as `"Cat_Num"`, and `NA` values appear as an
explicit `"(Missing)"` fill level rather than vanishing, so you can colour and
count them. The `y_label` argument sets the title and fill-legend name in place
of the raw column name. Here, sex by surgery year.

```{r}
#| label: eda_binary_count
plot(hv_eda(dta_eda, x_col = "year", y_col = "male",
            y_label = "Sex")) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick", "(Missing)" = "grey80"),
    labels = c("0" = "Female", "1" = "Male", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  theme_hv_manuscript()
```

### Binary categorical: percentage bars

When you care about the *mix* rather than the absolute volume, set
`show_percent = TRUE`. That switches `geom_bar()` to `position = "fill"`, so
every year's bar runs the full height and the fill shows the proportion in each
category. This is the version to use when annual cohort sizes vary a lot and a
raw count would tell you more about volume than about case mix.

```{r}
#| label: eda_binary_percent
plot(hv_eda(dta_eda, x_col = "year", y_col = "cabg",
            y_label = "Concomitant CABG", show_percent = TRUE)) +
  scale_fill_manual(
    values = c("0" = "grey70", "1" = "steelblue", "(Missing)" = "grey90"),
    labels = c("0" = "No CABG", "1" = "CABG", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Surgery Year", y = "Proportion") +
  theme_hv_manuscript()
```

### Ordinal and multi-level categorical

Columns with more than two numeric levels render as stacked count bars, one
level per fill colour. For an ordered grade such as NYHA class, a reversed
diverging palette (`"RdYlGn"`) carries the severity ordering visually, from
green at the mild end through yellow to red at the severe end, so the reader
sees worsening case mix without reading the legend.

```{r}
#| label: eda_ordinal
plot(hv_eda(dta_eda, x_col = "year", y_col = "nyha",
            y_label = "Preoperative NYHA Class")) +
  scale_fill_brewer(
    palette = "RdYlGn", direction = -1,
    labels  = c("1" = "I", "2" = "II", "3" = "III", "4" = "IV",
                "(Missing)" = "Missing"),
    name    = "NYHA"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  theme_hv_manuscript()
```

### Character categorical

String columns are classified as `"Cat_Char"` and also produce stacked count
bars. The difference is that character levels are ordered alphabetically by
default rather than by an implied scale, so use `scale_fill_manual()` to assign
colours that carry clinical meaning, here one hue per valve morphology type.

```{r}
#| label: eda_char_cat
plot(hv_eda(dta_eda, x_col = "year", y_col = "valve_morph",
            y_label = "Valve Morphology")) +
  scale_fill_manual(
    values = c(Bicuspid   = "steelblue",
               Tricuspid  = "firebrick",
               Unicuspid  = "goldenrod3",
               "(Missing)" = "grey80"),
    name = "Morphology"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  theme_hv_manuscript()
```

### Longitudinal: the numeric table panel

`hv_longitudinal()` carries a second panel. `plot(lc, type = "table")` renders
the same counts as coloured text below the x-axis labels, the numeric companion
to the bars. On its own it is sparse; it is meant to sit under the bar chart.

```{r}
#| label: lc_table
p_lc_tbl <- plot(lc, type = "table") +
  ggplot2::scale_colour_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    guide  = "none"
  ) +
  theme_hv_manuscript()

p_lc_tbl
```

### Longitudinal: the two-panel composite

Stack the bar chart above the table with `patchwork`'s `/` operator, then use
`plot_layout(heights = c(3, 1))` to give the bars three times the vertical space.
This is the figure you actually publish: the bars carry the visual story and the
table gives the reader the exact counts.

```{r}
#| label: lc_combined
#| fig-height: 5
p_lc_bar / p_lc_tbl +
  patchwork::plot_layout(heights = c(3, 1))
```

## Pitfalls

- **Count versus proportion.** A raw count bar and a `show_percent = TRUE` bar
  answer different questions. If annual cohort sizes swing widely, a count chart
  can make a stable case mix look like a trend, and a percentage chart can hide a
  collapse in volume. Show whichever matches the claim, and say which it is.
- **Dropping the missing level.** `hv_eda()` keeps `NA` as an explicit
  `"(Missing)"` level on purpose. Recolouring it to white or styling it away
  hides how much of each bar is unknown. Leave it visible, or at least account
  for it in the caption.
- **Feeding `hv_longitudinal()` raw records.** It expects pre-aggregated
  one-row-per-window-per-series data. Hand it patient-level rows and the counts
  will be wrong without an error. Aggregate first, or use
  `sample_longitudinal_counts_data()` as the template for the shape.
- **Reading a stacked bar's middle segments.** Only the bottom segment of a
  stacked bar starts from a common baseline; the middle segments float, so their
  heights are hard to compare across years. If a middle category is the story,
  pull it out into its own binary count or percentage bar.
