# Box Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

When you want to compare a continuous measurement across a handful of groups,
the box plot is the compact summary that fits in a single panel. It draws the
five-number summary, the median, the two quartiles, and the whiskers, for each
group side by side, so the reader can scan across categories and see at once
which group sits higher, which is more variable, and which has stragglers. Reach
for it when the question is "does this number differ by group?" and the groups
number in the low single digits. With twenty groups the boxes get too thin to
read; with two groups you might prefer to overlay densities. In the sweet spot of
three to six categories, the box plot is hard to beat.

`hvtiPlotR` does not provide a box-plot helper, so this chapter builds the plot
directly with ggplot2's [@R-ggplot2] `geom_boxplot()` and styles it with the
house theme `theme_hv_manuscript()` so it stays visually consistent with the
rest of the book. As with the other bare-ggplot recipes, you write the
`ggplot()` call yourself and finish with `+ theme_hv_manuscript()`.

## The data it needs

A box plot needs a categorical column for the x-axis and a continuous column for
the y-axis. We use `hvtiPlotR::sample_eda_data()` and detect a numeric column and
a categorical grouping column programmatically. The first numeric column here is
`year` (the surgery year) and the first categorical is `valve_morph` (valve
morphology: Bicuspid, Tricuspid, or Unicuspid).

```{r}
#| label: data
dat <- sample_eda_data()
num <- names(dat)[vapply(dat, is.numeric, logical(1))][1]
grp <- names(dat)[vapply(dat, function(x) is.factor(x) || is.character(x), logical(1))][1]
c(numeric = num, group = grp)
```

The `.data[[grp]]` idiom is how ggplot2 reads a column whose name lives in a
variable, so the same recipe runs against whatever columns the detection step
selected.

## Build it

The grouped box plot is the base case: one box per level of the categorical
variable, all on a shared y-axis.

```{r}
#| label: box-grouped
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]])) +
  geom_boxplot() +
  labs(
    title = "Numeric distribution by group",
    x = grp,
    y = num
  ) +
  theme_hv_manuscript()
```

Each box packs a lot in. The line inside the box is the median; the box edges
are the first and third quartiles (the 25th and 75th percentiles), so the box
spans the middle half of the data, the interquartile range or IQR. The whiskers
reach out to the most extreme points within 1.5 times the IQR of the box, and
anything past the whiskers is drawn as an individual outlier point.

## Read it

Read a box plot one box at a time, then across.

- **The median line.** This is the typical value for the group, and comparing
  median lines across boxes is the first thing your eye should do. A box that
  sits higher has a higher typical value.
- **The box height (IQR).** A tall box means the middle half of patients are
  spread out; a short box means they cluster. This is the spread you compare
  across groups, and it is more stable than the whiskers.
- **The whiskers and outliers.** The whiskers show the reach of the bulk of the
  data, and they extend to 1.5 times the IQR, not to the minimum and maximum.
  Points drawn beyond them are flagged as outliers by that same rule. They are
  candidates for a second look, not automatically errors.
- **Asymmetry.** If the median sits low in the box and the upper whisker is long,
  the group is right-skewed; the box plot shows skew even though it hides the
  finer shape.

What a box plot does not show is the shape inside the box. Two groups can have
identical boxes while one is a single clump and the other is two separated
clusters. That blind spot is the reason for the next variation.

## Variations

### Box plot with jittered points

Overlaying the raw observations with `geom_jitter()` shows the underlying sample
size and spread that a box plot alone hides. We turn off the box's own outlier
points (`outlier.shape = NA`) so they are not drawn twice, then scatter the
points with a little horizontal jitter and low alpha so dense regions read as
darker.

```{r}
#| label: box-jitter
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]])) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 0.7) +
  labs(
    title = "Box plot with jittered observations",
    x = grp,
    y = num
  ) +
  theme_hv_manuscript()
```

Now you can see how many patients sit behind each box, and whether the points
under a box form one cloud or two. For the small group in this sample, the
handful of points makes plain how little the box is summarising.

### Letting the box width show the group size

A box plot draws every box the same width regardless of how many patients it
summarises, which can mislead. Setting `varwidth = TRUE` makes each box's width
proportional to the square root of its group size, so a thin box is a visual
warning that it rests on few patients.

```{r}
#| label: box-varwidth
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]])) +
  geom_boxplot(varwidth = TRUE) +
  labs(
    title = "Box width scaled to group size",
    x = grp,
    y = num
  ) +
  theme_hv_manuscript()
```

## Pitfalls

- **Hiding the shape.** A box plot collapses the distribution to five numbers,
  so a bimodal group (two clusters with a gap between them) looks the same as a
  single clump centred at the same median. When the shape might matter, overlay
  the points as in the jitter variation, or draw a density alongside.
- **Small-n boxes.** A box computed from a dozen patients has quartiles and
  whiskers that can swing on one or two values, yet it is drawn with the same
  authority as a box from hundreds. Treat a box over a small group with caution:
  scale the width to n, overlay the points, or just report the numbers.
- **Misreading the whiskers.** The whiskers are not the minimum and maximum. They
  reach to the most extreme point within 1.5 times the IQR, and points past them
  are flagged as outliers by that rule. A long whisker does not mean the data run
  exactly that far, and an "outlier" point is a statistical flag, not a verdict
  that the value is wrong.
- **Too many or unordered groups.** Past five or six categories the boxes get too
  narrow to compare; and leaving them in alphabetical order when the variable has
  a natural order (severity, era) makes the trend harder to see than reordering
  the factor would.
