# Box Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

`hvtiPlotR` does not provide a dedicated box-plot helper, so this chapter builds
box plots directly with ggplot2's `geom_boxplot()` and styles them with the house
theme `theme_hv_manuscript()` so they stay visually consistent with the rest of
the book.

We use `hvtiPlotR::sample_eda_data()` and detect a numeric column and a
categorical grouping column programmatically.

```{r}
#| label: data
dat <- sample_eda_data()
num <- names(dat)[vapply(dat, is.numeric, logical(1))][1]
grp <- names(dat)[vapply(dat, function(x) is.factor(x) || is.character(x), logical(1))][1]
c(numeric = num, group = grp)
```

## Grouped box plot

```{r}
#| label: box-grouped
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]])) +
  geom_boxplot() +
  labs(
    title = "Numeric distribution by group",
    x = grp,
    y = num
  ) +
  theme_hv_manuscript()
```

## Box plot with jittered points

Overlaying the raw observations with `geom_jitter()` shows the underlying sample
size and spread that a box plot alone hides.

```{r}
#| label: box-jitter
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]])) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 0.7) +
  labs(
    title = "Box plot with jittered observations",
    x = grp,
    y = num
  ) +
  theme_hv_manuscript()
```
