# Color templates for consistency

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

hvtiPlotR plots carry no colour scale by default. ggplot2's [@R-ggplot2]
`scale_colour_*` controls
line and point colours; `scale_fill_*` controls filled areas (ribbons, bars).
Both take the same `name` (legend title) and `guide` (legend display)
arguments — set them together so the two scales stay in sync.

We use a grouped trends plot so the colour mapping is visible across levels.

```{r}
#| label: base
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))
```

## Manual colours

Use `scale_colour_manual()` / `scale_fill_manual()` to assign specific brand or
convention colours to known levels. Here a single-stratum KM curve is set to one
colour.

```{r}
#| label: manual
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)

plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()
```

## ColorBrewer palettes

`scale_colour_brewer()` applies a perceptually uniform, print-safe palette —
which matters when figures go to black-and-white PDF. Use `palette = "Set1"`
for categorical data, `"RdYlGn"` for diverging, `"Blues"` for sequential.

```{r}
#| label: brewer
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()
```

## Suppressing legends

Pass `guide = "none"` to any scale and that aesthetic drops out of the legend —
do this when axis labels or an annotation already name the group, so a redundant
legend does not eat panel space.

```{r}
#| label: no_legend
p_base +
  scale_colour_brewer(palette = "Dark2", guide = "none") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    guide = "none"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()
```
