# Combination figures

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
library(patchwork)
```

The `patchwork` [@R-patchwork] package composes multiple ggplot objects into a single figure.
The two patterns we reach for most are side-by-side comparisons and a main plot
stacked above a companion panel. Each panel is built independently and composed
at the last step, so you can adjust one without touching the other.

## Building the panels

We build two small hvtiPlotR [@R-hvtiPlotR] plots from the `sample_*()` generators: a survival
curve and a covariate-balance panel.

```{r}
#| label: combo_panels
km <- hv_survival(sample_survival_data(n = 300, seed = 42))

p_km <- plot(km) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()

cb <- hv_balance(sample_covariate_balance_data(n_vars = 12))

p_cb <- plot(cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  labs(x = "Standardized difference (%)", y = "") +
  theme_hv_poster()
```

## Side-by-side plots

The `|` operator places two plots next to each other in the same row, sharing
the figure height; `/` stacks them vertically.

```{r}
#| label: combo_side_by_side
#| fig-width: 10
p_cb | p_km
```

## Controlling relative widths and heights

`plot_layout(widths = ...)` sets relative widths as a numeric vector —
`c(2, 1)` makes the left panel twice as wide as the right. Use `heights` the
same way for stacked layouts.

```{r}
#| label: combo_layout
#| fig-width: 10
(p_cb | p_km) +
  plot_layout(widths = c(2, 1))   # left panel twice as wide as right
```

## Stacking a plot above a risk table

`hv_survival()` stores the numbers-at-risk table as a data frame at
`km$tables$risk` — columns `strata`, `report_time`, `n.risk`. Build a ggplot
text panel from it, then stack with `/`.

```{r}
#| label: combo_risk_table
#| fig-height: 5
risk_df <- km$tables$risk

rt_panel <- ggplot(risk_df,
                   aes(x = report_time, y = factor(strata),
                       label = n.risk)) +
  geom_text(size = 3) +
  labs(x = "Years after Operation", y = NULL) +
  theme_hv_poster() +
  theme(
    axis.line  = element_blank(),
    axis.ticks = element_blank()
  )

p_km / rt_panel +
  plot_layout(heights = c(4, 1))
```

## Shared panel tags

`plot_annotation()` adds a shared title or panel tags (A, B, C...) across all
panels. Setting `tag_levels = "A"` generates the tags most journals require.
The `&` operator applies a shared `theme()` to every panel at once.

```{r}
#| label: combo_annotation
#| fig-width: 10
(p_cb | p_km) +
  plot_annotation(
    title = "Figure 1. Outcomes after cardiac surgery",
    tag_levels = "A"
  ) &
  theme(plot.tag = element_text(size = 12, face = "bold"))
```
