# CONSORT patient-flow diagrams

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

`hv_consort()` builds a CONSORT-style patient-flow diagram via a multi-step
API: `hv_consort_start()` initializes a patient-level tracker,
`hv_consort_exclude()` adds exclusion stages, and `hv_consort()` renders the
result. The diagram is drawn by the `consort` package (grid graphics), so it
is **not** a `ggplot` — do not decorate it with `+` or `theme_*()`.

## Sample data

`sample_consort_data()` simulates a cardiac-surgery cohort and returns a
three-stage `hv_consort_tracker`.

```{r}
#| label: consort_sample
tracker <- sample_consort_data(n = 300, seed = 42)
tracker
```

## Building a tracker from your own data

Start from a data frame with one row per patient. `hv_consort_start()` records
the patient identifier and marks every patient as screened; each
`hv_consort_exclude()` call adds an exclusion stage. Exclusion rules are
two-sided formulas — `<condition> ~ "<reason>"` — evaluated against the data.
The first matching rule wins, and patients dropped earlier are skipped later,
so each stage operates only on the survivors of the last.

```{r}
#| label: consort_build
set.seed(42)
cohort <- data.frame(
  mrn         = paste0("P", 1:300),
  age         = sample(12:85, 300, replace = TRUE),
  had_surgery = sample(c(TRUE, FALSE), 300, replace = TRUE, prob = c(0.9, 0.1)),
  echo        = sample(c(TRUE, FALSE), 300, replace = TRUE, prob = c(0.85, 0.15))
)

tracker2 <- hv_consort_start(cohort, patient_id = mrn, label = "Screened") |>
  hv_consort_exclude(
    label        = "Eligible",
    col          = "excl_screen",
    age < 18     ~ "Age < 18",
    !had_surgery ~ "No qualifying surgery"
  ) |>
  hv_consort_exclude(
    label = "Analyzed",
    col   = "excl_eligible",
    !echo ~ "Missing echocardiogram"
  )
tracker2
```

## The diagram

`hv_consort()` derives the box layout from the tracker's stage metadata;
`plot()` draws it. Every exclusion column is shown as a side box by default —
pass a character vector to `side_box` to select specific ones.

```{r}
#| label: consort_plot
fig <- hv_consort(tracker)
plot(fig)
```

## Auditing the cohort

`hv_consort_summary()` returns a per-stage count table suitable for a methods
section:

```{r}
#| label: consort_summary
hv_consort_summary(tracker)
```

`hv_consort_patients()` returns the patient IDs active at a stage, or the
subset excluded for a specific reason:

```{r}
#| label: consort_patients
# IDs still in the analysed cohort
head(hv_consort_patients(tracker, "Analyzed"))

# IDs dropped at screening for being under 18
head(hv_consort_patients(tracker, "Screened", reason = "Age < 18"))
```
