# Data governance and reproducibility

Reproducible analysis starts before the first plot. A documented dataset, typed and labelled variables, and a manifest pinning the exact extract are the difference between a figure you can defend and one you cannot. The `hvtiRutilities` package supplies the primitives: synthetic data for examples, a data dictionary, label and type maps, dataset manifests, and version diffs.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hvtiRutilities)
library(gt)
```

## A reproducible sample dataset

`generate_survival_data()` produces a deterministic synthetic survival cohort. Fixing the seed makes every downstream table and figure reproducible without touching protected health information.

```{r}
#| label: sample-data
dat <- generate_survival_data(n = 200, seed = 42)
str(dat)
```

Each column carries a `label` attribute, which the governance helpers below read and propagate.

## Data dictionary

`data_dictionary()` returns a per-column summary as a data frame: variable name, label, class, distinct values, percent missing, and a compact distribution summary. It renders directly through `gt()`.

```{r}
#| label: dict
dd <- data_dictionary(dat)
gt(dd) |>
  tab_header(title = "Data dictionary")
```

## Variable labels

`label_map()` returns the current variable-to-label mapping as a two-column (`key`, `label`) data frame. Holding labels in one place means they travel into tables and figures unchanged, so axis titles and column headers stay consistent across the analysis.

```{r}
#| label: labels
lm <- label_map(dat)
gt(head(lm, 10))
```

## Inferring R data types

`r_data_types()` classifies each column by role and returns the dataset with columns coerced to the inferred type — character IDs become factors, 0/1 indicators become logical. Comparing the inferred class against the source class shows what the classifier changed.

```{r}
#| label: dtypes
typed <- r_data_types(dat)
data.frame(
  variable = names(dat),
  source   = vapply(dat,   \(x) class(x)[1], character(1)),
  inferred = vapply(typed, \(x) class(x)[1], character(1)),
  row.names = NULL
) |>
  subset(source != inferred) |>
  gt() |>
  tab_header(title = "Columns retyped by r_data_types()")
```

## Dataset manifests

A manifest records provenance for every extract: source system, extract date, row count, and a SHA-256 hash of the file. `update_manifest()` appends an entry (hashing the file on disk) and `verify_manifest()` re-hashes the data directory against the manifest to confirm nothing drifted. The chunk below is not evaluated because it writes files; run it interactively.

```{r}
#| label: manifest
#| eval: false
saveRDS(dat, "cohort.rds")
mp <- tempfile(fileext = ".yaml")

update_manifest(
  file         = "cohort.rds",
  manifest_path = mp,
  extract_date = Sys.Date(),
  n_rows       = nrow(dat),
  source       = "CORR DWH"
)

verify_manifest(manifest_path = mp, data_dir = ".", stop_on_error = FALSE)
```

A representative manifest entry:

```yaml
datasets:
- file: cohort.rds
  extract_date: '2026-05-28'
  n_rows: 200
  sha256: 8b3e0188eae294e61a16f0e09e3989e8b5249601209bde987fb5153c9ab1bd76
  source: CORR DWH
```

## Comparing dataset versions

When a data warehouse re-delivers an extract, `compare_datasets()` diffs the old and new versions and reports row-count and structural changes through its print method — a fast check that a re-delivery is what you expected.

```{r}
#| label: compare
old <- generate_survival_data(n = 150, seed = 42)
new <- generate_survival_data(n = 200, seed = 99)
compare_datasets(old, new)
```
