# Density Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

A histogram answers "how many patients fall in each bin," and that is often
enough. But the moment you want to compare the *shape* of two distributions, or
you want a smooth curve to put in a figure rather than a staircase of bars, the
kernel density estimate is the cleaner tool. A density plot draws a continuous
curve whose height at any value reflects how concentrated the data are there,
and the area under the whole curve is one. Reach for it when the question is
about the form of a single continuous variable: is it roughly symmetric, is it
skewed toward high values, does it have one peak or two?

`hvtiPlotR` does not provide a density helper, so this chapter builds the plot
directly with ggplot2's [@R-ggplot2] `geom_density()`. We still style it with the
house theme `theme_hv_manuscript()` so a density panel sits comfortably next to
the survival curves and scatter plots in the same paper. The pattern is the one
you will use for any bare-ggplot recipe in the book: write the `ggplot()` call
yourself, then finish with `+ theme_hv_manuscript()`.

## The data it needs

A density plot needs one continuous column. To compare groups you add a
categorical column to map to colour. We use `hvtiPlotR::sample_eda_data()` and
detect a numeric column and a categorical grouping column programmatically, so
the recipe keeps working if the sample schema changes. The first numeric column
here is `year` (the surgery year) and the first categorical is `valve_morph`
(valve morphology: Bicuspid, Tricuspid, or Unicuspid).

```{r}
#| label: data
dat <- sample_eda_data()
num <- names(dat)[vapply(dat, is.numeric, logical(1))][1]
grp <- names(dat)[vapply(dat, function(x) is.factor(x) || is.character(x), logical(1))][1]
c(numeric = num, group = grp)
```

The `.data[[num]]` idiom you will see below is how ggplot2 reads a column whose
name lives in a variable rather than typed out as a bare symbol. It lets the
same recipe run against whatever column the detection step picked.

## Build it

Start with a single curve so you can see the estimate before any grouping. The
`fill` gives the curve a soft body and `alpha` keeps it from reading as a solid
block.

```{r}
#| label: density-single
ggplot(dat, aes(x = .data[[num]])) +
  geom_density(fill = "grey70", alpha = 0.7) +
  labs(
    title = "Distribution of a single numeric variable",
    x = num,
    y = "Density"
  ) +
  theme_hv_manuscript()
```

That curve is a smoothed version of the histogram you would otherwise draw. The
y-axis is density, not count: it is scaled so the total area is one, which is
what lets you put two of these on the same panel and compare them fairly even
when the groups have different sizes.

## Read it

A density curve gives you the shape of the distribution at a glance. Look for:

- **Peaks (modes).** One hump means the data cluster around a single typical
  value. Two humps (bimodal) usually mean you have mixed two populations that
  belong in separate groups, an early-era and a late-era cohort, say, and the
  average between the peaks describes nobody.
- **Skew.** A long tail to the right (a few large values pulling the curve out)
  is right skew; a long left tail is left skew. Skew tells you whether the mean
  and the median will agree, and whether a log transform might help downstream.
- **Width.** A tall, narrow curve means the values are tightly concentrated; a
  low, broad curve means they are spread out. The height is not a count, so do
  not read a taller curve as "more patients."

One thing to keep in the back of your mind: the curve depends on a smoothing
choice called the bandwidth, and the same data can look smoother or bumpier
depending on it. We come back to that in the pitfalls.

## Variations

### Grouped density

Mapping the grouping variable to `fill` overlays one density curve per group,
with transparency so the overlapping regions stay readable. This is the
comparison you actually want most of the time: not "what does this variable look
like" but "does it look different across valve morphologies?"

```{r}
#| label: density-grouped
ggplot(dat, aes(x = .data[[num]], fill = .data[[grp]])) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density by group",
    x = num,
    y = "Density",
    fill = grp
  ) +
  theme_hv_manuscript()
```

Because each curve integrates to one, a small group and a large group are drawn
on the same footing. That is the right default for comparing shape, but it also
hides how many patients sit under each curve, which is the trap we flag below.

### Tuning the bandwidth

The single most consequential knob is the bandwidth, passed through `adjust`
(a multiplier on the automatic bandwidth). Drawing two settings side by side is
the honest way to show how much of a bump is signal and how much is the smoother.

```{r}
#| label: density-bandwidth
ggplot(dat, aes(x = .data[[num]])) +
  geom_density(aes(colour = "wiggly (adjust = 0.4)"),
               adjust = 0.4, fill = NA, linewidth = 0.8) +
  geom_density(aes(colour = "smooth (adjust = 1.5)"),
               adjust = 1.5, fill = NA, linewidth = 0.8) +
  scale_colour_manual(values = c("wiggly (adjust = 0.4)" = "firebrick",
                                 "smooth (adjust = 1.5)" = "steelblue"),
                      name = "Bandwidth") +
  labs(
    title = "Same data, two bandwidths",
    x = num,
    y = "Density"
  ) +
  theme_hv_manuscript()
```

If a feature survives both settings it is probably real; if it appears only at
the small bandwidth it is probably noise the smoother stopped hiding.

## Pitfalls

- **Over-smoothing and under-smoothing.** Too large a bandwidth erases real
  structure, a second mode flattens into a shoulder. Too small a bandwidth
  invents structure, every cluster of three points becomes its own little peak.
  When a bump matters to your story, redraw it at a wider and a narrower
  bandwidth before you commit, the way the variation above does.
- **Treating a density like a histogram.** The y-axis is a density, not a count.
  A taller curve does not mean more patients; it means the data are more
  concentrated there. If the reader needs counts, draw a histogram or label the
  group sizes.
- **Comparing groups with very different n.** Each curve is normalised to area
  one, so a 19-patient group and a 167-patient group look equally trustworthy
  even though one rests on a tenth of the data. The small group's curve is far
  more sensitive to a handful of values. Note the group sizes in the caption so
  the reader weights the comparison correctly.
- **Densities that spill past the possible range.** A kernel density will happily
  put mass below zero for a strictly positive measurement, or above 100 for a
  percent. The smoother does not know the bounds. Constrain the axis or
  transform the variable rather than letting the curve imply impossible values.
