# Density Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

`hvtiPlotR` does not provide a dedicated density-plot helper, so this chapter
builds density plots directly with ggplot2's `geom_density()` and styles them with
the house theme `theme_hv_manuscript()` so they stay visually consistent with the
rest of the book.

We use `hvtiPlotR::sample_eda_data()` and detect a numeric column and a
categorical grouping column programmatically, so the recipes are robust to schema
changes.

```{r}
#| label: data
dat <- sample_eda_data()
num <- names(dat)[vapply(dat, is.numeric, logical(1))][1]
grp <- names(dat)[vapply(dat, function(x) is.factor(x) || is.character(x), logical(1))][1]
c(numeric = num, group = grp)
```

## Single density

```{r}
#| label: density-single
ggplot(dat, aes(x = .data[[num]])) +
  geom_density(fill = "grey70", alpha = 0.7) +
  labs(
    title = "Distribution of a single numeric variable",
    x = num,
    y = "Density"
  ) +
  theme_hv_manuscript()
```

## Grouped density

Mapping the grouping variable to `fill` overlays one density curve per group,
with transparency so overlapping regions remain readable.

```{r}
#| label: density-grouped
ggplot(dat, aes(x = .data[[num]], fill = .data[[grp]])) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density by group",
    x = num,
    y = "Density",
    fill = grp
  ) +
  theme_hv_manuscript()
```
