# Tables that accompany figures

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hvtiPlotR)
library(gt)
library(ggplot2)
library(patchwork)
```

Many manuscript figures are paired with a small numeric table: a
numbers-at-risk strip below a Kaplan-Meier curve, or the count panel
beneath a longitudinal participation bar chart. The `hvtiPlotR` plot
objects carry these companion tables alongside the plot data, so we can
pull the numbers straight out and render them with `gt`. As with the
[publication tables](qt_tables.qmd) chapter, **these tables will migrate
to the planned hvtiRtables package** once it is available.

## Numbers at risk

`hv_survival()` stores its companion tables in the object's `$tables`
slot. The numbers-at-risk strip lives in `$tables$risk` (one row per
strata × report-time) and the per-time-point summary in `$tables$report`.

```{r}
#| label: km_build
km <- hv_survival(sample_survival_data(n = 500, seed = 42))
names(km$tables)
str(km$tables$risk)
```

We render the at-risk table directly with `gt`, relabelling the columns
for presentation.

```{r}
#| label: km_risk_gt
km$tables$risk |>
  gt() |>
  tab_header(
    title = "Numbers at Risk",
    subtitle = "Patients remaining under follow-up at each report time"
  ) |>
  cols_label(
    strata      = "Stratum",
    report_time = "Time (years)",
    n.risk      = "At risk (n)"
  )
```

The companion report table carries the survival estimates and confidence
limits at the same report times — useful as the figure caption's numeric
backing.

```{r}
#| label: km_report_gt
km$tables$report |>
  gt() |>
  tab_header(title = "Survival Estimates at Report Times") |>
  cols_label(
    strata      = "Stratum",
    report_time = "Time (years)",
    surv        = "S(t)",
    lower       = "95% LCL",
    upper       = "95% UCL",
    n.risk      = "At risk",
    n.event     = "Events"
  ) |>
  fmt_number(columns = c(surv, lower, upper), decimals = 3)
```

## Longitudinal count table

`hv_longitudinal()` produces a two-panel figure: a grouped bar chart of
participation counts and a numeric table panel below it. The underlying
counts live in the object's `$data` slot in long format (one row per
time-window × series), with the column roles recorded in `$meta`.

```{r}
#| label: lc_build
lc <- hv_longitudinal(sample_longitudinal_counts_data(n_patients = 300, seed = 42L))
names(lc)
lc$meta[c("x_col", "group_col", "count_col")]
str(lc$data)
```

We reshape the long counts into a wide table — one row per follow-up
window, one column per series — and render it with `gt`. This is the
tabular companion to the bar chart's table panel.

```{r}
#| label: lc_table_gt
lc_wide <- lc$data |>
  tidyr::pivot_wider(
    names_from  = series,
    values_from = count
  )
lc_wide

lc_wide |>
  gt(rowname_col = "time_label") |>
  tab_header(
    title    = "Longitudinal Participation Counts",
    subtitle = "Patients and measurements at each follow-up window"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 0) |>
  tab_stubhead(label = "Follow-up window")
```

For reference, the same counts rendered as the in-figure table panel
appear below the bar chart when the two are composed with `patchwork`.

```{r}
#| label: lc_figure
#| fig-height: 5
p_bar <- plot(lc) +
  scale_fill_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    name   = NULL
  ) +
  labs(x = NULL, y = "Count (n)")

p_tbl <- plot(lc, type = "table") +
  scale_colour_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    guide  = "none"
  )

p_bar / p_tbl + plot_layout(heights = c(3, 1))
```
