# Hazard and nonparametric curves

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

This chapter covers three families of pre-computed curve plots: the
parametric `hazard_plot()`, the nonparametric temporal-trend curve
(`hv_nonparametric()`), and the ordinal-outcome curve (`hv_ordinal()`).

## Hazard plot

`hazard_plot()` plots pre-computed parametric curves — survival, hazard, or
cumulative hazard — with an optional Kaplan-Meier empirical overlay. You pass
the estimate column and its CI bounds explicitly. Two helpers generate the
demonstration data: `sample_hazard_data()` for the parametric prediction grid
(`survival`, `hazard`, `cumhaz` plus CI bounds) and `sample_hazard_empirical()`
for the binned KM overlay.

```{r}
#| label: hazard_data
dat_hp <- sample_hazard_data(n = 500, time_max = 10)
emp_hp <- sample_hazard_empirical(n = 500, time_max = 10, n_bins = 6)
head(dat_hp)
```

### Survival curve with KM overlay

The most common variant: a smooth parametric survival curve with confidence
band plus discrete KM empirical points and error bars.

```{r}
#| label: hazard_survival
hazard_plot(
  dat_hp,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_hp,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival (%)") +
  theme_hv_poster()
```

### Hazard rate

Switch `estimate_col` to `"hazard"` and the matching CI columns for the
instantaneous hazard rate (%/year).

```{r}
#| label: hazard_rate
hazard_plot(
  dat_hp,
  estimate_col = "hazard",
  lower_col    = "haz_lower",
  upper_col    = "haz_upper"
) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 30),
                     labels = function(x) paste0(x, "%/yr")) +
  labs(x = "Years", y = "Hazard (%/year)") +
  theme_hv_poster()
```

### Cumulative hazard

The `"cumhaz"` column equals `-log(S) * 100`; used for readmission and
repeated-event analyses.

```{r}
#| label: hazard_cumhaz
hazard_plot(
  dat_hp,
  estimate_col  = "cumhaz",
  lower_col     = "cumhaz_lower",
  upper_col     = "cumhaz_upper"
) +
  scale_colour_manual(values = c("darkorange"), guide = "none") +
  scale_fill_manual(values = c("darkorange"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Cumulative Hazard (%)") +
  theme_hv_poster()
```

## Nonparametric temporal trend curve

`hv_nonparametric()` prepares pre-computed average curves with bootstrap CI
ribbons from a two-phase nonparametric temporal-trend model.
`sample_nonparametric_curve_data()` generates the average curve and CI;
`sample_nonparametric_curve_points()` generates the binned summary points that
overlay it. Build the S3 object once with the matching `lower_col`/`upper_col`
and optional `data_points`, then `plot()`.

```{r}
#| label: np_curve_data
curve_dat <- sample_nonparametric_curve_data(
  n            = 500,
  time_max     = 12,
  outcome_type = "probability",
  ci_level     = 0.68
)
pts_dat <- sample_nonparametric_curve_points(n = 500, time_max = 12)

np <- hv_nonparametric(
  curve_data  = curve_dat,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = pts_dat
)
```

### Single average curve with CI ribbon

The shaded ribbon here is the 68% bootstrap CI (one standard error); pass
`ci_level = 0.95` to `sample_nonparametric_curve_data()` for 95% bands.

```{r}
#| label: np_curve_single
plot(np) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values   = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2),
                     labels = function(x) paste(x, "yr")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = scales::percent) +
  labs(x = "Follow-up (years)", y = "Prevalence (%)") +
  theme_hv_poster()
```

### Two-group comparison

Pass `group_col` to the constructor to compare two average curves in a single
panel; each group gets its own CI ribbon.

```{r}
#| label: np_curve_groups
curve_grp <- sample_nonparametric_curve_data(
  n            = 400,
  time_max     = 7,
  groups       = c("Ozaki" = 0.7, "CE-Pericardial" = 1.3),
  outcome_type = "continuous",
  ci_level     = 0.68
)
pts_grp <- sample_nonparametric_curve_points(
  n = 400, time_max = 7,
  groups = c("Ozaki" = 0.7, "CE-Pericardial" = 1.3),
  outcome_type = "continuous"
)
np_grp <- hv_nonparametric(
  curve_data  = curve_grp,
  group_col   = "group",
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = pts_grp
)

plot(np_grp) +
  scale_colour_manual(
    values = c("Ozaki" = "steelblue", "CE-Pericardial" = "firebrick"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("Ozaki" = "steelblue", "CE-Pericardial" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 7), breaks = 0:7) +
  labs(x = "Follow-up (years)", y = "AV Peak Gradient (mmHg)") +
  theme_hv_poster() +
  theme(legend.position = c(0.15, 0.85))
```

## Nonparametric ordinal outcome curve

`hv_ordinal()` prepares grade-specific probability curves from a cumulative
proportional-odds model (e.g. TR grade or AR severity prevalence over time).
The curve data is long format — one row per time × grade.
`sample_nonparametric_ordinal_data()` generates the curves and
`sample_nonparametric_ordinal_points()` the binned summary points.

```{r}
#| label: np_ordinal_data
ord_dat <- sample_nonparametric_ordinal_data(
  n = 800, time_max = 5,
  grade_labels = c("None", "Mild", "Moderate", "Severe")
)
ord_pts <- sample_nonparametric_ordinal_points(
  n = 800, time_max = 5,
  grade_labels = c("None", "Mild", "Moderate", "Severe")
)
head(ord_dat)
```

### Grade probability curves with data points

Build the object with both curve data and summary points. Each line is one
grade level; the lines sum to approximately 1.0 at every time point.

```{r}
#| label: np_ordinal_basic
np_ord <- hv_ordinal(curve_data = ord_dat, data_points = ord_pts)

plot(np_ord) +
  scale_colour_manual(
    values = c("None" = "grey40", "Mild" = "steelblue",
               "Moderate" = "darkorange", "Severe" = "firebrick"),
    name = "AR Grade"
  ) +
  scale_x_continuous(limits = c(0, 5), breaks = 0:5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = scales::percent) +
  labs(x = "Follow-up (years)", y = "Grade prevalence (%)") +
  theme_hv_poster() +
  theme(legend.position = c(0.75, 0.6))
```
