# Histogram Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## Mirrored propensity-score histogram

In propensity-matched analyses we routinely produce a mirrored histogram to
show how well matching compressed the overlap between two treatment groups.
`hv_mirror_hist()`, from hvtiPlotR [@R-hvtiPlotR], prepares the data; `plot()`
hands back a bare ggplot to dress with colour and labels. Two display modes are selected by the
constructor arguments: **binary-match mode** (`match_col`) shows all
observations before matching with the matched subset overlaid, and **weighted
IPTW mode** (`weight_col`) shows raw counts with per-bin IPTW weight sums
overlaid.

### Binary-match mode (SAVR vs. TF-TAVR)

`sample_mirror_histogram_data()` simulates 2,000 patients with a continuous
propensity score (`prob_t`), a binary group indicator (`tavr`), and a match
flag (`match`). `score_multiplier = 100` puts the score on a 0--100 percent
scale and `binwidth = 5` sets 5-point bins.

```{r}
#| label: mirror_histogram_build
mirror_dta <- sample_mirror_histogram_data(n = 2000, separation = 1.5)

mh <- hv_mirror_hist(
  data             = mirror_dta,
  score_col        = "prob_t",
  group_col        = "tavr",
  match_col        = "match",
  group_levels     = c(0, 1),
  group_labels     = c("SAVR", "TF-TAVR"),
  matched_value    = 1,
  score_multiplier = 100,
  binwidth         = 5
)
```

The bare panel shows two mirrored bar charts — upper bars for the first group,
lower for the second — with white fill and no scale, label, or theme yet.

```{r}
#| label: mirror_histogram_bare
p <- plot(mh, alpha = 0.8)
p
```

`scale_fill_manual()` maps the four internal fill levels (`before_g0`,
`matched_g0`, `before_g1`, `matched_g1`) to white (pre-match) and two greens
(matched subsets). The annotations use `y = Inf`/`-Inf` with `vjust` to anchor
labels at the panel edges; `theme_hv_poster()` sizes text for a conference
poster.

```{r}
#| label: mirror_histogram_decorated
p +
  ggplot2::scale_fill_manual(
    values = c(
      before_g0  = "white",  matched_g0 = "green1",
      before_g1  = "white",  matched_g1 = "green4"
    ),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)
  ) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::annotate("text", x = 20, y =  Inf, vjust =  2,
                    label = mh$meta$group_labels[1], size = 7) +
  ggplot2::annotate("text", x = 20, y = -Inf, vjust = -1,
                    label = mh$meta$group_labels[2], size = 7) +
  ggplot2::labs(x = "Propensity Score (%)", y = "Number of Patients") +
  theme_hv_poster()
```

The constructor stores group counts and standardized mean differences (SMD)
before and after matching in `$tables$diagnostics`:

```{r}
#| label: mirror_histogram_diagnostics
mh$tables$diagnostics$smd_before
mh$tables$diagnostics$smd_matched
mh$tables$diagnostics$group_counts_before
mh$tables$diagnostics$group_counts_matched
```

### Weighted IPTW mode (Limited vs. Extended)

When the analysis uses inverse probability of treatment weighting rather than
1:1 matching, pass `weight_col` instead of `match_col`. Each bin's overlay
height is the sum of IPTW weights in that bin. `add_weights = TRUE` in the
sample generator attaches an `mt_wt` column.

```{r}
#| label: mirror_histogram_weighted_build
wt_dta <- sample_mirror_histogram_data(
  n = 2000, separation = 1.5, add_weights = TRUE
)

mh_wt <- hv_mirror_hist(
  data             = wt_dta,
  score_col        = "prob_t",
  group_col        = "tavr",
  group_levels     = c(0, 1),
  group_labels     = c("Limited", "Extended"),
  weight_col       = "mt_wt",
  score_multiplier = 100,
  binwidth         = 5
)
```

The bare weighted panel looks like the binary-match bare plot, but the overlay
encodes IPTW weight sums, not counts.

```{r}
#| label: mirror_histogram_weighted_bare
p_wt <- plot(mh_wt, alpha = 0.8)
p_wt
```

The IPTW variant uses blue/red for the Limited/Extended groups, with the group
labels coloured to match.

```{r}
#| label: mirror_histogram_weighted_decorated
p_wt +
  ggplot2::scale_fill_manual(
    values = c(
      before_g0   = "white", weighted_g0 = "blue",
      before_g1   = "white", weighted_g1 = "red"
    ),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)
  ) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::annotate("text", x = 30, y =  Inf, vjust =  2,
                    label = mh_wt$meta$group_labels[1], color = "blue", size = 5) +
  ggplot2::annotate("text", x = 30, y = -Inf, vjust = -1,
                    label = mh_wt$meta$group_labels[2], color = "red",  size = 5) +
  ggplot2::labs(x = "Propensity Score (%)", y = "#") +
  theme_hv_poster()
```

Weighted diagnostics include the effective N and weighted SMD:

```{r}
#| label: mirror_histogram_weighted_diagnostics
mh_wt$tables$diagnostics$smd_weighted
mh_wt$tables$diagnostics$effective_n_by_group
```

## Stacked histogram

The stacked histogram shows how the composition of a numeric variable shifts
over time or across a grouping dimension. `hv_stacked()` prepares the data;
`plot()` hands back a bare ggplot. `sample_stacked_histogram_data()` generates
a reproducible dataset with `year` and `category` columns.

```{r}
#| label: stacked_histogram_data
hist_dta <- sample_stacked_histogram_data(n_years = 20, start_year = 2000,
                                           n_categories = 3)
head(hist_dta)
```

### Count histogram

The default `position = "stack"` shows raw counts within each bin. Build the
object with `hv_stacked()`, then dress it with colour scales and a theme.

```{r}
#| label: stacked_histogram_count
sh <- hv_stacked(hist_dta, x_col = "year", group_col = "category")
p_count <- plot(sh)

p_count +
  scale_fill_brewer(palette = "Set1", name = "Category") +
  scale_color_brewer(palette = "Set1", name = "Category") +
  labs(x = "Year", y = "Count") +
  theme_hv_manuscript()
```

### Proportion (fill) histogram

`position = "fill"` rescales each bin so the bars sum to 1, making relative
composition comparable across years without count differences obscuring the
trend.

```{r}
#| label: stacked_histogram_fill
sh2 <- hv_stacked(hist_dta, x_col = "year", group_col = "category",
                  position = "fill")
p_fill <- plot(sh2)

p_fill +
  scale_fill_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    labels = c("1" = "Group A", "2" = "Group B", "3" = "Group C"),
    name   = "Category"
  ) +
  scale_color_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    guide  = "none"
  ) +
  labs(x = "Year", y = "Proportion") +
  theme_hv_manuscript()
```
