# Using legends

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

ggplot2 [@R-ggplot2] places the legend outside the panel by default. For publication figures
we often move it inside, drop it, or shrink it — the axis labels frequently do
the identification work already. Legend behaviour is controlled through
`theme()` (or directly via a theme's `...`), and per-aesthetic via `guide =`.

We use a grouped trends plot so there is a multi-level legend to position.

```{r}
#| label: base
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))
```

## Inside the panel

Pass fractional coordinates `c(x, y)` to `legend.position`. `c(0, 0)` is
bottom-left; `c(1, 1)` is top-right. A light background box keeps it legible
over data.

```{r}
#| label: inside
p_base +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = NULL
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.position   = c(0.15, 0.2),
    legend.background = element_rect(fill = "white", colour = "grey80",
                                     linewidth = 0.3)
  )
```

## Outside the panel (explicit sides)

Set `legend.position` to `"right"`, `"left"`, `"top"`, or `"bottom"`.
`"bottom"` with `legend.direction = "horizontal"` is the common choice for
multi-group figures where a corner legend would crowd the panel.

```{r}
#| label: outside
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.position  = "bottom",
    legend.direction = "horizontal"
  )
```

## Suppress all legends

`theme(legend.position = "none")` hides every legend in one shot.
Alternatively, `guide = "none"` on each `scale_*()` call drops only that
aesthetic's legend — handy when you want a colour legend but no shape legend.

```{r}
#| label: none
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)

plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()
```

## Legend text and key size

The legend inherits the theme's font size, often larger than needed inside a
crowded panel. `legend.text` controls the label font; `legend.key.size` shrinks
the colour/shape swatch.

```{r}
#| label: text
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.text     = element_text(size = 9),
    legend.key.size = unit(0.4, "cm")
  )
```
