# Preparing figures for manuscripts

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

Journals impose hard constraints on figures: a fixed column or page width, a
maximum height, and a vector format (PDF/EPS) so type and lines stay crisp at
print resolution. The trap is that two figures saved at the same `ggsave()`
dimensions rarely share the same *data region*: a longer axis label or a wider
legend steals space from the panel, so the plotting area drifts from figure to
figure. `theme_hv_manuscript()` sets a print-appropriate base size, and
`hv_ggsave_dims()` pins the panel content area to a constant rectangle across a
figure set.

## A manuscript-styled plot

`theme_hv_manuscript()` uses a 12 pt base on a white background — sized for the
final printed page rather than the screen.

```{r}
#| label: build-manuscript
p    <- plot(hv_survival(sample_survival_data(n = 500, seed = 42)))
p_ms <- p + theme_hv_manuscript()
p_ms
```

## Manuscript PDF

For most figures the quickest path is `save_manuscript()`, which writes a
finished plot at the house manuscript size — 6 × 4 inches — in one call.

```{r}
#| label: save-manuscript-default
#| eval: false
save_manuscript(p_ms, "trends_manuscript.pdf")
```

It is the manuscript counterpart of `save_ppt()`: it fixes the output geometry,
not the theme, so the 12 pt type comes from the `theme_hv_manuscript()` you have
already applied. Override the size when a journal asks for it — for a single
column, say `save_manuscript(p_ms, "fig.pdf", width = 3.5, height = 3)` — and
pass `device = cairo_pdf` to embed fonts where your system has cairo. The chunks
in this section are `eval: false` because each writes a binary file.

### Holding the panel constant across a figure set

`save_manuscript()`, like `ggsave()`, fixes the *file* dimensions but lets the
panel float inside them. When you need the panel content area identical across
several figures — so the data regions line up regardless of how much room each
plot's axis labels take — `hv_ggsave_dims()` inverts the problem: you specify the
panel content area you want (the rectangular bounding box of the gtable cells
tagged `panel`) and it returns the `width`/`height`/`units` such that `ggsave()`
yields exactly that panel size, expanding the file to absorb axes, titles, and
legend (the "chrome"). The result is a named list shaped to splat directly into
`ggsave()` via `do.call()`.

```{r}
#| label: save-manuscript
#| eval: false
dims <- hv_ggsave_dims(p_ms, width = 6.5, height = 4.0, units = "in")

do.call(
  ggsave,
  c(
    list(filename = "trends_manuscript.pdf", plot = p_ms, device = cairo_pdf),
    dims
  )
)
```

`cairo_pdf` is the device of choice here: it embeds fonts and renders
transparency and custom typefaces correctly, where the base `pdf` device can
substitute fonts. The chunk is `eval: false` because it writes a binary file.
The code is shown, not run during the book build.

For multi-panel figure sets where the data region must be identical across
several PDFs regardless of label length, call `hv_ggsave_dims()` with the same
`width`/`height` for each figure. Its slide-layout analogue,
`hv_ph_location()`, computes `officer::ph_location()` coordinates so a panel
lands at the same slide position in a deck.

## PowerPoint export

`save_ppt()` writes the plot into a PowerPoint slide as **editable DrawingML
vector graphics** via the `officer` and `rvg` packages: shapes, lines, and
text remain selectable in PowerPoint, so co-authors can re-label or recolor
without round-tripping back to R. It appends a slide to a template `.pptx`
rather than overwriting it.

```{r}
#| label: save-ppt
#| eval: false
template <- system.file("ClevelandClinic.pptx", package = "hvtiPlotR")

p_ppt <- p + theme_hv_ppt_dark()

save_ppt(
  object       = p_ppt,
  template     = template,
  powerpoint   = "trends_slides.pptx",
  slide_titles = "Overall Survival",
  width        = 10.1,
  height       = 5.8
)
```

This chunk is `eval: false` because it writes a `.pptx` file. To anchor the
panel content area at identical slide coordinates across a multi-slide deck
(useful when axis-label widths differ between plots), pass
`panel_box = list(width, height, left, top)`; `save_ppt()` delegates to
`hv_ph_location()` per slide.
