# Number needed to treat and survival difference

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

A hazard ratio tells a reviewer that one treatment is better, but it does not
tell a patient what the choice is worth to them. These two figures translate a
survival comparison into the language clinicians and patients actually use.
The survival difference plots the gap between two survival curves over time: how
many more patients out of a hundred are still alive at five years if they take
the better arm. The number needed to treat (NNT) inverts that gap into a count:
how many patients you must treat to prevent one event by a given time.

Reach for the survival difference when you want to show the size and timing of a
benefit and whether its confidence band clears zero. Reach for NNT when you want
the bedside number, the figure that makes a small absolute benefit feel concrete
or an over-hyped one feel modest. Both functions, from hvtiPlotR [@R-hvtiPlotR],
take a pre-computed data frame and a single call: you name the estimate column
and its CI bounds, then decorate the returned ggplot.

## The data it needs

Both functions expect one row per time point with an estimate column and
explicit lower/upper CI columns that you name in the call. `sample_nnt_data()`
generates two survival curves from a hazard-ratio contrast and computes NNT and
ARR (absolute risk reduction) at each time point with CI bounds. The `groups`
argument names the arms and sets their hazard multipliers; here the internal
thoracic artery graft (ITA) carries 0.75 times the hazard of the saphenous vein
graft (SVG) reference.

```{r}
#| label: nnt_data
nnt_dat <- sample_nnt_data(
  n        = 500,
  time_max = 20,
  groups   = c("SVG" = 1.0, "ITA" = 0.75)
)
head(nnt_dat)
```

## Build it

The NNT curve takes the pre-computed estimate and its CI columns directly. The
default `estimate_col` is `"nnt"`, so you need only name the CI bounds.

```{r}
#| label: fig-nnt
#| fig-cap: "Number needed to treat over follow-up with its confidence band, falling as the survival gap widens"
plot(hv_nnt(
  nnt_dat,
  lower_col = "nnt_lower",
  upper_col = "nnt_upper"
)) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_y_continuous(limits = c(0, 50), breaks = seq(0, 50, 10)) +
  labs(x = "Years", y = "Number Needed to Treat") +
  theme_hv_poster()
```

## Read it

An NNT curve falls over time, and that fall is the whole point. Look for:

- **The downward slope.** Early in follow-up the two arms have barely diverged,
  so you must treat a great many patients to prevent one event and the NNT is
  large. As the survival gap widens the NNT drops: the same treatment buys more
  per patient the longer you wait. Read the value at the time horizon that
  matters for your decision, not at the visual endpoint.
- **The confidence band.** A wide band, especially early when NNT is large and
  unstable, means the count is poorly determined. If the band runs off to very
  large values, the benefit at that time is essentially unmeasurable.
- **The y-axis scale.** NNT is a ratio and blows up when the absolute benefit is
  near zero. A capped axis (here 0 to 50) keeps the figure readable, but
  remember the curve is heading to infinity at time zero, not flattening.

## Variations

### Absolute risk reduction

The same data plotted as ARR (%) by setting `estimate_col = "arr"` and the
matching CI columns. ARR is the reciprocal of NNT and rises over time as the
benefit accumulates, which many readers find easier to take in than a falling
count. Show whichever frames your benefit most honestly; show both if a reviewer
asks.

```{r}
#| label: fig-nnt-arr
#| fig-cap: "Absolute risk reduction over follow-up, the reciprocal of NNT, rising as the benefit accumulates"
plot(hv_nnt(
  nnt_dat,
  estimate_col = "arr",
  lower_col    = "arr_lower",
  upper_col    = "arr_upper"
)) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_y_continuous(limits = c(0, 50),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Absolute Risk Reduction (%)") +
  theme_hv_poster()
```

### Survival difference

`hv_survival_difference()` plots the difference `S_2(t) - S_1(t)` between two
groups over time with an optional confidence band.
`sample_survival_difference_data()` generates the point-wise difference with
bootstrap CI bounds from a hazard-ratio contrast. Both arms are simulated at the
same n to keep the band width realistic.

```{r}
#| label: surv_diff_data
diff_dat <- sample_survival_difference_data(
  n      = 500,
  groups = c("Control" = 1.0, "Treatment" = 0.70)
)
head(diff_dat)
```

The dashed line at zero is the no-benefit baseline; a positive difference means
the treatment group has higher survival. No grouping argument is needed for the
two-group case. Read the figure two ways: the height of the curve is the size of
the benefit at each time, and the moment the lower edge of the band clears zero
is when the benefit becomes statistically convincing.

```{r}
#| label: fig-nnt-survival-difference
#| fig-cap: "Survival difference between two groups over time with its confidence band and a dashed no-benefit baseline"
plot(hv_survival_difference(
  diff_dat,
  lower_col = "diff_lower",
  upper_col = "diff_upper"
)) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(-5, 40),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival Difference (%)") +
  theme_hv_poster()
```

### Multiple comparisons

Combine several two-group differences into one long data frame and use
`group_col` to overlay them. This is how we put one reference arm against
several alternatives in a single panel: each line is one contrast, and the ones
that sit highest and clear zero earliest are the comparisons with the strongest
benefit.

```{r}
#| label: fig-nnt-survival-difference-multi
#| fig-cap: "Several two-group survival differences overlaid in one panel, one line per contrast against a shared reference arm"
d1 <- sample_survival_difference_data(
  groups = c("Medical Mgmt" = 1.0, "TF-TAVR" = 0.70), seed = 1L
)
d1$comparison <- "TF-TAVR vs Medical Mgmt"

d2 <- sample_survival_difference_data(
  groups = c("TA-TAVR" = 0.90, "TF-TAVR" = 0.70), seed = 2L
)
d2$comparison <- "TF-TAVR vs TA-TAVR"

d3 <- sample_survival_difference_data(
  groups = c("AVR" = 0.80, "TF-TAVR" = 0.70), seed = 3L
)
d3$comparison <- "TF-TAVR vs AVR"

plot(hv_survival_difference(rbind(d1, d2, d3), group_col = "comparison")) +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Survival Difference (%)") +
  theme_hv_poster()
```

## Pitfalls

- **Quoting NNT at time zero.** NNT is a reciprocal, so it explodes near the
  start of follow-up where the arms have not yet diverged. Always report it at a
  named horizon, never as a single headline number with no time attached.
- **Mismatched CI columns.** NNT and ARR each have their own pair of bounds.
  Plot ARR with the NNT bounds and the band will be wrong. Name the CI columns
  that go with the estimate you are showing.
- **Reading the survival difference where the band straddles zero.** A positive
  point estimate whose confidence band still includes zero is not yet a
  demonstrated benefit. The figure is most honest when you let the band do the
  arguing.
- **Forgetting the zero line.** The dashed reference at zero is what makes the
  survival-difference figure readable. Add the `geom_hline()` every time; without
  it the eye has no anchor for "no benefit".
