# Number needed to treat and survival difference

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

Both plots in this chapter take a pre-computed data frame and a single call —
you name the estimate column and its CI bounds, then decorate the returned
ggplot.

## Number needed to treat

`nnt_plot()` plots number needed to treat (or absolute risk reduction) over
time. `sample_nnt_data()` generates two survival curves from a hazard-ratio
contrast and computes NNT and ARR at each time point with CI bounds.

```{r}
#| label: nnt_data
nnt_dat <- sample_nnt_data(
  n        = 500,
  time_max = 20,
  groups   = c("SVG" = 1.0, "ITA" = 0.75)
)
head(nnt_dat)
```

### NNT curve

NNT decreases over time as the treatment benefit accumulates: early on you
must treat many patients to prevent one event; later the survival gap widens
and fewer are needed.

```{r}
#| label: nnt_curve
nnt_plot(
  nnt_dat,
  lower_col = "nnt_lower",
  upper_col = "nnt_upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_y_continuous(limits = c(0, 50), breaks = seq(0, 50, 10)) +
  labs(x = "Years", y = "Number Needed to Treat") +
  theme_hv_poster()
```

### Absolute risk reduction

The same data plotted as ARR (%) by setting `estimate_col = "arr"` and the
matching CI columns.

```{r}
#| label: nnt_arr
nnt_plot(
  nnt_dat,
  estimate_col = "arr",
  lower_col    = "arr_lower",
  upper_col    = "arr_upper"
) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_y_continuous(limits = c(0, 50),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Absolute Risk Reduction (%)") +
  theme_hv_poster()
```

## Survival difference

`survival_difference_plot()` plots the difference `S_2(t) - S_1(t)` between two
groups over time with an optional confidence band.
`sample_survival_difference_data()` generates the point-wise difference with
bootstrap CI bounds from a hazard-ratio contrast.

```{r}
#| label: surv_diff_data
diff_dat <- sample_survival_difference_data(
  n      = 500,
  groups = c("Control" = 1.0, "Treatment" = 0.70)
)
head(diff_dat)
```

### Survival difference curve

The dashed line at zero is the no-benefit baseline; a positive difference
means the treatment group has higher survival. No grouping argument is needed
for the two-group case.

```{r}
#| label: surv_diff_plot
survival_difference_plot(
  diff_dat,
  lower_col = "diff_lower",
  upper_col = "diff_upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(-5, 40),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival Difference (%)") +
  theme_hv_poster()
```

### Multiple comparisons

Combine several two-group differences into one long data frame and use
`group_col` to overlay them.

```{r}
#| label: surv_diff_multi
d1 <- sample_survival_difference_data(
  groups = c("Medical Mgmt" = 1.0, "TF-TAVR" = 0.70), seed = 1L
)
d1$comparison <- "TF-TAVR vs Medical Mgmt"

d2 <- sample_survival_difference_data(
  groups = c("TA-TAVR" = 0.90, "TF-TAVR" = 0.70), seed = 2L
)
d2$comparison <- "TF-TAVR vs TA-TAVR"

d3 <- sample_survival_difference_data(
  groups = c("AVR" = 0.80, "TF-TAVR" = 0.70), seed = 3L
)
d3$comparison <- "TF-TAVR vs AVR"

survival_difference_plot(rbind(d1, d2, d3), group_col = "comparison") +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Survival Difference (%)") +
  theme_hv_poster()
```
