# Postage Stamp Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

Sometimes overlaying every group on one panel turns into spaghetti: too many
colours, too many overlapping clouds, and the comparison you wanted is lost.
The fix is to give each group its own small panel and line them up on a shared
grid. The result is a "postage stamp" plot, the same plot repeated across the
levels of a categorical variable, each panel reading like a stamp on a sheet.
Because every panel shares the same axes and geometry, your eye compares them
the way it compares photographs in a contact sheet: the layout does the
alignment for you. Reach for it when one figure has to carry a per-subgroup
comparison and an overlay would be too busy.

`hvtiPlotR` has no helper for this layout, so we build it with base ggplot2's
[@R-ggplot2] `facet_wrap()` and apply the house theme `theme_hv_manuscript()` for
visual consistency. The faceting is the whole trick: you write one `ggplot()`
call describing a single panel, add `facet_wrap()`, and ggplot2 repeats the panel
once per group.

## The data it needs

A faceted scatter needs an x and a y for the relationship inside each panel, plus
a categorical column to split the panels on. We use
`hvtiPlotR::sample_eda_data()` with `op_years` (years since first surgery) for x
and `ef` (ejection fraction) for y, and we facet on operative `year`. A postage
stamp only earns its keep when there are *many* small panels, so we restrict to a
nine-year window, which lays out as a clean three-by-three sheet.

```{r}
#| label: data
dat <- sample_eda_data()
# A nine-year window facets into a 3 x 3 sheet of small multiples.
dat      <- dat[dat$year %in% 2012:2020, ]
dat$year <- factor(dat$year)
xvar <- "op_years"
yvar <- "ef"
grp  <- "year"
c(x = xvar, y = yvar, facet = grp, panels = nlevels(dat[[grp]]))
```

The `.data[[xvar]]` idiom is how ggplot2 reads a column name held in a variable,
so the recipe runs against whatever columns you name. Faceting on a window of
years turns one busy overlay into a contact sheet you read year by year;
substitute your own grouping (centre, severity class, era) and the mechanics are
identical.

## Build it

Each panel shows the relationship between the two numeric variables for one level
of the faceting variable, with a loess trend (a flexible local-regression curve)
to summarise the pattern within each subgroup. `facet_wrap()` lays the panels out
left to right and wraps to a new row when it runs out of width.

```{r}
#| label: postage-facet
#| fig-height: 5
ggplot(dat, aes(x = .data[[xvar]], y = .data[[yvar]])) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(vars(.data[[grp]]), ncol = 3) +
  labs(
    title = "Small-multiple panels by operative year",
    x = xvar,
    y = yvar
  ) +
  theme_hv_manuscript()
```

The key is that all nine panels share one x-axis and one y-axis. The smooth in
each panel is computed from that panel's data alone, so the curves are directly
comparable: a higher curve in one panel really does mean a higher y there, not a
trick of a rescaled axis.

## Read it

Read a postage stamp plot by comparing panels, not by studying any one in
isolation.

- **Compare on the shared scale.** The point of the layout is that the axes are
  fixed across panels, so a difference in height or slope between two panels is a
  real difference in the data. Trust the comparison only as long as the scales
  are shared; the moment you free the scales (see below) the panels no longer
  line up and cross-panel reading breaks.
- **Look at the trend per panel.** Each panel's smooth summarises that subgroup.
  Do the curves rise in some panels and fall in others? A pattern that differs by
  panel is exactly the interaction a single overlaid plot would have buried.
- **Watch the density of points.** A panel with few points (a year with fewer
  operations) gives a smooth you should not over-read; the curve is drawn with the
  same confidence as a panel backed by ten times the data.

## Variations

### Fixing the panel layout

`facet_wrap()` chooses a roughly square grid by default. When you want the panels
in a single row to emphasise a left-to-right comparison, or in a single column to
stack them, set `nrow` or `ncol` directly.

```{r}
#| label: postage-nrow
#| fig-height: 3.2
ggplot(dat, aes(x = .data[[xvar]], y = .data[[yvar]])) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(vars(.data[[grp]]), nrow = 1) +
  labs(
    title = "Panels in a single row",
    x = xvar,
    y = yvar
  ) +
  theme_hv_manuscript()
```

### Free scales, used deliberately

By default every panel shares both axes, which is what makes cross-panel reading
honest. Sometimes one group lives on a different range and the shared scale
squashes it flat. `scales = "free_y"` lets each panel set its own y-axis. Use it
only when you care about the shape inside each panel rather than comparing
heights across them, and say so in the caption, because the panels no longer line
up.

```{r}
#| label: postage-free
#| fig-height: 5
ggplot(dat, aes(x = .data[[xvar]], y = .data[[yvar]])) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(vars(.data[[grp]]), scales = "free_y") +
  labs(
    title = "Free y-axis per panel (compare shape, not height)",
    x = xvar,
    y = yvar
  ) +
  theme_hv_manuscript()
```

## Pitfalls

- **Free scales by accident.** The default shared scale is what makes the panels
  comparable. Switching to `scales = "free"` or `"free_y"` without flagging it
  lets a reader compare panel heights that are no longer on the same axis, which
  is the most common way a small-multiples figure misleads. Free the scales only
  on purpose, and label it.
- **Too many panels.** A facet over a high-cardinality variable produces a wall
  of tiny panels nobody can read. If a categorical has more than eight or so
  levels, collapse the rare ones into an "Other" group, or pick the levels worth
  showing, before you facet.
- **Unhelpful panel order.** `facet_wrap()` orders panels by the factor levels,
  which default to alphabetical. When the variable has a natural order (severity,
  era, anatomical class), reorder the factor first so the panels read in that
  order; otherwise the trend across panels is scrambled.
- **Over-reading a thin panel.** A subgroup with few observations gets a panel and
  a smooth that look just as authoritative as a well-populated one. Note the
  per-panel counts, or let the sparse panel speak for itself, rather than reading
  a confident-looking curve off a dozen points.
