# Postage Stamp Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

A "postage stamp" plot is a grid of small multiples: the same plot repeated
across the levels of one or more categorical variables, laid out in faceted
panels. Each small panel reads like a postage stamp on a sheet, and the grid lets
you compare many subgroups at a glance while holding the axes and geometry
constant. `hvtiPlotR` has no dedicated helper for this layout, so we use base
ggplot2's `facet_wrap()` and apply the house theme `theme_hv_manuscript()` for
visual consistency.

We use `hvtiPlotR::sample_eda_data()`, choosing two numeric columns for the x/y
relationship and a categorical column to facet on.

```{r}
#| label: data
dat <- sample_eda_data()
nums <- names(dat)[vapply(dat, is.numeric, logical(1))]
grp  <- names(dat)[vapply(dat, function(x) is.factor(x) || is.character(x), logical(1))][1]
xvar <- nums[2]
yvar <- nums[3]
c(x = xvar, y = yvar, facet = grp)
```

## Small multiples with a fitted trend

Each panel shows the relationship between two numeric variables for one level of
the faceting variable, with a loess trend to summarise the pattern within each
subgroup.

```{r}
#| label: postage-facet
#| fig-height: 5
ggplot(dat, aes(x = .data[[xvar]], y = .data[[yvar]])) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(vars(.data[[grp]])) +
  labs(
    title = "Small-multiple panels by group",
    x = xvar,
    y = yvar
  ) +
  theme_hv_manuscript()
```
