# Slides and presentations

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

A slide is read from across a room and sits on a coloured background, not white
paper. Two things change from a manuscript figure. The type scales up the way a
poster's does, so it carries to the back of the room; and the theme has to match
the deck — dark for a projected slide, light or transparent for a printed
handout or a light template. hvtiPlotR ships a pair of PowerPoint themes for
exactly this, plus `save_ppt()`, which drops the finished plot onto a slide as an
*editable* vector graphic rather than a flat image.

## A slide-styled plot

`theme_hv_ppt_dark()` styles a plot for a projected deck: white ink, no
gridlines, and a large base size so the figure reads from the back of the room.
It deliberately leaves the figure background *transparent* so the slide's own
background shows through — which is what you want on a real deck, but it makes the
white axis text vanish on this white page. For the preview below we add a dark
`plot.background` to stand in for the slide; on the actual deck, drop that line
and let `save_ppt()` place the transparent figure on the slide.

```{r}
#| label: build-ppt-dark
p <- plot(hv_survival(sample_survival_data(n = 500, seed = 42)))
p + theme_hv_ppt_dark(base_size = 18) +
  theme(plot.background = element_rect(fill = "grey15", colour = NA))
```

`theme_hv_ppt_light()` is the same typography on a light or transparent
background, for a printed handout or a light template.

```{r}
#| label: build-ppt-light
p + theme_hv_ppt_light(base_size = 18)
```

Both themes default to a 32 pt Arial base — sized for a slide projected across a
room. We drop `base_size` to 18 here so the figure previews legibly on the page;
leave it at the default for the real deck, and raise it further for a large hall.

## Editable PowerPoint

A figure exported as a PNG is frozen: a reviewer who wants the title reworded or
a colour swapped has to come back to you. `save_ppt()` instead writes the plot
onto a slide as a native vector graphic, so its text, lines, and colours stay
editable in PowerPoint itself. It anchors the plot in the panel rectangle of a
house slide template, so every figure lands in the same place across a deck.

```{r}
#| label: save-ppt
#| eval: false
save_ppt(
  p + theme_hv_ppt_dark(),
  template   = "RD-dark.pptx",
  powerpoint = "survival_slide.pptx"
)
```

The chunk is `eval: false` because it writes a binary `.pptx`. The code is shown,
not executed during the book build. Pair the dark theme with a dark `template`
and the light theme with a light one, and keep `theme_hv_ppt_dark()` at its 32 pt
default here — the template is laid out for that size. Pass `panel_box` to move or
resize the plot rectangle, and `slide_titles` to set the slide heading.
```
