# Publication tables

For a thorough treatment of effective display tables in `gt` [@R-gt], see
Rich Iannone's guide:
<https://rich-iannone.github.io/gt-effective-display-tables/>.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(gt)
library(dplyr)
library(hvtiPlotR)
```

This chapter builds publication-ready tables directly with the `gt`
package. The recipes here are deliberately self-contained; once a
dedicated table package matures, **these tables will migrate to the
planned hvtiRtables package** and the call sites will collapse to a
single helper.

## A "Table 1" baseline-characteristics summary

The canonical first table in any clinical manuscript is a
baseline-characteristics summary ("Table 1"): rows of patient
characteristics, columns split by treatment or exposure group. We build
one from `hvtiRutilities::generate_survival_data()`, which returns a
richly labelled patient-level cohort with a natural grouping variable
(`sex`).

```{r}
#| label: t1_data
dta <- hvtiRutilities::generate_survival_data(n = 200, seed = 42)
str(dta[, c("sex", "age", "bmi", "gfr_bs", "diabetes")])
```

### Summarise numeric variables as mean ± SD

We summarise three continuous baseline variables (age, BMI, baseline
eGFR) as mean and standard deviation within each `sex` group.

```{r}
#| label: t1_numeric
num_summary <- dta |>
  group_by(sex) |>
  summarise(
    n        = dplyr::n(),
    age_mean = mean(age),     age_sd = sd(age),
    bmi_mean = mean(bmi),     bmi_sd = sd(bmi),
    gfr_mean = mean(gfr_bs),  gfr_sd = sd(gfr_bs),
    .groups  = "drop"
  )
num_summary
```

### Summarise a categorical variable as n (%)

Diabetes status is summarised as a count and within-group percentage.

```{r}
#| label: t1_categorical
cat_summary <- dta |>
  group_by(sex) |>
  summarise(
    diabetes_n   = sum(diabetes == "Yes"),
    diabetes_pct = 100 * mean(diabetes == "Yes"),
    .groups      = "drop"
  )
cat_summary
```

### Render with `gt()`

We join the two summaries and present the result with a header, relabelled
columns, and consistent numeric formatting. Mean ± SD pairs are formatted
to one decimal place; the diabetes percentage to one decimal place.

```{r}
#| label: t1_gt
table1 <- num_summary |>
  left_join(cat_summary, by = "sex") |>
  gt(rowname_col = "sex") |>
  tab_header(
    title    = "Table 1. Baseline Characteristics by Sex",
    subtitle = "Simulated cohort, n = 200"
  ) |>
  cols_label(
    n            = "N",
    age_mean     = "Age (mean)",
    age_sd       = "Age (SD)",
    bmi_mean     = "BMI (mean)",
    bmi_sd       = "BMI (SD)",
    gfr_mean     = "eGFR (mean)",
    gfr_sd       = "eGFR (SD)",
    diabetes_n   = "Diabetes (n)",
    diabetes_pct = "Diabetes (%)"
  ) |>
  fmt_number(
    columns  = c(age_mean, age_sd, bmi_mean, bmi_sd,
                 gfr_mean, gfr_sd, diabetes_pct),
    decimals = 1
  ) |>
  tab_spanner(label = "Age (years)",  columns = c(age_mean, age_sd)) |>
  tab_spanner(label = "BMI (kg/m²)",  columns = c(bmi_mean, bmi_sd)) |>
  tab_spanner(label = "eGFR (mL/min)", columns = c(gfr_mean, gfr_sd))

table1
```

The table renders as static HTML in the book and as a publication-quality
PDF or Word object when the document is rendered to those formats.
