# Random forests: fit then visualize

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

A Kaplan-Meier curve answers one question at a time, and a Cox model asks you to
commit up front to which covariates enter, in what functional form, and which
ones interact. Outcomes data after cardiac surgery rarely behave that cleanly.
The effect of age on reoperation risk is not a straight line; the effect of an
implant type depends on the patient it went into; some predictors are continuous
(gradient, ejection fraction) and some are categorical (valve position, urgency),
and the interactions among them are exactly the part you did not anticipate.

Reach for a random forest when you want a model to find that structure for you.
A forest grows many decision trees on bootstrap resamples of the cohort and
averages their predictions. Because each tree splits the data on whatever
variable helps most at each node, the ensemble picks up nonlinear effects and
interactions without you specifying them, and it handles continuous and
categorical predictors side by side. You get one more thing for free: every
observation is left out of roughly a third of the trees, so the forest can
predict each patient using only the trees that never saw them. That OOB
(out-of-bag) prediction is an honest, cross-validation-like estimate of how the
model performs on data it was not trained on, and it costs nothing extra to
compute.

We treat a random forest analysis in this book as two steps, always in the same
order.

1. **Fit** a forest with `randomForestSRC::rfsrc()` [@R-randomForestSRC]. This is
   the modeling engine: it grows the trees, computes the OOB predictions, and
   (optionally) variable importance.
2. **Visualize** the fitted object with a `ggRandomForests` [@R-ggRandomForests]
   `gg_*()` *constructor*. Each constructor extracts a tidy slice of the forest,
   error rates, predicted survival, variable importance, partial dependence, and
   returns a lightweight object with `plot()` and `autoplot()` methods that
   produce `ggplot` graphics.

Because every `gg_*()` plot is a standard `ggplot`, the house manuscript style
is applied the same way as everywhere else in the book: add
`+ theme_hv_manuscript()`.

## The data it needs

`rfsrc()` works from a formula and a data frame, the same interface you already
use for `lm()` or `coxph()`. For a survival forest the left-hand side is a
`Surv(time, status)` object: a follow-up time and an event indicator (`1` =
event, `0` = censored), one row per patient. The right-hand side is the set of
predictors, and `~ .` means "use every other column." Predictors can be a mix of
numeric and factor variables; the forest does not need them centered, scaled, or
transformed.

The `veteran` lung-cancer trial data ships with `randomForestSRC` and is shaped
exactly this way, so the recipe runs end to end before you point it at your own
cohort.

```{r}
#| label: rf-data
data(veteran, package = "randomForestSRC")
head(veteran)
```

`time` and `status` are the outcome; the remaining columns (treatment, cell
type, Karnofsky score, age, and so on) are the predictors the forest will split
on.

## Build it

Start from the bare end-to-end pattern so you can see what the two steps produce
before any styling. We fit a survival forest, then look at how the OOB error
converges as trees are added. One argument matters here: pass `block.size = 1`
so the forest records the OOB error after *every* tree. The default records it
only at the final tree, which would leave `gg_error()` with a single point and
no convergence curve to read.

```{r}
#| label: rf-minimal
rf <- rfsrc(Surv(time, status) ~ ., data = veteran, ntree = 100, block.size = 1)
plot(gg_error(rf)) + theme_hv_manuscript()
```

That is the whole pattern: `rfsrc()` once, then a `gg_*()` constructor wrapped in
`plot()` and themed. Everything else in this part of the book is a variation on
those two lines. The survival curves, the importance rankings, and the
dependence plots all start from a fitted forest and differ only in which `gg_*()`
constructor reads it.

::: callout-note
## Tree count

Every `rfsrc()` call in these chapters uses `ntree = 100` to keep render times
short. For a real analysis use the package default (`ntree = 500` or more) and
confirm convergence with `gg_error()` before you trust anything downstream.
:::

## Read it

The figure above is the error-convergence curve, and it is the first thing to
look at after any fit. Read it as a settling process:

- **The drop on the left** is the ensemble learning. Each of the first few dozen
  trees still carries information the forest did not already have, so the error
  falls fast.
- **The flat tail on the right** means additional trees no longer change the
  prediction. The forest has converged, and the number of trees you grew is
  enough.
- **A tail still trending downward** at the right edge is the signal to refit
  with a larger `ntree`. The curve here is flat well before 100 trees, so for
  this small dataset 100 is adequate.

This is also your cue that the rest of the analysis is safe to read. Variable
importance and dependence plots are only as stable as the forest underneath
them; a forest that has not converged will give you rankings that shift if you
add more trees.

## Variations

Each chapter that follows takes the fitted forest and reads a different slice of
it. The constructor is the only thing that changes.

| Chapter | Constructor | Shows |
|---|---|---|
| Error convergence | `gg_error()` | OOB error vs. number of trees |
| Predicted survival | `gg_rfsrc()`, `gg_survival()` | per-observation and grouped survival |
| Variable importance | `gg_vimp()` | VIMP ranking of predictors |
| Variable dependence | `gg_variable()`, `gg_partial()` | marginal and partial effects |
| ROC / Brier | `gg_roc()` | classification performance |
| varPro | varPro constructors | rule-based variable selection |

The forest does not have to be a survival forest. Give `rfsrc()` a numeric
outcome and you get a regression forest; give it a factor and you get a
classification forest. The `gg_*()` constructors adapt to the family of the fit,
so the same `plot(gg_error(rf))` works whether the error is `1 - C-index`,
mean-squared error, or misclassification rate.

## Pitfalls

- **Forgetting `block.size = 1` for error curves.** The default records OOB error
  only at the final tree, so `gg_error()` collapses to a single point and you
  lose the convergence curve entirely. Set `block.size = 1` whenever you plan to
  inspect convergence.
- **Trusting the toy tree count.** The `ntree = 100` in these chapters is a
  rendering shortcut. A real analysis uses the default (500 or more) and checks
  `gg_error()` to confirm the forest actually settled before you quote any
  importance or dependence result.
- **Reading downstream plots before checking convergence.** A forest that is
  still improving at the right edge of the error curve gives you unstable
  rankings. Always glance at `gg_error()` first; it is the cheapest sanity check
  in the workflow.
