# Random forests: fit then visualize

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

Random forest analysis in this book follows a consistent two-step workflow.

1. **Fit** a forest with `randomForestSRC::rfsrc()` [@R-randomForestSRC]. This is the modeling
   engine: it grows the trees, computes out-of-bag (OOB) predictions, and
   (optionally) variable importance.
2. **Visualize** the fitted object with a `ggRandomForests` [@R-ggRandomForests] `gg_*()`
   *constructor*. Each constructor extracts a tidy slice of the forest --
   error rates, predicted survival, variable importance, partial dependence --
   and returns a lightweight object with `plot()` and `autoplot()` methods that
   produce `ggplot` graphics.

Because every `gg_*()` plot is a standard `ggplot`, the house manuscript style
is applied the same way as everywhere else in the book: add
`+ theme_hv_manuscript()`.

## A minimal end-to-end example

The `veteran` lung-cancer trial data ships with `randomForestSRC`. We fit a
survival forest and look at how the OOB error converges as trees are added.
Pass `block.size = 1` so the forest records the OOB error after *every* tree --
the default records it only at the final tree, which would leave `gg_error()`
with a single point and no convergence curve to read.

```{r}
#| label: rf-minimal
data(veteran, package = "randomForestSRC")
rf <- rfsrc(Surv(time, status) ~ ., data = veteran, ntree = 100, block.size = 1)
plot(gg_error(rf)) + theme_hv_manuscript()
```

That is the whole pattern: `rfsrc()` once, then a `gg_*()` constructor wrapped
in `plot()` and themed. The pattern is identical for the survival curves,
importance, and dependence plots in the chapters that follow.

::: callout-note
## Tree count

Every `rfsrc()` call in these chapters uses `ntree = 100` to keep render times
short. For real analyses use the package default (`ntree = 500` or more) and
confirm convergence with `gg_error()`.
:::

## What the following chapters cover

| Chapter | Constructor | Shows |
|---|---|---|
| Error convergence | `gg_error()` | OOB error vs. number of trees |
| Predicted survival | `gg_rfsrc()`, `gg_survival()` | per-observation and grouped survival |
| Variable importance | `gg_vimp()` | VIMP ranking of predictors |
| Variable dependence | `gg_variable()`, `gg_partial()` | marginal and partial effects |
| ROC / Brier | `gg_roc()` | classification performance |
| varPro | varPro constructors | rule-based variable selection |
