# Variable and partial dependence

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

Variable importance ranks predictors but says nothing about the *shape* of
their effect. Two complementary plots fill that gap:

- **Variable dependence** shows the forest's predicted value for every
  observation against an actual predictor value -- the raw, marginal picture.
- **Partial dependence** averages out all other predictors, isolating the
  effect of one variable on the prediction -- a smoother, model-level summary.

We reuse the `pbc` survival forest from the previous chapter.

```{r}
#| label: rf-fit
data(pbc, package = "randomForestSRC")
rf <- rfsrc(Surv(days, status) ~ ., data = pbc, ntree = 100,
            importance = TRUE)
```

For survival forests we evaluate dependence at a fixed follow-up time. The
median follow-up is a natural choice.

```{r}
#| label: tval
t_eval <- median(pbc$days)
t_eval
```

## Variable dependence

`gg_variable()`, from ggRandomForests [@R-ggRandomForests], returns one predicted survival probability per observation. For
a survival forest the `plot()` method needs a `time` to evaluate, and an `xvar`
selection. Setting `panel = TRUE` lays the chosen predictors out side by side;
each point is one patient and the blue curve is a loess smooth.

```{r}
#| label: gg-variable
#| fig-height: 4
gg_v <- gg_variable(rf)
plot(gg_v, xvar = c("bili", "age", "albumin"), time = t_eval,
     panel = TRUE) +
  theme_hv_manuscript()
```

Reading the smooths: predicted survival at `r t_eval` days falls steeply as
serum bilirubin rises, and declines with age, while higher albumin tracks with
better survival. These are the marginal trends actually present in the data --
including the scatter, which warns you how much spread underlies each trend.

## Partial dependence

Variable-dependence scatter is confounded by every other predictor. Partial
dependence removes that confounding: for a grid of `bili` values the forest
predicts each patient as if they had that value, then averages. The result is
the marginal effect of `bili` with everything else held at its observed
distribution.

`gg_partial_rfsrc()` drives `randomForestSRC`'s partial-prediction machinery
directly. It is the slow step -- we restrict it to a single variable at the
median time and a modest evaluation grid (`n_eval`) to keep the render fast.

```{r}
#| label: gg-partial
gg_p <- gg_partial_rfsrc(rf, xvar.names = "bili",
                         partial.time = t_eval, n_eval = 25)
plot(gg_p) + theme_hv_manuscript()
```

The partial-dependence curve is the *adjusted* version of the bilirubin panel
above: the same downward trend, but now free of the influence of age, albumin
and the rest. The flat scatter has collapsed into a single, interpretable
dose-response line -- the cleanest summary of how bilirubin moves predicted
survival.

## Practical notes

- **Pick a time.** Survival dependence is always read at a follow-up time;
  changing `time`/`partial.time` can change the shape, so report which you used.
- **Cost.** Partial dependence refits predictions across the grid. If it is too
  slow, lower `n_eval`, restrict to one variable, or reduce `ntree`.
- **Order of operations.** Use VIMP to choose *which* variables to inspect here;
  there is no point profiling a variable the forest ignores.
