# Random forest error convergence

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

A random forest is an *ensemble*: predictions are averaged over many trees.
The first practical question is always "how many trees are enough?" The
out-of-bag (OOB) error answers it directly. Each observation is predicted only
by the trees that did **not** see it during training, so the OOB error is an
honest, cross-validation-like estimate computed for free while the forest
grows.

## Fit the forest

We use the `veteran` survival data again and grow 100 trees. The `block.size = 1`
argument tells `rfsrc()` [@R-randomForestSRC] to record the OOB error after every tree rather than
only at the final tree (the default). Without it, `gg_error()` [@R-ggRandomForests] has a single
error value and the convergence curve collapses to one point.

```{r}
#| label: rf-fit
data(veteran, package = "randomForestSRC")
rf <- rfsrc(Surv(time, status) ~ ., data = veteran, ntree = 100, block.size = 1)
rf
```

## Plot the error curve

`gg_error()` pulls the cumulative OOB error -- for survival forests this is
`1 - C-index`, where lower is better -- after each additional tree. The result
plots as error against the number of trees.

```{r}
#| label: gg-error
plot(gg_error(rf)) + theme_hv_manuscript()
```

## Reading the curve

- The curve drops quickly while the first few dozen trees are added: each new
  tree still carries information the ensemble did not already have.
- It then **flattens into a tail**. A flat tail means additional trees no
  longer change the prediction -- the forest has converged and you have grown
  enough trees.
- If the tail were still trending downward at the right edge, that is your
  signal to refit with a larger `ntree`.

Here the curve is essentially flat well before 100 trees, so for this small
dataset 100 trees is adequate. Always glance at `gg_error()` before trusting
any downstream importance or dependence plot: those are only as stable as the
underlying forest.
