# Random forest error convergence

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

A random forest is an *ensemble*: its prediction is an average over many trees.
That raises the first practical question of every fit, before you read a single
importance ranking or survival curve. How many trees are enough? Too few and the
average is still jittering from tree to tree; the rankings and curves you draw
from it will shift if you refit. Enough, and the forest has settled into a stable
prediction you can build on.

Reach for `gg_error()` whenever you have just fit a forest and want to confirm it
converged. It reads the OOB (out-of-bag) error the forest recorded as it grew and
plots that error against the number of trees. Each observation is predicted only
by the trees that did *not* see it during training, so the OOB error is an
honest, cross-validation-like estimate computed for free while the forest grows.
The curve it produces is the single cheapest diagnostic in the random-forest
workflow, and it is the one to look at first.

## The data it needs

`gg_error()` does not work on a data frame. It works on a fitted `rfsrc` object,
and it needs that object to have recorded its error history. By default `rfsrc()`
records the OOB error only at the final tree, which leaves a single number and
nothing to plot. The fix is one argument at fit time: `block.size = 1` tells the
forest to record the OOB error after every tree.

We use the `veteran` survival data again, one row per patient with a follow-up
`time` and an event `status`, and grow 100 trees.

```{r}
#| label: rf-fit
data(veteran, package = "randomForestSRC")
rf <- rfsrc(Surv(time, status) ~ ., data = veteran, ntree = 100, block.size = 1)
rf
```

The printed summary confirms the family (survival), the number of trees, and the
error rate at the final tree. `gg_error()` [@R-ggRandomForests] turns that final
number into the whole history.

## Build it

With the forest fit and `block.size = 1` in place, the plot is one line.
`gg_error()` pulls the cumulative OOB error after each additional tree. For a
survival forest the error is `1 - C-index` (the concordance index, the
probability the model ranks a random pair of patients in the right order), so
lower is better.

```{r}
#| label: gg-error
plot(gg_error(rf)) + theme_hv_manuscript()
```

The result plots error on the y-axis against the number of trees on the x-axis.
`rfsrc()` [@R-randomForestSRC] did the work; `gg_error()` just reads it back.

## Read it

A convergence curve tells its story by its shape, not its level. Three things to
look at:

- **The drop on the left.** While the first few dozen trees are added, each new
  tree still carries information the ensemble did not already have, so the error
  falls quickly.
- **The flat tail on the right.** A tail that has gone flat means additional
  trees no longer change the prediction. The forest has converged, and you have
  grown enough trees.
- **A tail still sloping down.** If the curve is still trending downward at the
  right edge, the forest is not done improving. That is your signal to refit with
  a larger `ntree`.

Here the curve is essentially flat well before 100 trees, so for this small
dataset 100 trees is adequate. Make this glance a habit: check `gg_error()`
before you trust any downstream importance or dependence plot, because those are
only as stable as the forest underneath them.

## Variations

### A forest that has not converged

To see what an undersized forest looks like, refit with far too few trees. The
curve never reaches a flat tail; it is still descending at the right edge, which
is precisely the pattern that should send you back to a larger `ntree`.

```{r}
#| label: gg-error-short
rf_short <- rfsrc(Surv(time, status) ~ ., data = veteran, ntree = 25,
                  block.size = 1)
plot(gg_error(rf_short)) + theme_hv_manuscript()
```

Compare this to the 100-tree curve above. The shape, not the number on the axis,
is what tells you whether to grow more trees.

## Pitfalls

- **Forgetting `block.size = 1`.** This is the one that bites everyone. Without
  it, `rfsrc()` records OOB error only at the final tree, so `gg_error()` has a
  single value and the convergence curve collapses to one point. Set it at fit
  time whenever you intend to inspect convergence.
- **Reading the level instead of the shape.** The y-axis value depends on the
  outcome and the cohort; a `1 - C-index` of 0.4 is not "bad" in any absolute
  sense. What you are reading is whether the curve has flattened, not where it
  sits.
- **Mistaking OOB error for an external test.** The OOB estimate is honest and
  free, and it is the right tool for judging convergence. It is not a substitute
  for an external validation set when the cohort is small: with few patients the
  OOB error itself is noisy, and a flat curve confirms the forest settled, not
  that it will generalize to a new institution.
