# Predicted response and survival

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

For a survival forest, the prediction is not a single number but an entire
**survival curve** for each observation. `gg_rfsrc()` extracts these predicted
curves so they can be plotted directly. This chapter shows two views: the
per-observation curves, and a population summary stratified by a grouping
variable.

## Per-observation predicted survival

`gg_rfsrc(object)` returns one out-of-bag predicted survival curve per subject.
Plotting them all together shows the spread of predicted risk across the
cohort.

```{r}
#| label: rf-fit
data(veteran, package = "randomForestSRC")
rf <- rfsrc(Surv(time, status) ~ ., data = veteran, ntree = 100)
plot(gg_rfsrc(rf)) + theme_hv_manuscript()
```

Each line is one patient's predicted survival probability over time. The fan of
curves is the forest's picture of heterogeneity in the cohort: low-risk
subjects stay high, high-risk subjects drop early.

## Predicted survival by group

To compare groups, pass `by =` to `gg_rfsrc()`. It averages the predicted
curves within each level of the grouping variable, giving a model-based
analogue of stratified survival. We switch to the `pbc` primary biliary
cirrhosis trial, which has a randomized `treatment` variable.

```{r}
#| label: rf-by-group
data(pbc, package = "randomForestSRC")
rf2 <- rfsrc(Surv(days, status) ~ ., data = pbc, ntree = 100)
plot(gg_rfsrc(rf2, by = "treatment")) + theme_hv_manuscript()
```

The two treatment arms track closely, consistent with the trial's null
finding -- but now the comparison comes from the full multivariable forest, not
a univariate split.

## Aside: empirical survival with `gg_survival()`

`gg_rfsrc()` shows the *model's* predicted survival. When you want the raw
**empirical** (Kaplan-Meier-style) survival of the data instead, use
`gg_survival()`. It works directly on the data frame -- no forest required --
taking the time `interval`, the `censor` indicator, an optional `by` grouping,
and the `data`.

```{r}
#| label: gg-survival
gg_dta <- gg_survival(
  interval = "days",
  censor   = "status",
  by       = "treatment",
  data     = pbc
)
plot(gg_dta) + theme_hv_manuscript() +
  labs(y = "Survival probability", x = "Time (days)")
```

The empirical curves and the forest-predicted curves should tell a consistent
story; divergence between them is itself diagnostic of how much the other
covariates are doing in the model.
