# Predicted response and survival

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

Once the error curve says the forest has converged, the next question is what the
forest actually predicts. For a survival forest the answer is unusual: the
prediction is not a single number but an entire **survival curve** for each
observation. The model gives every patient their own estimated probability of
being event-free over time, conditioned on all of their covariates at once.

Reach for `gg_rfsrc()` when you want to see those predicted curves. It is the
view to use when the question is about heterogeneity (how much do predicted
risks vary across the cohort?) or about a group contrast that comes from the full
multivariable model rather than a single univariate split. Two displays cover
most needs: the per-observation curves, which show the spread of predicted risk
across patients, and a population summary that averages those curves within
levels of a grouping variable. We finish the chapter with a related tool,
`gg_survival()`, for the raw empirical survival you compare the model against.

## The data it needs

`gg_rfsrc()` reads a fitted `rfsrc` survival object, so the input is the same
formula and data frame you used for the error chapter: a `Surv(time, status)`
outcome (follow-up time plus a `1`/`0` event indicator) and the predictors on the
right-hand side. Nothing about this view needs `block.size = 1`; that argument
only matters for the error curve, so we drop it here.

```{r}
#| label: rf-fit
data(veteran, package = "randomForestSRC")
rf <- rfsrc(Surv(time, status) ~ ., data = veteran, ntree = 100)
rf
```

For the grouped and empirical views later in the chapter we switch to the `pbc`
primary biliary cirrhosis trial, which carries a randomized `treatment` variable
to stratify on.

## Build it

`gg_rfsrc(object)` returns one OOB (out-of-bag) predicted survival curve per
subject. Plotting them together gives the bare view: every patient's predicted
survival probability over time, drawn on one set of axes.

```{r}
#| label: fig-rf_predicted
#| fig-cap: "Out-of-bag predicted survival curves, one per patient, showing the spread of predicted risk across the cohort"
plot(gg_rfsrc(rf)) + theme_hv_manuscript()
```

That fan of curves is the forest's picture of the cohort. The pattern is the same
two lines as everywhere else in this part of the book: a `gg_*()` constructor on
the fitted forest, wrapped in `plot()` and themed.

## Read it

A panel of predicted curves rewards reading the spread, not any single line.

- **The width of the fan** is the model's sense of how much patients differ.
  Low-risk subjects stay high; high-risk subjects drop early. A tight bundle of
  curves says the covariates do not separate patients much; a wide fan says they
  do.
- **The overall level** is the cohort's typical survival. Read it the way you
  would a Kaplan-Meier curve, as a probability that starts near 1 and steps down.
- **The far-right tail** rests on the patients still at risk late in follow-up,
  which are usually few. Treat the spread there with the same caution you give the
  tail of any survival estimate.

Because these are OOB predictions, each curve was produced by trees that never saw
that patient, so the fan is an honest picture rather than a re-description of the
training data.

## Variations

### Predicted survival by group

To compare groups, pass `by =` to `gg_rfsrc()`. It averages the predicted curves
within each level of the grouping variable, giving a model-based analogue of
stratified survival. We use `pbc` and its randomized `treatment` arm.

```{r}
#| label: fig-rf_predicted-grouped
#| fig-cap: "Predicted survival averaged within each treatment arm of the pbc trial, the model-based analogue of stratified survival"
data(pbc, package = "randomForestSRC")
rf2 <- rfsrc(Surv(days, status) ~ ., data = pbc, ntree = 100)
plot(gg_rfsrc(rf2, by = "treatment")) + theme_hv_manuscript()
```

The two treatment arms track closely, consistent with the trial's null finding.
The difference from a univariate comparison is that this contrast comes out of the
full multivariable forest: it is the predicted survival by arm after the model has
accounted for every other covariate, not a raw split on treatment alone.

### Empirical survival with `gg_survival()`

`gg_rfsrc()` shows the *model's* predicted survival. When you want the raw
**empirical** (Kaplan-Meier-style) survival of the data instead, use
`gg_survival()` [@R-ggRandomForests]. It works directly on the data frame, no
forest required, taking the time `interval`, the `censor` indicator, an optional
`by` grouping, and the `data`.

```{r}
#| label: fig-rf_predicted-empirical
#| fig-cap: "Empirical Kaplan-Meier survival by treatment arm computed directly from the pbc data, for comparison with the forest's predicted curves"
gg_dta <- gg_survival(
  interval = "days",
  censor   = "status",
  by       = "treatment",
  data     = pbc
)
plot(gg_dta) + theme_hv_manuscript() +
  labs(y = "Survival probability", x = "Time (days)")
```

The empirical curves and the forest-predicted curves should tell a consistent
story. When they diverge, that gap is itself diagnostic: it shows how much the
other covariates are doing in the model, since the forest curves reflect the full
covariate set while the empirical curves reflect only the grouping variable.

## Pitfalls

- **Reading predicted curves as observed data.** The fan from `gg_rfsrc()` is the
  model's estimate, smoothed through the ensemble, not the raw event times. For
  the observed picture, plot `gg_survival()` alongside it and compare.
- **Over-reading a group contrast.** `by =` averages predicted curves within a
  level, but it does not test a difference. Closely tracking arms are consistent
  with a null effect; clearly separated arms are a starting point for a formal
  comparison, not the comparison itself.
- **Forgetting these are OOB on a small cohort.** The predicted curves are honest
  out-of-bag estimates, which is their strength. With a small cohort they are also
  noisy, and they are not a substitute for an external test set: a clean-looking
  fan still needs outside validation before you carry it to a new population.
