# ROC and Brier performance

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

A *classification* forest does not just label each patient, it returns a
probability for each class, and a probability is only useful if you can say how
trustworthy it is. Reach for these plots whenever you need to report how well a
classification forest performs: whether it can tell recurrence from
non-recurrence, how it compares against another model, or whether the predicted
probabilities are calibrated well enough to act on. They are the performance
half of any classification result, the figures a reviewer looks for right after
the model description.

We judge a probabilistic classifier the same way we judge any other:

- The **ROC curve** traces the trade-off between the true-positive rate
  (sensitivity, the share of true events you catch) and the false-positive rate
  (1 minus specificity, the share of non-events you wrongly flag) as the decision
  threshold sweeps from 0 to 1. The **area under the curve (AUC)** boils that
  whole curve down to one number: 0.5 is a coin flip, 1.0 is perfect separation.
- The **Brier score** measures calibration, the mean squared difference between
  predicted probabilities and observed outcomes. Lower is better.

Both come from ggRandomForests [@R-ggRandomForests], and a useful detail comes
for free: these are computed out-of-bag (from patients each tree never saw
during fitting), so they are honest estimates that need no separate test set.

## The data it needs

ROC reads from a fitted classification forest, one whose outcome is a factor so
the forest predicts class probabilities. We use the `breast` cancer data, where
`status` records recurrence with two levels.

```{r}
#| label: rf-fit
data(breast, package = "randomForestSRC")
levels(breast$status)
rf <- rfsrc(status ~ ., data = breast, ntree = 100)
rf$family
```

The `family` field confirms a classification forest, and the levels are `N` (no
recurrence) and `R` (recurrence). The clinically interesting event is
recurrence, so that is the class we will treat as positive.

## Build it

`gg_roc()` builds the out-of-bag ROC curve for one outcome. The `which_outcome`
argument names the class to treat as positive, and the important thing is that
it takes a *numeric index*, not the level name: `2` selects the second level,
`R`. Pass the string `"R"` and the call errors with "subscript out of bounds".

```{r}
#| label: gg-roc
plot(gg_roc(rf, which_outcome = 2)) + theme_hv_manuscript()
```

The curve bows toward the top-left corner. The further it pulls away from the
diagonal (the line a random classifier would trace), the better the forest
separates recurrence from non-recurrence, and the AUC printed on the panel
quantifies that gap.

## Read it

- **Distance from the diagonal is discrimination.** A curve hugging the top-left
  corner means the forest ranks events above non-events almost every time; a
  curve sagging toward the 45-degree line means it is barely better than
  guessing. The AUC is that distance as a single number, easy to compare across
  models.
- **AUC ranks, Brier calibrates.** AUC tells you whether the *ranking* is right,
  whether events get higher predicted probabilities than non-events. It says
  nothing about whether a predicted 0.8 actually happens 80% of the time. That is
  calibration, and it is what the Brier score checks. A forest can rank well (high
  AUC) yet be poorly calibrated, so read both.
- **Read per-class curves against each other.** In the multi-class plot below,
  the class whose curve sags closest to the diagonal is the one the forest finds
  hardest to separate from the rest. That is usually the more informative finding
  than the overall headline number.

## Variations

### Multi-class ROC

With more than two classes you draw one curve per class, each class treated as
positive in turn against all the others (one-vs-rest). The `per_class = TRUE`
flag does this in a single call. We illustrate on `iris`, where the three
species give three curves.

```{r}
#| label: gg-roc-iris
rf_iris <- rfsrc(Species ~ ., data = iris, ntree = 100)
plot(gg_roc(rf_iris, which_outcome = 1, per_class = TRUE)) +
  theme_hv_manuscript()
```

Each coloured curve is one species-versus-rest classifier. The well-separated
species hug the top-left corner; any class whose curve sags toward the diagonal
is the one the forest struggles to tell apart from the others.

### Brier score (on a survival forest)

The Brier score is a calibration measure, but in ggRandomForests `gg_brier()`
only supports right-censored *survival* forests. It errors on a classification
forest, so we cannot run it on the `breast` model above; instead we refit `pbc`
as a survival forest and plot the integrated Brier score over follow-up time,
the survival analogue of the calibration check we want for classification.

```{r}
#| label: gg-brier
data(pbc, package = "randomForestSRC")
rf_surv <- rfsrc(Surv(days, status) ~ ., data = pbc, ntree = 100)
plot(gg_brier(rf_surv)) + theme_hv_manuscript()
```

Lower curves mean better-calibrated survival predictions across time. Read it
alongside the error curve and AUC: discrimination (ROC and AUC) tells you
whether the ranking is right, while the Brier score tells you whether the
predicted probabilities themselves can be trusted.

## Pitfalls

- **`which_outcome` is a number, not a name.** It is the *index* of the class to
  treat as positive. Use `2` to select the second level; passing the level name
  like `"R"` throws "subscript out of bounds". Check `levels()` to confirm which
  index is the class you mean.
- **`gg_brier()` is survival-only.** It works on right-censored survival forests
  and errors on a classification forest, which is why the Brier figure here is
  built on a `pbc` survival fit rather than the `breast` classification fit used
  for ROC. Do not expect a Brier curve straight off your classifier.
- **Match the metric to the family.** ROC and AUC need a classification or
  probability forest; `gg_brier()` here targets survival. Picking the right tool
  for the forest family saves you a confusing error message.
- **OOB is the honest estimate, but it is not magic.** These curves come from
  out-of-bag patients, so they approximate test-set performance without holding
  data back. They still inherit the forest's variance, so confirm the forest has
  converged before quoting a headline AUC.
