# ROC and Brier performance

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

For a *classification* forest the prediction is a class probability, and we
judge it the same way we judge any probabilistic classifier:

- The **ROC curve** traces the trade-off between the true-positive rate
  (sensitivity) and the false-positive rate (1 - specificity) as the decision
  threshold sweeps from 0 to 1. The **area under the curve (AUC)** summarises it
  in a single number: 0.5 is random guessing, 1.0 is perfect separation.
- The **Brier score** measures calibration -- the mean squared difference
  between predicted probabilities and observed outcomes. Lower is better.

All of these are computed out-of-bag, so they are honest estimates that need no
separate test set.

## Fit a classification forest

We use the `breast` cancer data, where `status` records recurrence with two
levels.

```{r}
#| label: rf-fit
data(breast, package = "randomForestSRC")
levels(breast$status)
rf <- rfsrc(status ~ ., data = breast, ntree = 100)
rf$family
```

The outcome levels are `N` (no recurrence) and `R` (recurrence). The clinically
interesting event is recurrence.

## ROC curve

`gg_roc()`, from ggRandomForests [@R-ggRandomForests], builds the OOB ROC curve for one outcome. The `which_outcome`
argument is the *index* of the class to treat as positive (here `2` selects the
second level, `R`), not the level name.

```{r}
#| label: gg-roc
plot(gg_roc(rf, which_outcome = 2)) + theme_hv_manuscript()
```

The curve bows toward the top-left corner; the further from the diagonal, the
better the forest discriminates recurrence from non-recurrence. The AUC printed
on the panel quantifies that gap.

## Multi-class ROC

With more than two classes you draw one curve per class (each class treated as
positive in turn). The `per_class = TRUE` flag does this in one call; we
illustrate on `iris`.

```{r}
#| label: gg-roc-iris
rf_iris <- rfsrc(Species ~ ., data = iris, ntree = 100)
plot(gg_roc(rf_iris, which_outcome = 1, per_class = TRUE)) +
  theme_hv_manuscript()
```

Each coloured curve is one species-versus-rest classifier. The well-separated
species hug the top-left corner; any class whose curve sags toward the diagonal
is the one the forest finds hardest to distinguish.

## Brier score

The Brier score is a calibration measure. In `ggRandomForests` `gg_brier()`
operates on a right-censored *survival* forest, plotting the (integrated) Brier
score over follow-up time -- the survival analogue of the calibration check we
do for classification. We refit `pbc` as a survival forest to demonstrate.

```{r}
#| label: gg-brier
data(pbc, package = "randomForestSRC")
rf_surv <- rfsrc(Surv(days, status) ~ ., data = pbc, ntree = 100)
plot(gg_brier(rf_surv)) + theme_hv_manuscript()
```

Lower curves mean better-calibrated survival predictions across time. Read it
alongside the error curve and AUC: discrimination (ROC/AUC) tells you whether
the ranking is right, while the Brier score tells you whether the predicted
probabilities themselves can be trusted.

## Reading performance plots

- **AUC ranks, Brier calibrates.** A model can discriminate well (high AUC) yet
  be poorly calibrated (high Brier), and vice versa -- always look at both.
- **OOB by default.** These estimates come from observations each tree never
  saw, so they approximate test-set performance without holding data back.
- **Right forest, right tool.** ROC needs a classification/probability forest;
  `gg_brier()` here targets survival forests. Match the metric to the family.
