# Variable importance

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

Once a forest has converged (see the previous chapter), the next question is
*which variables drive the prediction?* The standard random forest answer is
**permutation variable importance** (VIMP). For each predictor, its values are
randomly permuted in the out-of-bag (OOB) sample, breaking the relationship
between that predictor and the outcome, and the forest is re-evaluated. The
increase in prediction error is the variable's importance: a large increase
means the forest relied heavily on that variable; a near-zero or negative VIMP
means the variable contributed little.

## Fit the forest

We use the primary biliary cirrhosis (`pbc`) survival data. VIMP is computed at
fit time, so we pass `importance = TRUE` -- this avoids a second pass (and an
auto-calculation warning) later.

```{r}
#| label: rf-fit
data(pbc, package = "randomForestSRC")
rf <- rfsrc(Surv(days, status) ~ ., data = pbc, ntree = 100,
            importance = TRUE)
rf
```

## Rank every predictor

`gg_vimp()` extracts the VIMP for every variable and orders them. The `plot()`
method draws a horizontal bar for each predictor, sorted from most to least
important, so the ranking reads top to bottom.

```{r}
#| label: gg-vimp
plot(gg_vimp(rf)) + theme_hv_manuscript()
```

Bars are coloured by the sign of the importance. A handful of clinical markers
-- bilirubin, albumin, prothrombin time -- separate clearly from the pack,
which is exactly what we would expect for liver disease.

## Show only the top predictors

For a manuscript figure you rarely want every variable. The `nvar` argument
keeps just the top-N, giving a clean, focused panel.

```{r}
#| label: gg-vimp-top
plot(gg_vimp(rf, nvar = 10)) + theme_hv_manuscript()
```

## Reading VIMP

- **Ranking, not effect size.** VIMP tells you *whether* a variable matters to
  the forest, not the *direction* or *shape* of its effect. For that, see the
  variable- and partial-dependence plots in the next chapter.
- **Sign matters.** A negative VIMP means permuting the variable *improved*
  OOB error -- a strong signal the variable is noise.
- **Stability.** VIMP inherits the forest's variability. Confirm the forest has
  converged (`gg_error()`) before trusting the ranking, and be cautious reading
  fine distinctions between adjacent low-importance variables.
