# Alluvial and Sankey plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## Alluvial plots

`hv_alluvial()`, from hvtiPlotR [@R-hvtiPlotR], prepares an alluvial
(Sankey-style) diagram using `ggalluvial` [@R-ggalluvial]
so you can see how patients flow between states across time points or treatment
stages. Each row of the input is a unique combination of axis values with an
associated patient count; `plot()` draws flows proportional to that count.

### Sample data

`sample_alluvial_data()` simulates 300 patients with pre-operative AR grade,
procedure type, and post-operative AR grade columns. The `axes` vector defines
the left-to-right display order; each unique combination of axis values becomes
one band, sized by its `freq` count.

```{r}
#| label: alluvial_data
dta_al  <- sample_alluvial_data(n = 300, seed = 42)
axes    <- c("pre_ar", "procedure", "post_ar")
head(dta_al)
```

### Bare plot

The bare panel shows flows between the three axis stages with uniform fill and
no colour, labels, or theme.

```{r}
#| label: alluvial_bare
al <- hv_alluvial(dta_al, axes = axes, y_col = "freq")
plot(al)
```

### Fill flows by pre-operative grade

Pass `fill_col` to the constructor to colour each flow band by a categorical
column. Here pre-operative AR grade colours the bands so you can trace how each
grade distributes across procedure types and post-operative outcomes.

```{r}
#| label: alluvial_filled
al_filled <- hv_alluvial(dta_al, axes = axes, y_col = "freq",
                         fill_col = "pre_ar")

plot(al_filled) +
  scale_fill_manual(
    values = c(None     = "steelblue",
               Mild     = "goldenrod",
               Moderate = "darkorange",
               Severe   = "firebrick"),
    name = "Pre-op AR"
  ) +
  scale_colour_manual(
    values = c(None     = "steelblue",
               Mild     = "goldenrod",
               Moderate = "darkorange",
               Severe   = "firebrick"),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = 1:3,
    labels = c("Pre-op AR", "Procedure", "Post-op AR"),
    expand = c(0.05, 0.05)
  ) +
  labs(y = "Patients (n)",
       title = "AV Regurgitation: Pre- to Post-operative") +
  theme_hv_poster()
```

### Two-axis before / after comparison

When you only need to compare two time points, drop the middle axis and pass
just two columns to `axes`. Use `axis_labels` to replace the raw column names,
and `annotate()` to call out the direction of change on the panel.

```{r}
#| label: alluvial_two_axis
al2 <- hv_alluvial(
  dta_al,
  axes        = c("pre_ar", "post_ar"),
  y_col       = "freq",
  fill_col    = "pre_ar",
  axis_labels = c("Pre-operative", "Post-operative")
)

plot(al2) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1,
                    name = "AR Grade") +
  scale_colour_brewer(palette = "RdYlGn", direction = -1,
                      guide = "none") +
  annotate("text", x = 1.5, y = 250,
           label = "Improvement after surgery",
           size = 3.5, fontface = "italic") +
  labs(y = "Patients (n)",
       title = "AV Regurgitation Before and After Surgery") +
  theme_hv_poster()
```

## Cluster stability Sankey plots

`hv_sankey()` prepares a Sankey diagram (via `ggsankey` [@R-ggsankey]) that lets you watch
how patient cluster assignments shift as you increase K — a visual test of
cluster stability. Each column represents one value of K (default K = 2 to 9);
each band shows the fraction of patients whose assignment changes between
consecutive K values. Node labels show the cluster letter and count.

### Sample data

`sample_cluster_sankey_data()` returns 300 patients with nine
cluster-assignment columns (`C2` through `C9`), where each column gives the PAM
cluster label at that K.

```{r}
#| label: sankey_data
dta_san <- sample_cluster_sankey_data(n = 300, seed = 42)
head(dta_san)
table(dta_san$C9)
```

### Default plot (K = 2 to 9)

`hv_sankey()` reads the nine cluster-assignment columns and builds the Sankey
data automatically. The default Set1 palette assigns one colour per cluster
letter. Wide bands that stay intact across K indicate a stable partition; heavy
crossing and fragmentation signal that K has grown past what the data supports.

```{r}
#| label: sankey_default
#| fig-width: 8
#| fig-height: 5
sk <- hv_sankey(dta_san)
plot(sk) +
  labs(title = "Cluster Stability: K = 2 to 9") +
  theme_hv_poster()
```

### Custom colour palette

Replace the default Set1 colours with a fully custom named vector. Names must
match the node labels in the data.

```{r}
#| label: sankey_custom_cols
#| fig-width: 8
#| fig-height: 5
my_cols <- c(
  A = "#1f77b4", B = "#ff7f0e", C = "#2ca02c", D = "#d62728",
  E = "#9467bd", F = "#8c564b", G = "#e377c2", H = "#7f7f7f",
  I = "#bcbd22"
)
sk_custom <- hv_sankey(dta_san, node_colours = my_cols)
plot(sk_custom) +
  labs(title = "Cluster Stability: K = 2 to 9") +
  theme_hv_poster()
```

### Subset of K values

Pass a shorter `cluster_cols` vector to the constructor to show only a range of
K — useful for focusing a manuscript panel on, say, K = 2 to 6.

```{r}
#| label: sankey_subset
#| fig-width: 6
#| fig-height: 5
sk_sub <- hv_sankey(dta_san, cluster_cols = paste0("C", 2:6))
plot(sk_sub) +
  labs(title = "Cluster Stability: K = 2 to 6") +
  theme_hv_poster()
```
