# Scatter plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

Before you fit anything, you want to look at the data. A continuous measurement
plotted against a time axis answers the most basic exploratory question we ask
of a longitudinal cohort: is this number drifting as the years go by? Reach for
the scatter plot whenever you have a continuous variable (ejection fraction, LV
mass, peak gradient) and a continuous reference axis (years since first surgery,
operation date) and you want to see the cloud of points and the trend through
them at the same time.

`hv_eda()` [@R-hvtiPlotR] is the exploratory workhorse. Point it at a column and
a reference axis and it picks the right display: when `y_col` is continuous it
returns a scatter plot with a LOESS smoother (locally weighted regression, a
flexible curve that follows the data without assuming a straight line) drawn
through the cloud. The same function draws bar charts for categorical columns,
but those live in their own chapter; here we stay with the single continuous
case. As with the rest of the package, `plot()` on the `hv_eda` object hands
back a bare ggplot. You finish it with colour scales, labels, and a house theme
using the usual `+`.

One detail that matters in practice: observations missing `y_col` are not
silently dropped. They are drawn as rug ticks along the x-axis, so you can see
where in time your measurement is missing rather than pretending the data are
complete.

## The data it needs

`hv_eda()` expects one row per observation with a continuous `x_col` (the
reference axis) and a continuous `y_col` (the measurement). Nothing else is
required. `sample_eda_data()` returns a mixed-type cohort of 300 patients;
here we use the continuous `op_years` (years since the patient's first surgery)
for x and `ef` (ejection fraction, the percent of blood the left ventricle
pumps with each beat) for y.

```{r}
#| label: eda_data
dta_eda <- sample_eda_data(n = 300, seed = 42)
head(dta_eda)
```

If you are not sure a column will be treated as continuous, ask the classifier
directly. `eda_classify_var()` reports `"Cont"` for continuous columns,
`"Cat_Num"` for numeric categories, and `"Cat_Char"` for character categories,
so you can confirm the plot type before you build it.

```{r}
#| label: eda_classify
sapply(dta_eda[c("op_years", "ef")], eda_classify_var)
```

## Build it

Start from the bare panel so you can see what the constructor produced before
any styling. Build the object with `x_col`/`y_col`; the `y_label` argument sets
the axis and legend text in place of the raw column name.

```{r}
#| label: eda_bare
plot(hv_eda(dta_eda, x_col = "op_years", y_col = "ef",
            y_label = "Ejection Fraction (%)"))
```

That is the raw material: the scatter cloud, the LOESS smooth through it, and
the rug of missing-EF observations along the bottom. Now layer on the house
style. We set a single fill colour (one variable, one colour), constrain the
axes to a sensible range, and add a caption that tells the reader what the
x-axis ticks mean.

```{r}
#| label: fig-scatter
#| fig-cap: "Ejection fraction against years from first surgery, with a LOESS smoother and a rug of observations missing EF"
plot(hv_eda(dta_eda, x_col = "op_years", y_col = "ef",
            y_label = "Ejection Fraction (%)")) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, 20)) +
  labs(x = "Years from First Surgery Year",
       caption = "Tick marks on x-axis: observations with missing EF") +
  theme_hv_poster()
```

Swap `theme_hv_poster()` for `theme_hv_manuscript()` when you want the journal
version at smaller text sizes.

## Read it

A scatter plot with a LOESS smooth gives you two things to read at once. Look
for:

- **The trend, not the points.** The smoother is the signal. Is the curve flat
  (the measurement is stable across the cohort), sloping (a drift you may need
  to adjust for), or bending (a non-linear pattern a straight line would miss)?
  In our example the LOESS line stays roughly level, which says ejection
  fraction is not drifting systematically with time since surgery.
- **The spread around the trend.** A tight cloud means the trend is well
  determined; a wide one means individual patients vary a lot and the smooth
  is an average you should not over-read.
- **The rug at the bottom.** Each tick is a patient whose EF is missing. If the
  ticks cluster in one region of time, your measurement is missing in a pattern,
  not at random, and the trend in that region rests on fewer points than the
  cloud suggests.

## Variations

### Tuning the LOESS smoother

`plot.hv_eda()` exposes `smooth_method` and `smooth_span` to control the
overlaid curve. `smooth_span` is the fraction of the data used in each local
fit: a smaller span follows local structure more closely (and wiggles more),
a larger span gives a flatter, more conservative trend. When you are unsure
whether a bump is real, compare two spans before you commit to a story.

```{r}
#| label: fig-scatter-span
#| fig-cap: "The same scatter with a narrower LOESS span (0.5), which follows local structure more closely than the default"
plot(hv_eda(dta_eda, x_col = "op_years", y_col = "ef",
            y_label = "Ejection Fraction (%)"),
     smooth_method = "loess", smooth_span = 0.5) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, 20)) +
  labs(x = "Years from First Surgery Year",
       y = "Ejection Fraction (%)") +
  theme_hv_poster()
```

## Pitfalls

- **Reading the points instead of the trend.** With a few hundred patients the
  cloud looks busy and your eye finds patterns that are not there. The LOESS
  curve is what you report; the scatter is context.
- **Over-fitting the span.** A small `smooth_span` will track every local dip,
  including dips driven by one or two patients. If a feature disappears when you
  widen the span, it was probably noise.
- **Trusting the ends of the curve.** LOESS is least stable at the extremes of
  the x-axis, where it has data on only one side. Treat the leftmost and
  rightmost bends with the same caution you would give a sparse tail.
- **Ignoring the rug.** Missing values become rug ticks, not gaps in the trend.
  If the ticks pile up in one time window, the smooth there is built on thin
  data even though the line looks just as confident.
