# Scatter Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

`hv_eda()` plots a column against a reference axis. When `y_col` is continuous
it returns a scatter plot with a LOESS smoother overlay; observations missing
`y_col` are drawn as rug ticks on the x-axis. `plot()` on the `hv_eda` object
returns a bare `ggplot` that you decorate with colour scales, labels, and
`theme_hv_manuscript()` (or `theme_hv_poster()`).

## Sample data

`sample_eda_data()` returns a mixed-type cohort of 300 patients. Here we use
the continuous `op_years` (x) and `ef` (ejection fraction, y) columns.

```{r}
#| label: eda_data
dta_eda <- sample_eda_data(n = 300, seed = 42)
head(dta_eda)
```

## Continuous scatter + LOESS

Build the object with `x_col`/`y_col`, then decorate. The `y_label` argument
sets the legend/title text in place of the raw column name.

```{r}
#| label: eda_continuous
plot(hv_eda(dta_eda, x_col = "op_years", y_col = "ef",
            y_label = "Ejection Fraction (%)")) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, 20)) +
  labs(x = "Years from First Surgery Year",
       caption = "Tick marks on x-axis: observations with missing EF") +
  theme_hv_poster()
```

## Tuning the LOESS smoother

`plot.hv_eda()` exposes `smooth_method` and `smooth_span` to control the
overlaid smoother. A smaller `smooth_span` follows local structure more
closely; a larger span gives a flatter trend.

```{r}
#| label: eda_continuous_smooth
plot(hv_eda(dta_eda, x_col = "op_years", y_col = "ef",
            y_label = "Ejection Fraction (%)"),
     smooth_method = "loess", smooth_span = 0.5) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, 20)) +
  labs(x = "Years from First Surgery Year",
       y = "Ejection Fraction (%)") +
  theme_hv_poster()
```
