# Spaghetti Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

You measured the same thing on the same patients more than once, and you want to
see how it moves. That is the case for a spaghetti plot. Each patient becomes one
line connecting their repeated measurements over time, and the tangle of lines (the
"spaghetti" the name promises) gives you the shape of the cohort: where the
values cluster, how much individual patients swing, whether the bundle drifts up
or down as the years pass. A serial echo measurement (aortic valve gradient,
valve area, regurgitation grade) is the natural candidate.

This chapter covers two related displays. The spaghetti plot, `hv_spaghetti()`,
keeps every patient's trajectory visible, so you see the spread as well as the
trend. Its companion, `hv_trends()`, throws the individual lines away and plots a
single summary per time point (a mean or median per year), which is what you want
when the question is the population-level drift and the per-patient detail is
noise. You reach for the spaghetti when individual variation is the story, and
for the trend curve when the annual summary is.

Both return a bare ggplot you finish with scales, labels, and a house theme.

## The data it needs

`hv_spaghetti()` expects long-format data: one row per measurement, with an id
column identifying the patient, a time column, and the value being tracked.
Optionally, a grouping column lets you colour trajectories by stratum.
`sample_spaghetti_data()` generates 150 patients with up to 6 observations each,
split by a named proportion vector that stands in for a sex stratification. Build
both the plain and the colour-stratified objects up front so the variants below
can reuse them.

```{r}
#| label: spaghetti_data
dta_sp <- sample_spaghetti_data(
  n_patients = 150,
  max_obs    = 6,
  groups     = c(Female = 0.45, Male = 0.55),
  seed       = 42L
)
head(dta_sp)

# Build S3 objects for reuse below
sp     <- hv_spaghetti(dta_sp)
sp_col <- hv_spaghetti(dta_sp, colour_col = "group")
```

`hv_trends()` works the other way: you hand it patient-level data and it computes
the annual summary internally, so you do not pre-aggregate. The `year_range` and
`groups` arguments set the study window and group structure.

## Build it

Start from the bare spaghetti panel so you can see what the constructor produced
before any styling: one thin trajectory per patient over time, no colour, no axis
limits, no theme.

```{r}
#| label: spaghetti_bare
p_sp <- plot(sp)
p_sp
```

Now set sensible axis limits and breaks with the usual scale layers, then add a
theme. The y-axis here covers the full range of AV mean gradient.

```{r}
#| label: spaghetti_unstrat_full
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

## Read it

A spaghetti plot is read as a cloud first and individual lines second. Look for:

- **The shape of the bundle.** Where the lines are dense is where most patients
  sit; where they fan out is where patients diverge. A bundle that drifts upward
  across the x-axis is a cohort whose measurement is rising over time, even if no
  single line makes that obvious.
- **The crossing lines.** A patient whose trajectory cuts across the bundle, from
  the bottom to the top or back, is a patient whose value changed a lot. A few of
  these are normal; a whole sheaf of steep lines means the measurement is volatile
  and a single summary will hide more than it shows.
- **Where the lines stop.** Trajectories that end early are patients lost to
  follow-up. If the lines that drop out tend to sit high or low in the bundle,
  your later time points are a biased sample of the early cohort, and the apparent
  trend there may be attrition rather than change.

## Variations

### Spaghetti, stratified by group

Build the object with `colour_col = "group"` so each patient's line is coloured by
its stratum, then assign colours with `scale_colour_manual()`. Now you can read
two bundles at once and compare their shapes.

```{r}
#| label: spaghetti_sex_strat
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

### Spaghetti, with a LOESS overlay

When the bundle is too dense to read by eye, pass `add_smooth = TRUE` to overlay
a LOESS trend line per group (locally weighted regression, a flexible curve that
follows the data without assuming a straight line). The individual trajectories
stay as faint context while the smooth carries the trend.

```{r}
#| label: spaghetti_smooth
plot(sp_col, add_smooth = TRUE) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

### Temporal trend: a single annual series

When the per-patient detail stops being the point, `hv_trends()` collapses the
cohort to one summary per year. Subset to a single group and fit with
`group_col = NULL` for a single annual series. `hv_trends()` plots the annual mean
of the continuous `value` column, so set the x-axis to the study window and let
the y-axis span the observed range of the mean (here roughly 20 to 70). One
detail to watch, called out again in the pitfalls below: setting `limits` tighter
than the data, for instance `c(0, 10)`, silently drops every point and line and
leaves a blank panel. The `c(0, 80)` limits here comfortably contain the series.

```{r}
#| label: trends_data
dta_tr <- sample_trends_data(
  n          = 600,
  year_range = c(1968L, 2000L),
  groups     = c("Group I", "Group II", "Group III", "Group IV")
)
head(dta_tr)
```

```{r}
#| label: trends_n_year
one_grp <- dta_tr[dta_tr$group == "Group I", ]
tr1 <- hv_trends(one_grp, group_col = NULL)

plot(tr1) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(0, 80),      breaks = seq(0, 80, 20)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript()
```

### Temporal trend: multiple groups

When `group_col` is set, the constructor computes per-group annual means and
`plot()` draws one line per group. Pairing `scale_colour_brewer()` with
`scale_shape_manual()` gives each group both a distinct colour and a distinct
marker, so the figure stays readable when it is printed or photocopied in
greyscale.

```{r}
#| label: trends_multi_brewer
tr <- hv_trends(dta_tr)
plot(tr) +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15L, "Group II" = 19L,
               "Group III" = 17L, "Group IV" = 18L),
    name = "Group"
  ) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript()
```

### Temporal trend: with a confidence ribbon

Pass `se = TRUE` to add a `geom_ribbon()` around the mean line, and `alpha` to
control its opacity. The ribbon turns a bare trend line into an honest one: it
shows where the annual mean is well determined and where it rests on too few
patients to trust.

```{r}
#| label: trends_ribbon
plot(tr1, se = TRUE, alpha = 0.2) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript()
```

## Pitfalls

- **Axis limits that blank the panel.** This is the one that catches everyone.
  `scale_*_continuous(limits = ...)` does not zoom, it *filters*: any point
  outside the limits is dropped before plotting. Set the limits tighter than the
  data, say `c(0, 10)` on a series that runs to 70, and the trend line silently
  vanishes, leaving an empty panel with no warning. When a panel comes back blank,
  check the limits against the data range first. Use `coord_cartesian()` when you
  want to zoom without discarding points.
- **Too many lines.** Past a few hundred patients the spaghetti turns into a solid
  block and the individual trajectories stop being legible. At that point lower
  the line `alpha`, or switch to the trend curve, which is built for exactly this.
- **Reading a trend line without its ribbon.** A bare `hv_trends()` line looks
  equally confident at every year, but the early and late years of a study often
  rest on a handful of patients. Add `se = TRUE` so the reader can see where the
  estimate is thin.
- **Mistaking attrition for change.** Both displays connect only the patients who
  were measured. If sicker patients drop out, the surviving bundle drifts toward
  the healthy end, and the trend line follows. A trend is only a trend if the same
  kind of patient is being measured throughout.
