# Spaghetti Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## Spaghetti / profile plot

`hv_spaghetti()`, from hvtiPlotR [@R-hvtiPlotR], draws one trajectory line per
subject over time, with optional stratification by a grouping variable. `plot()` accepts `add_smooth = TRUE` for
an optional LOESS overlay and returns a bare ggplot. `sample_spaghetti_data()`
generates 150 patients with up to 6 observations each, stratified by a named
proportion vector.

```{r}
#| label: spaghetti_data
dta_sp <- sample_spaghetti_data(
  n_patients = 150,
  max_obs    = 6,
  groups     = c(Female = 0.45, Male = 0.55),
  seed       = 42L
)
head(dta_sp)

# Build S3 objects for reuse below
sp     <- hv_spaghetti(dta_sp)
sp_col <- hv_spaghetti(dta_sp, colour_col = "group")
```

### Bare plot

The bare `plot(sp)` panel draws one thin trajectory per patient over time with
no colour, axis limits, or theme.

```{r}
#| label: spaghetti_bare
p_sp <- plot(sp)
p_sp
```

### Unstratified

Set axis limits and breaks with the usual scale layers, then add a theme.

```{r}
#| label: spaghetti_unstrat_full
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

### Stratified by group

Build the object with `colour_col = "group"` so each subject's line is coloured
by its stratum, then assign colours with `scale_colour_manual()`.

```{r}
#| label: spaghetti_sex_strat
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

### With LOESS smooth overlay

Pass `add_smooth = TRUE` to `plot()` to add a LOESS trend line per group over
the individual trajectories.

```{r}
#| label: spaghetti_smooth
plot(sp_col, add_smooth = TRUE) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

## Temporal trend plot

`hv_trends()` accepts patient-level data and computes annual summaries (mean or
median) internally. `plot()` returns a bare ggplot; add scales, labels, and a
theme. `sample_trends_data()` generates patient-level data that `hv_trends()`
aggregates by year; the `year_range` and `groups` arguments mirror the study
period and group structure.

```{r}
#| label: trends_data
dta_tr <- sample_trends_data(
  n          = 600,
  year_range = c(1968L, 2000L),
  groups     = c("Group I", "Group II", "Group III", "Group IV")
)
head(dta_tr)
```

### Single variable: annual mean

Subset to a single group and fit with `group_col = NULL` for a single annual
series. `hv_trends()` plots the annual mean of the continuous `value` column, so
set the x-axis to the study window and let the y-axis span the observed range of
the mean (here roughly 20--70). Setting `limits` tighter than the data — e.g.
`c(0, 10)` — silently drops every point and line, leaving a blank panel.

```{r}
#| label: trends_n_year
one_grp <- dta_tr[dta_tr$group == "Group I", ]
tr1 <- hv_trends(one_grp, group_col = NULL)

plot(tr1) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(0, 80),      breaks = seq(0, 80, 20)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript()
```

### Multiple groups

When `group_col` is set, the constructor computes per-group annual means and
`plot()` draws one line per group. `scale_colour_brewer()` and
`scale_shape_manual()` give each group a distinct colour and marker, keeping
the figure readable in greyscale.

```{r}
#| label: trends_multi_brewer
tr <- hv_trends(dta_tr)
plot(tr) +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15L, "Group II" = 19L,
               "Group III" = 17L, "Group IV" = 18L),
    name = "Group"
  ) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript()
```

### With confidence ribbon

Pass `se = TRUE` to `plot()` to add a `geom_ribbon()` around the mean line; the
`alpha` argument controls ribbon opacity.

```{r}
#| label: trends_ribbon
plot(tr1, se = TRUE, alpha = 0.2) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript()
```
