# Survival Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

The Kaplan-Meier curve is the workhorse of every CORR outcomes paper. Reach for
it whenever the question is *how long until an event* and some patients have not
had the event yet by the end of follow-up: death, reoperation, readmission.
That last part is what makes survival analysis its own method: a patient who is
alive at last contact is **right-censored**, not a non-event, and the
product-limit estimator is what lets those partial observations still count.

`hv_survival()` [@R-hvtiPlotR], built on the survival package [@R-survival],
estimates the product-limit survival function once and stores everything five
related panels need (the survival curve, cumulative hazard, hazard rate,
log-log check, and integrated survivorship), plus tidy summary tables in
`$tables`. If you have used the SAS `%kaplan` macro, these are the same five
plots it draws. You build the S3 object once and call `plot()` with a `type`
argument to pull out the panel you want; each one comes back as a bare ggplot
you finish with the usual `+`.

## The data it needs

`hv_survival()` expects long-format data: one row per patient with a follow-up
`time` and an `event` indicator (`1` = event, `0` = censored). Nothing else is
required for an unstratified curve; add a grouping column when you want curves
per arm. `sample_survival_data()` builds a simulated right-censored cohort in
exactly this shape, so the recipe runs end to end before you point it at your
own data.

```{r}
#| label: km_data
dta_km <- sample_survival_data(n = 500, seed = 42)
head(dta_km)

# Build the S3 object once; reuse it across every panel below.
km <- hv_survival(dta_km)
```

## Build it

Start from the bare panel so you can see what the constructor produced before
any styling. The default `type = "survival"` is the curve with a logit-transform
95% confidence ribbon.

```{r}
#| label: km_bare
plot(km)
```

Then layer on the house style. The pattern below is the one we use most often in
CORR: a single steelblue for a one-cohort figure, percent labels on the y-axis,
an `n =` callout in the lower-left, and a poster-sized theme. Swap
`theme_hv_poster()` for `theme_hv_manuscript()` when you want the journal
version.

```{r}
#| label: km_styled
plot(km, alpha = 0.8) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)",
       title = "Overall Survival") +
  annotate("text", x = 1, y = 5, label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  theme_hv_poster()
```

## Read it

A Kaplan-Meier curve starts at 100% and steps downward at each event time; it
never rises. A few things to look at:

- **The ribbon widens to the right.** As the at-risk population thins, each
  remaining event moves the estimate more, so the confidence band fans out. Be
  cautious reading the far-right tail, which may rest on a handful of patients.
- **Median survival** is where the curve crosses 50%. If it never crosses, the
  median is not reached within follow-up and you report a different landmark
  (e.g. 5-year survival) instead.
- **The numbers at risk** below the figure tell the reader how much data backs
  each region of the curve. Always show them.

The at-risk strip and the per-time-point report (point estimates with CIs) are
stored on the object, so you never recompute them:

```{r}
#| label: km_tables
km$tables$risk
km$tables$report
```

## Numbers at risk

The "always show them" advice has a one-call answer. `hv_atrisk()` turns the
object's risk table into an aligned panel, and `hv_atrisk_compose()` stacks it
under the curve, matching the table's x-range to the plot's so the counts sit
beneath the time points they describe. Style the curve as usual, build the table
from the same object, and compose the two.

```{r}
#| label: km_atrisk
#| fig-height: 5
curve <- plot(km, alpha = 0.8) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_manuscript()

hv_atrisk_compose(curve, hv_atrisk(km))
```

The counts under the axis are the same `km$tables$risk` numbers from above, now
positioned at the curve's own time ticks. A stratified figure gives the table
one row per group, labelled to match the curves.

## Variations

### Stratified by group

To compare arms, pass `group_col` and the constructor fits a separate curve per
stratum. Here we simulate a two-arm valve-type cohort with a 1.4x hazard ratio.
Look for clearly separated curves whose CI ribbons stop overlapping in the
windows where the treatment difference matters.

```{r}
#| label: km_strata
dta_km_s <- sample_survival_data(
  n             = 500,
  strata_levels = c("Type A", "Type B"),
  hazard_ratios = c(1, 1.4),
  seed          = 42
)
km_s <- hv_survival(dta_km_s, group_col = "valve_type")

plot(km_s, alpha = 0.8) +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  scale_fill_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)",
       title = "Survival by Valve Type") +
  theme_hv_poster()
```

### Cumulative hazard

`type = "cumhaz"` gives the Nelson-Aalen cumulative hazard
`H(t) = -log S(t)`, a monotonically non-decreasing curve whose slope at any
point is the instantaneous hazard rate. Same object, different `type`.

```{r}
#| label: km_cumhaz
plot(km, type = "cumhaz") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = "Cumulative Hazard H(t)",
       title = "Nelson-Aalen Cumulative Hazard") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  theme_hv_poster()
```

### Hazard rate

`type = "hazard"` is the instantaneous hazard. The raw point estimates are
noisy, so for a figure overlay a LOESS smooth. The smoothed curve shows the
early-risk peak and later plateau a reviewer is looking for.

```{r}
#| label: km_hazard
plot(km, type = "hazard") +
  geom_smooth(aes(x = mid_time, y = hazard, color = strata),
              method = "loess", se = FALSE, span = 0.6) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = "Instantaneous Hazard",
       title = "Hazard Rate") +
  theme_hv_poster()
```

### Log-log survival (proportional-hazards check)

`type = "loglog"` plots `log(-log S(t))` against `log(time)`. Parallel lines
across strata mean the proportional-hazards assumption is reasonable. Run this
check before you quote a single hazard ratio. Fit with `group_col` so
there are strata to compare.

```{r}
#| label: km_loglog
plot(km_s, type = "loglog") +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  labs(x = "log(Years after Operation)", y = "log(-log S(t))",
       title = "Log-Log Survival") +
  theme_hv_poster()
```

### Integrated life table (restricted mean survival)

`type = "life"` is the integrated survivorship: the area under the survival
curve up to time `t`, equal to the restricted mean survival time (RMST) at that
horizon. It rises linearly while `S(t)` is near 1 and bends as events
accumulate. RMST is the summary to reach for when the log-log check above shows
hazards are *not* proportional and a single hazard ratio would mislead.

```{r}
#| label: km_life
plot(km, type = "life") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation",
       y = "Restricted Mean Survival (years)",
       title = "Integral of Survivorship") +
  theme_hv_poster()
```

## Pitfalls

- **Event coding.** `event` must be `1` for the event and `0` for censored. Flip
  them and you estimate freedom from *being censored*, which is meaningless. Check
  the coding before you trust the curve.
- **Reading the tail.** The right end of the curve often rests on very few
  patients; the widening ribbon is the warning. Quote a landmark estimate (with
  its at-risk count) rather than the visual endpoint.
- **Non-proportional hazards.** If the log-log curves cross, a single hazard
  ratio is the wrong summary -- report RMST from the integrated panel instead.
