# Survival Plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

`hv_survival()` estimates the Kaplan-Meier product-limit survival function and
stores the data for five companion panels plus tidy summary tables in
`$tables`. Build the S3 object once, then call `plot()` with the `type`
argument to render a bare ggplot for the selected panel (default
`"survival"`). Decorate each panel with scales, labels, and a theme.
Confidence intervals use the logit transform at 0.95 by default.

## Sample data

`sample_survival_data()` builds a simulated cohort in the long format
`hv_survival()` expects: one row per patient with `time` and `event` columns.
Build the object once and reuse it across every variant below.

```{r}
#| label: km_data
dta_km <- sample_survival_data(n = 500, seed = 42)
head(dta_km)

# Build the S3 object once; reuse across all plot variants below
km <- hv_survival(dta_km)
```

## Survival curve

The default panel is the survival curve with the logit-transform 95% CI
ribbon. Layer scales, labels, and a theme onto the bare plot with `+`.

```{r}
#| label: km_styled
plot(km, alpha = 0.8) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)",
       title = "Overall Survival") +
  annotate("text", x = 1, y = 5, label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  theme_hv_poster()
```

The at-risk strip and the per-time-point report (point estimates with CIs)
are stored in `$tables`; read either out directly with `km$tables$risk` or
`km$tables$report`.

## Cumulative hazard

`type = "cumhaz"` gives the Nelson-Aalen cumulative hazard `H(t) = -log S(t)`,
a monotonically non-decreasing curve whose slope is the instantaneous hazard.

```{r}
#| label: km_cumhaz
plot(km, type = "cumhaz") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = "Cumulative Hazard H(t)",
       title = "Nelson-Aalen Cumulative Hazard") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  theme_hv_poster()
```

## Log-log survival (proportional-hazards check)

`type = "loglog"` plots `log(-log S(t))` against `log(time)`. Parallel lines
across strata indicate proportional hazards. To see two strata, fit with
`group_col`.

```{r}
#| label: km_loglog
dta_km_s <- sample_survival_data(
  n             = 500,
  strata_levels = c("Type A", "Type B"),
  hazard_ratios = c(1, 1.4),
  seed          = 42
)
km_s <- hv_survival(dta_km_s, group_col = "valve_type")

plot(km_s, type = "loglog") +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  labs(x = "log(Years after Operation)", y = "log(-log S(t))",
       title = "Log-Log Survival") +
  theme_hv_poster()
```

## Integrated life table (restricted mean survival)

`type = "life"` is the integrated survivorship — the area under the survival
curve up to time `t`, equal to restricted mean survival time at that horizon.
It rises linearly while `S(t) ~ 1` and bends as events accumulate.

```{r}
#| label: km_life
plot(km, type = "life") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation",
       y = "Restricted Mean Survival (years)",
       title = "Integral of Survivorship") +
  theme_hv_poster()
```
