# Parametric hazard modeling with TemporalHazard

`TemporalHazard` [@R-TemporalHazard] is the team's R port of the SAS/C `HAZARD`
module that Blackstone and colleagues wrote for cardiac-surgery outcomes
research. It does two jobs. The first is the nonparametric pair you already know
from the survival chapter: Kaplan-Meier survival and the Nelson-Aalen cumulative
hazard, both built on the survival package [@R-survival]. The second is the part
that makes the package worth its own chapter: a *parametric* multi-phase hazard
fit, the classic early/constant/late decomposition of risk over time.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(TemporalHazard)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

A Kaplan-Meier curve answers "how many are still event-free at time *t*." A
hazard curve answers a different and often more clinically interesting question:
"how dangerous is *right now*." After a cardiac operation that instantaneous
risk is rarely flat. It spikes in the first weeks (the surgical insult, the
early failures), settles into a long low plateau, then climbs again years later
as the cohort ages. Reach for a parametric hazard model when you want to *see*
those phases as a smooth curve rather than the noisy step function a raw event
count gives you, and when you want each phase summarised by a parameter you can
report and compare across cohorts.

The other reason to reach for this package is lineage. For decades the team
described risk with the SAS-HAZARD module, and reviewers in this field expect
that early-peak, constant-phase, late-rise vocabulary. `TemporalHazard` keeps
the same decomposition in R, so a figure you build here reads the way the
reviewers already think. You fit the model once with `hazard()`, then ask it for
survival, cumulative hazard, or the instantaneous hazard over whatever time grid
you like.

## The data it needs

Everything in this chapter runs on data bundled with the package, so the recipe
works before you point it at your own extract. For the survival and hazard
sections we use `cabgkul`: 5,880 patients with a follow-up time `int_dead`
(interval to death, in years) and an event indicator `dead` (`1` = death,
`0` = censored). That is the same long format the survival chapter used, one row
per patient with a time and a 0/1 event. The competing-risks section switches to
`omc`, which records thromboembolic events and death on the same patients.

```{r}
#| label: load-data
data(cabgkul, package = "TemporalHazard")
str(cabgkul)
```

## Build it

### Nonparametric estimates first

Before fitting anything parametric, look at the empirical curves. They are the
yardstick every parametric fit is later checked against, and they make no
distributional assumption. `hzr_kaplan()` returns a data frame (class
`hzr_kaplan`) with one row per event time and the columns you need to draw a
product-limit curve: `time`, `n_risk`, `n_event`, `survival`, the
logit-transformed band `cl_lower`/`cl_upper`, and the derived `cumhaz`,
`hazard`, `density`, and `life`.

```{r}
#| label: fig-temporal_hazard-km
#| fig-cap: "Empirical Kaplan-Meier survival for the cabgkul cohort with its logit-transformed confidence band"
km <- hzr_kaplan(cabgkul$int_dead, cabgkul$dead)
head(km)

ggplot(km, aes(x = time, y = survival)) +
  geom_ribbon(aes(ymin = cl_lower, ymax = cl_upper), alpha = 0.2, fill = "steelblue") +
  geom_step(colour = "steelblue", linewidth = 0.7) +
  labs(x = "Years since operation", y = "Survival probability",
       title = "Kaplan-Meier survival (cabgkul)") +
  theme_hv_manuscript()
```

`hzr_nelson()` gives the Nelson-Aalen cumulative hazard (class `hzr_nelson`):
`time`, `n_risk`, `n_event`, `weight_sum`, `cumhaz` with its lognormal limits
`cl_lower`/`cl_upper`, `hazard`, and a running `cum_events`. The cumulative
hazard is the same information as the survival curve turned the other way up; its
slope at any point *is* the instantaneous hazard, which is what the parametric
fit below will try to summarise.

```{r}
#| label: fig-temporal_hazard-nelson
#| fig-cap: "Empirical Nelson-Aalen cumulative hazard for the cabgkul cohort with lognormal confidence limits"
na <- hzr_nelson(cabgkul$int_dead, cabgkul$dead)
head(na)

ggplot(na, aes(x = time, y = cumhaz)) +
  geom_ribbon(aes(ymin = cl_lower, ymax = cl_upper), alpha = 0.2, fill = "firebrick") +
  geom_step(colour = "firebrick", linewidth = 0.7) +
  labs(x = "Years since operation", y = "Cumulative hazard",
       title = "Nelson-Aalen cumulative hazard (cabgkul)") +
  theme_hv_manuscript()
```

### The parametric fit

`hazard()` fits the parametric model when you pass `fit = TRUE`. The one
argument you must not skip is `theta`, the starting values for the parameter
vector. The optimiser needs somewhere to begin, and there is no safe default it
can guess for you, so leaving `theta` out makes the fit (and every later
`predict()`) error. For a Weibull the two parameters are the scale `mu` and the
shape `nu`; `theta = c(0.01, 1.0)` is a reasonable start for both. The returned
object (class `hazard`) is a list with `call`, `spec`, `data`, `fit`,
`legacy_args`, and `engine`; the fitted parameters live in `fit$fit$theta`.

```{r}
#| label: fit
fit <- hazard(time = cabgkul$int_dead, status = cabgkul$dead,
              theta = c(0.01, 1.0), dist = "weibull", fit = TRUE)
summary(fit)
```

`predict.hazard()` evaluates the fitted model over any time grid you hand it in
`newdata`. For `type = "survival"` and `type = "cumulative_hazard"` it returns a
plain numeric vector lined up with the `time` column. The instantaneous hazard
takes one extra step. This model is intercept-only, so `type = "hazard"` returns
the constant relative hazard `exp(eta) = 1` rather than the curve we want; the
honest way to recover the instantaneous rate is to difference the cumulative
hazard, since the hazard is by definition its slope.

```{r}
#| label: predict
t_grid <- seq(0.5, max(cabgkul$int_dead), length.out = 120)
nd <- data.frame(time = t_grid)

pred <- data.frame(
  time     = t_grid,
  survival = predict(fit, newdata = nd, type = "survival"),
  cumhaz   = predict(fit, newdata = nd, type = "cumulative_hazard")
)
# Instantaneous hazard = d(cumhaz)/dt
pred$hazard <- c(NA, diff(pred$cumhaz) / diff(pred$time))
str(pred)
```

The fitted model gives us a smooth survival curve, its cumulative hazard, and
the instantaneous hazard rate, one panel each.

```{r}
#| label: fig-temporal_hazard-parametric-survival
#| fig-cap: "Parametric Weibull survival curve from the fitted model"
ggplot(pred, aes(x = time, y = survival)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  labs(x = "Years since operation", y = "Survival probability",
       title = "Parametric (Weibull) survival") +
  theme_hv_manuscript()
```

```{r}
#| label: fig-temporal_hazard-parametric-cumhaz
#| fig-cap: "Parametric Weibull cumulative hazard from the fitted model"
ggplot(pred, aes(x = time, y = cumhaz)) +
  geom_line(colour = "firebrick", linewidth = 0.8) +
  labs(x = "Years since operation", y = "Cumulative hazard",
       title = "Parametric (Weibull) cumulative hazard") +
  theme_hv_manuscript()
```

```{r}
#| label: fig-temporal_hazard-parametric-hazard
#| fig-cap: "Instantaneous hazard rate derived by differencing the parametric Weibull cumulative hazard"
ggplot(subset(pred, !is.na(hazard)), aes(x = time, y = hazard)) +
  geom_line(colour = "darkgreen", linewidth = 0.8) +
  labs(x = "Years since operation", y = "Instantaneous hazard rate",
       title = "Parametric (Weibull) hazard rate") +
  theme_hv_manuscript()
```

### Observed vs parametric

A parametric fit is only worth showing once you have checked it against the
empirical curve it is meant to summarise. So overlay them: the **observed**
Kaplan-Meier (or Nelson-Aalen) estimate is the yardstick, the **parametric**
Weibull is the smooth model laid over it. Two series on one panel means a legend,
and we place it inside the panel in the empty corner so it names the curves
without covering them.

```{r}
#| label: fig-temporal_hazard
#| fig-cap: "Observed Kaplan-Meier survival overlaid with the parametric Weibull curve as a fit check"
p_th_surv <- ggplot() +
  geom_ribbon(data = km,
              aes(x = time, ymin = cl_lower, ymax = cl_upper),
              alpha = 0.2, fill = "grey40") +
  geom_step(data = km,
            aes(x = time, y = survival, colour = "Observed (Kaplan-Meier)"),
            linewidth = 0.7) +
  geom_line(data = pred,
            aes(x = time, y = survival, colour = "Parametric (Weibull)"),
            linewidth = 0.9) +
  scale_colour_manual(
    name   = NULL,
    values = c("Observed (Kaplan-Meier)" = "grey40",
               "Parametric (Weibull)"    = "steelblue")
  ) +
  labs(x = "Years since operation", y = "Survival probability") +
  theme_hv_manuscript()

# Survival falls left-to-right, leaving the lower-left empty; hv_legend_inside()
# anchors the key there automatically.
hv_legend_inside(p_th_surv) +
  theme(legend.background = element_rect(fill = "white", colour = "grey80",
                                         linewidth = 0.3))
```

Read the overlay as a fit check: where the smooth Weibull tracks the step
function, the parametric form is describing the data; where the observed
Kaplan-Meier curve and its grey confidence band pull away from the Weibull line,
the single Weibull shape is the wrong summary and a multi-phase fit is the next
move. (The band is the observed estimate's; this parametric `predict()` returns
a point curve with no interval.) The same overlay works for cumulative hazard,
with the Nelson-Aalen estimate as the observed series.

```{r}
#| label: fig-temporal_hazard-compare-cumhaz
#| fig-cap: "Observed Nelson-Aalen cumulative hazard overlaid with the parametric Weibull curve"
p_th_cumhaz <- ggplot() +
  geom_step(data = na,
            aes(x = time, y = cumhaz, colour = "Observed (Nelson-Aalen)"),
            linewidth = 0.7) +
  geom_line(data = pred,
            aes(x = time, y = cumhaz, colour = "Parametric (Weibull)"),
            linewidth = 0.9) +
  scale_colour_manual(
    name   = NULL,
    values = c("Observed (Nelson-Aalen)" = "grey40",
               "Parametric (Weibull)"    = "firebrick")
  ) +
  labs(x = "Years since operation", y = "Cumulative hazard") +
  theme_hv_manuscript()

# Cumulative hazard rises left-to-right, leaving the upper-left empty.
hv_legend_inside(p_th_cumhaz) +
  theme(legend.background = element_rect(fill = "white", colour = "grey80",
                                         linewidth = 0.3))
```

### Numbers at risk

A survival figure should carry its denominators. The empirical curve here is a
plain data frame, with no `$tables$risk` to draw on, so we hand `hv_atrisk()` the
subject-level data directly — the `time` and `status` columns — and compose the
table under the curve, aligned to its ticks.

```{r}
#| label: fig-temporal_hazard-atrisk
#| fig-cap: "Kaplan-Meier survival with a numbers-at-risk table rebuilt from the subject-level data and composed beneath"
#| fig-height: 5
curve_km <- ggplot(km, aes(x = time, y = survival)) +
  geom_ribbon(aes(ymin = cl_lower, ymax = cl_upper),
              alpha = 0.2, fill = "steelblue") +
  geom_step(colour = "steelblue", linewidth = 0.7) +
  labs(x = "Years since operation", y = "Survival probability") +
  theme_hv_manuscript()

hv_atrisk_compose(
  curve_km,
  hv_atrisk(cabgkul, time = "int_dead", status = "dead")
)
```

For an `hv_survival()` object the at-risk table comes for free from
`$tables$risk` (see the survival chapter); for a curve built from raw columns
like this one, the subject-level path rebuilds it.

## Read it

A hazard curve is read by its shape, phase by phase. The Weibull above is the
simplest case: its shape parameter `nu < 1` produces a hazard that starts high
and falls monotonically, which is exactly the early-risk story of
post-operative mortality. Risk is greatest in the first months and then eases.

In the fuller multi-phase model the curve has three readable regions. An **early
phase** is the steep descent right after the operation, the perioperative risk
working itself out. A **constant phase** is the long flat middle, the background
rate of a stable cohort; a horizontal hazard there means risk is no longer
related to how long ago the surgery was. A **late phase** is an upturn years out,
typically the cohort aging or a prosthesis wearing. When you describe a figure
like this to a reviewer, name the phases: "an early peak that resolves by year
one, a flat interval, and a late rise after year ten." That is the language the
SAS-HAZARD lineage trained everyone to expect. Read the *cumulative* hazard
differently: it only ever rises, and it is its slope, not its height, that
carries the hazard. A flattening cumulative-hazard curve means low current risk
even though the total keeps climbing.

## Variations

### Competing risks

Death is not the only thing that can happen, and once a patient dies they can no
longer have a thromboembolic event. Treating one outcome as plain censoring for
the other overstates incidence; the cumulative incidence function is the honest
summary. `hzr_competing_risks()` takes an integer-coded `event` vector
(`0` = censored, `1, 2, ...` = competing event types) and returns a data frame
(class `hzr_competing_risks`) with `time`, `n_risk`, per-type `n_event_k`,
`n_censor`, the event-free `surv`, per-type cumulative incidence `incid_k`, and
standard errors.

The `omc` dataset records thromboembolic events (`te1`, `te2`, `te3`) and death
(`dead`). We collapse them into a two-cause outcome: type 1 is any
thromboembolic event, type 2 is death.

```{r}
#| label: cr-data
data(omc, package = "TemporalHazard")
event <- with(omc, ifelse(te1 == 1 | te2 == 1 | te3 == 1, 1L,
                          ifelse(dead == 1, 2L, 0L)))
table(event)

cr <- hzr_competing_risks(omc$int_dead, event)
head(cr)
```

Reshape the two incidence columns to long form and plot both curves together.
Read them as a pair: the two cumulative incidence curves climb toward a shared
ceiling, and the gap between them is the relative burden of each cause.

```{r}
#| label: fig-temporal_hazard-competing-risks
#| fig-cap: "Competing-risks cumulative incidence for thromboembolic events and death in the omc cohort"
cif <- data.frame(
  time      = rep(cr$time, 2),
  incidence = c(cr$incid_1, cr$incid_2),
  cause     = factor(rep(c("Thromboembolic", "Death"), each = nrow(cr)))
)

ggplot(cif, aes(x = time, y = incidence, colour = cause)) +
  geom_step(linewidth = 0.7) +
  labs(x = "Years since operation", y = "Cumulative incidence",
       colour = "Cause", title = "Competing-risks cumulative incidence (omc)") +
  theme_hv_manuscript()
```

### Diagnostics

A parametric fit is only worth reporting if it tracks the data. `hzr_gof()`
compares the fitted model against the Kaplan-Meier estimate. Call it on the fit
with no extra arguments: it populates the observed and expected event counts on
its own default grid (the observed event times), and handing it a custom grid
leaves those columns empty. The returned columns are `time`, `n_risk`,
`n_event`, `n_censor`, `km_surv`, `km_cumhaz`, `par_surv`, `par_cumhaz`,
`cum_observed`, and `cum_expected`, plus a `residual` (expected minus observed);
a `"summary"` attribute carries the scalar conservation diagnostics.

```{r}
#| label: gof
gof <- hzr_gof(fit)
attr(gof, "summary")
head(gof)
```

Overlaying the parametric survival on the Kaplan-Meier survival is the quickest
visual check. Where the two curves sit on top of each other the Weibull is
faithful; where they peel apart it is missing structure in the data.

```{r}
#| label: fig-temporal_hazard-gof
#| fig-cap: "Goodness-of-fit overlay of parametric and Kaplan-Meier survival on the same panel"
gof_long <- data.frame(
  time     = rep(gof$time, 2),
  survival = c(gof$km_surv, gof$par_surv),
  model    = factor(rep(c("Kaplan-Meier", "Parametric"), each = nrow(gof)))
)

ggplot(gof_long, aes(x = time, y = survival, colour = model)) +
  geom_line(linewidth = 0.6) +
  labs(x = "Years since operation", y = "Survival probability",
       colour = NULL, title = "Goodness of fit: observed vs. parametric") +
  theme_hv_manuscript()
```

The cumulative observed-versus-expected curve is the survival analogue of a
calibration plot. When the two lines stay together (a conservation ratio near 1)
the model is reproducing the actual event count over time, not just the shape of
the curve.

```{r}
#| label: fig-temporal_hazard-gof-oe
#| fig-cap: "Cumulative observed versus expected events, the survival analogue of a calibration plot"
oe <- data.frame(
  time   = rep(gof$time, 2),
  events = c(gof$cum_observed, gof$cum_expected),
  kind   = factor(rep(c("Observed", "Expected"), each = nrow(gof)))
)

ggplot(oe, aes(x = time, y = events, colour = kind)) +
  geom_line(linewidth = 0.6) +
  labs(x = "Years since operation", y = "Cumulative events",
       colour = NULL, title = "Cumulative observed vs. expected events") +
  theme_hv_manuscript()
```

Finally, `hzr_calibrate()` bins a continuous covariate into quantile groups and
reports the observed event probability per bin, returning `group`, `n`,
`events`, `mean`, `min`, `max`, `prob`, and `link_value` (the event probability
on the chosen link scale). Here we calibrate against follow-up time itself as a
worked example; against a real risk score the same plot tells you whether
predicted and observed risk agree across the range.

```{r}
#| label: fig-temporal_hazard-calibrate
#| fig-cap: "Observed event probability across follow-up-time deciles from hzr_calibrate()"
cal <- hzr_calibrate(cabgkul$int_dead, cabgkul$dead, groups = 10)
cal

ggplot(cal, aes(x = mean, y = prob)) +
  geom_point(size = 2, colour = "steelblue") +
  geom_line(colour = "steelblue") +
  labs(x = "Mean follow-up time in bin (years)", y = "Observed event probability",
       title = "Calibration across follow-up-time deciles") +
  theme_hv_manuscript()
```

## Pitfalls

- **`hazard()` needs starting values.** Pass `theta` explicitly, for example
  `theta = c(0.01, 1)`. Without it the optimiser has no starting point and the
  fit (and every `predict()` that follows) errors out. This is the single most
  common reason a `TemporalHazard` script fails on first run.
- **The instantaneous hazard is not free.** For an intercept-only model
  `predict(type = "hazard")` returns the constant relative hazard, not the curve
  you want. Derive the instantaneous rate by differencing the cumulative hazard,
  as in the `predict` chunk above; the hazard is its slope.
- **Let `hzr_gof()` choose its own grid.** It fills the observed and expected
  event counts only on its default event-time grid. Hand it a custom grid and
  `cum_observed`/`cum_expected` come back empty, and the calibration plots have
  nothing to draw. Call `hzr_gof(fit)` with no grid argument.
- **A pretty fit is not a faithful one.** Always run the goodness-of-fit overlay
  before you quote a parametric hazard. If the Weibull curve drifts off the
  Kaplan-Meier line, the smooth shape is hiding a misfit, and a richer
  multi-phase specification is the right next step.
