# Themes for consistency

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

hvtiPlotR [@R-hvtiPlotR] ships output-target themes via `theme_hv_*()`. Each sets a complete
non-data formatting baseline; the constructor and `plot()` add none of it. Pick
the theme that matches where the figure lands.

| Theme | Target |
|-------|--------|
| `theme_hv_manuscript()` | Journal PDF, black-on-white |
| `theme_hv_poster()` | Conference poster, larger text |
| `theme_hv_ppt_light()` | PowerPoint on white/light background |
| `theme_hv_ppt_dark()` | PowerPoint on dark/navy background |

We render the same base trends plot under each theme, then show targeted
overrides. The shared decorations are factored out once.

```{r}
#| label: base
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))

decorate <- function(p) {
  p +
    scale_colour_brewer(palette = "Set1", name = "Group") +
    scale_shape_manual(
      values = c("Group I" = 15, "Group II" = 19,
                 "Group III" = 17, "Group IV" = 18),
      name = "Group"
    ) +
    labs(x = "Surgery Year", y = "Outcome")
}
```

## Manuscript

Default for journal submissions — white background, black text, minimal chrome.
Font sizes tuned for letter-size PDFs reduced to a column width.

```{r}
#| label: manuscript
decorate(p_base) + theme_hv_manuscript()
```

## Poster

Larger axis text and tick weights, panel grid removed — reads at arm's length
on a foam board, and holds up on a projector.

```{r}
#| label: poster
decorate(p_base) + theme_hv_poster()
```

## Light PowerPoint

Dark text and lines for white or light-grey slide backgrounds — the Cleveland
Clinic standard template and most default Office decks.

```{r}
#| label: ppt_light
decorate(p_base) + theme_hv_ppt_light()
```

## Dark PowerPoint

White-on-dark for navy or dark-blue gradient decks. The theme sets a
transparent plot background so the slide shows through; the `plot.background`
override below simulates that navy in this rendered book and would be omitted
when saving to an actual `.pptx`.

```{r}
#| label: ppt_dark
decorate(p_base) +
  theme_hv_ppt_dark() +
  theme(plot.background = element_rect(fill = "navy", colour = "navy"))
```

## Theme overrides

Each theme is a baseline. Layer additional `theme()` calls after it to adjust
individual elements without touching the rest. The same overrides can also be
passed directly through the theme's `...`, e.g.
`theme_hv_poster(legend.position = "bottom")`.

### Axis text size

Resize tick labels (`axis.text`) and axis titles (`axis.title`) independently —
useful when scaling a figure down for a two-column layout.

```{r}
#| label: axis_text
decorate(p_base) +
  theme_hv_poster() +
  theme(
    axis.text  = element_text(size = 10),
    axis.title = element_text(size = 12)
  )
```

### Removing minor grid lines

Minor grid lines add density without precision. Blank them while keeping the
major grid.

```{r}
#| label: grid
decorate(p_base) +
  theme_hv_poster() +
  theme(panel.grid.minor = element_blank())
```

### Rotating x-axis labels

For time-point or long category labels. `angle = 45` with `hjust = 1, vjust = 1`
keeps the rotated label right-aligned to its tick.

```{r}
#| label: rotate
decorate(p_base) +
  theme_hv_poster() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

### Plot title and subtitle

Titles are stripped from the base themes (rarely used in journal figures) but
can be added back via `labs()` and styled in `theme()`.

```{r}
#| label: title
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)

plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title    = "Overall Survival",
    subtitle = paste0("n = ", nrow(dta_km), " patients"),
    x        = "Years after Operation",
    y        = "Freedom from Death (%)"
  ) +
  theme_hv_poster() +
  theme(
    plot.title    = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 11, colour = "grey40", hjust = 0)
  )
```

### Expanding plot margins

Add breathing room around the panel — for a figure placed directly on a poster,
or when labels clip a tight device edge. Arguments follow top / right / bottom /
left order.

```{r}
#| label: margin
decorate(p_base) +
  theme_hv_poster() +
  theme(plot.margin = margin(t = 10, r = 20, b = 10, l = 20, unit = "pt"))
```
