# Venn diagrams and UpSet plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

To show non-distinct groups and understand overlapping groupings, we turn to VENN diagrams, and UpSet plots. The general rule to show no more than four groups in VENN diagrams. If there are more than four groups, we look at UpSet Plots.


## VENN diagrams


## Upset plots

`hv_upset()`, from hvtiPlotR [@R-hvtiPlotR], builds an UpSet diagram via `ggupset::scale_x_upset()` to
visualise surgical procedure co-occurrences or any set-membership data. Where a
Venn diagram breaks down past three or four sets, UpSet scales cleanly to seven
or more.

`plot.hv_upset()` returns a **ggplot** when `set_size = FALSE`, or a
**patchwork composite** of an intersection-bar plot plus a set-size sidebar
when `set_size = TRUE` (the default). For the bare ggplot path, themes apply
via `+`; for the patchwork path, use patchwork's `&` operator to theme every
sub-panel.

`sample_upset_data()` returns a binary indicator matrix: one column per
procedure, one row per patient, with 1 indicating the procedure was performed.
Pass the column names to `intersect` to define the set-membership axes.

```{r}
#| label: upset_data
sets <- c("AV_Replacement", "AV_Repair", "MV_Replacement", "MV_Repair",
          "TV_Repair", "Aorta", "CABG")

upset_dta <- sample_upset_data(n = 400, seed = 42)
head(upset_dta)
colSums(upset_dta)

hu <- hv_upset(upset_dta, intersect = sets)
```

### Default plot

Intersection bars (top 10 by frequency) with a set-size sidebar on the right.
`set_size = TRUE` returns a patchwork composite, so theme it with `&` to cover
both sub-panels:

```{r}
#| label: upset_basic
#| fig-width: 9
#| fig-height: 5
plot(hu) &
  theme_hv_poster()
```

### Custom intersection bar colour

The intersection bars are a standard `geom_bar()` — change the colour via the
`bar_fill` argument. Setting `set_size = FALSE` returns just the
intersection-bar ggplot for full `+` customisation.

```{r}
#| label: upset_fill
#| fig-width: 9
#| fig-height: 5
plot(hu, bar_fill = "steelblue", set_size = FALSE) +
  ggplot2::labs(y = "Patients (n)") +
  theme_hv_poster()
```

### Colour bars by era

Pass `fill_col` to fill the intersection bars by a grouping column on the input
data, combined with `scale_fill_manual()` for explicit colour assignment.

```{r}
#| label: upset_era
#| fig-width: 9
#| fig-height: 5
upset_dta$era <- ifelse(seq_len(nrow(upset_dta)) <= 200, "Early", "Recent")
hu_era <- hv_upset(upset_dta, intersect = sets)

plot(hu_era, fill_col = "era", set_size = FALSE) +
  ggplot2::scale_fill_manual(
    values = c("Early" = "grey60", "Recent" = "steelblue"),
    name   = "Era"
  ) +
  ggplot2::labs(y = "Patients (n)") +
  theme_hv_poster()
```