# Variable priority with varPro

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(varPro)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

Permutation importance (the VIMP chapter) answers "how much does the forest lean
on this variable?" by breaking the variable and watching error rise. `varPro`
[@R-varPro] asks a different question: "how much do the forest's own splitting
rules *depend* on this variable?" Rather than permuting predictors, it scores
each one by how much a tree's release rules rely on it, and calls the result
variable *priority*. Reach for it when you want a selection that comes with a
built-in cutoff (so the method tells you which variables to keep, not just how to
rank them), when you want the direction and magnitude of a local effect, or when
you want to flag unusual observations the same forest can isolate.

`ggRandomForests` [@R-ggRandomForests] provides a family of `gg_*` extractors
that pull these scores out of a fitted `varpro` object and render them as tidy
ggplot2 graphics. This chapter works through the main extractors on a regression
fit and a classification fit. As elsewhere, each `plot()` hands back a bare
ggplot you finish with a house theme.

## The data it needs

The extractors read from a fitted `varpro` object, so the input is whatever
`varpro()` itself needs: a formula and a data frame. The outcome type decides
the family, and that matters here because several of the extractors are
family-specific. We start with a regression fit, modelling fuel economy (`mpg`)
on the rest of `mtcars`. A small `ntree` keeps the example fast; in practice you
would use the default.

```{r}
#| label: varpro-fit
set.seed(42)
o <- varpro(mpg ~ ., data = mtcars, ntree = 100)
plot(gg_varpro(o)) + theme_hv_manuscript()
```

`gg_varpro()` returns an honest 15/85 boxplot for each variable. The box spans
the importance values obtained when the variable is released across the rule
ensemble, and the dashed reference line at the `cutoff` (default 0.79 of the top
score on the local-standardized scale) separates selected from unselected
variables. Variables whose boxes sit to the right of the cutoff are retained.

## Build it

The plot above is the core figure. Two variants make it more informative.
Setting `faithful = TRUE` shows the per-tree importance draws as jittered points
behind the boxplot, exposing the spread that the summary hides:

```{r}
#| label: varpro-faithful
plot(gg_varpro(o, faithful = TRUE)) + theme_hv_manuscript()
```

Setting `nvar` restricts the display to the top-ranked variables, handy when a
wide problem would otherwise crowd the panel:

```{r}
#| label: varpro-nvar
plot(gg_varpro(o, nvar = 5)) + theme_hv_manuscript()
```

## Read it

The 15/85 boxplot is the figure to learn to read, and the cutoff line is what
makes it a *selection* rather than just a ranking.

- **The cutoff line decides membership.** Boxes to the right of the dashed line
  are the variables varPro keeps; boxes to the left are set aside. Unlike a VIMP
  bar chart, where you eyeball the break yourself, the method draws the line for
  you. You can move it with the `cutoff` argument when you want to be stricter or
  looser.
- **Box width is stability.** A tight box means the variable scored consistently
  across the rule ensemble; a wide box straddling the cutoff is a variable whose
  selection you should not bet the paper on. Turn on `faithful = TRUE` to see the
  raw draws behind the summary.
- **Priority is not the same as importance.** A variable can carry high priority
  (the rules depend on it) without the large error jump that permutation VIMP
  rewards, and vice versa. Read the two methods as complementary views, not as
  two attempts at the same number.

## Variations

### Beta-refined importance (regression only)

`gg_beta_varpro()` is regression-only. Within each release region it fits a
lasso and reports the refined slope (β) coefficients instead of a VIMP-style
importance, so you get not just *whether* a variable matters but the direction
and magnitude of its local linear effect.

```{r}
#| label: beta-varpro
plot(gg_beta_varpro(o)) + theme_hv_manuscript()
```

The per-rule lasso refinement (`beta.varpro`) is the expensive step; under
Quarto's `freeze` the result is cached after the first render. Because these are
lasso-refined β coefficients rather than importance scores, they can be negative.

### Individual variable priority

`gg_ivarpro()` computes *individual* (observation-level) variable priority,
decomposing the forest importance down to each case. On a regression fit with
weak per-observation signal the decomposition can collapse to nothing (it
returns zero rows), so we demonstrate it on a classification forest, which also
lets us show the `which_class` argument. We fit `Species ~ .` on `iris`:

```{r}
#| label: ivarpro-fit
set.seed(42)
oc <- varpro(Species ~ ., data = iris, ntree = 100)
plot(gg_ivarpro(oc)) + theme_hv_manuscript()
```

Each panel summarises how strongly individual observations rely on a given
variable. For multiclass problems you can focus on a single outcome level with
`which_class`:

```{r}
#| label: ivarpro-class
plot(gg_ivarpro(oc, which_class = "setosa")) + theme_hv_manuscript()
```

### Conditional priority by class

The classification fit also supports `conditional = TRUE` in `gg_varpro()`,
which facets the priority bars by outcome class so you can read which predictors
separate which classes. This option needs a classification forest; it errors on
the regression fit.

```{r}
#| label: varpro-conditional
plot(gg_varpro(oc, conditional = TRUE)) + theme_hv_manuscript()
```

### Anomaly / isolation scores

`gg_isopro()` visualises in-sample anomaly (isolation) scores. These come from
an `isopro` object, and there is no `.varpro` method that builds one for you: you
construct it first with `varPro::isopro()` on the fitted forest, then pass that
to `gg_isopro()`.

```{r}
#| label: isopro
iso <- isopro(object = o, ntree = 100)
plot(gg_isopro(iso)) + theme_hv_manuscript()
```

Higher isolation scores flag observations the forest separates with few splits,
candidate outliers. Pass `newdata` to `gg_isopro()` to score held-out
observations against the fitted isolation forest.

### Partial dependence curves

The extractors so far rank and refine variables; partial dependence asks a
different question — *how does the prediction move as one variable changes,
holding the rest as they are?* `gg_partial_varpro()` runs `varPro::partialpro()`
on the fitted forest and tidies the result into marginal-effect curves: one
panel per variable, the predicted response across that variable's range. On the
regression fit the y-axis is on the response scale (`mpg`).

```{r}
#| label: partial-varpro
plot(gg_partial_varpro(object = o)) & theme_hv_manuscript()
```

`plot()` returns a panel of curves (a `patchwork`), so we finish it with `&`
rather than `+` to apply the house theme to every panel. Each curve shows the
modelled effect of one predictor; a flat curve is a variable the forest barely
uses, a steep one a variable that moves the prediction. `gg_partialpro()` is the
superseded name for this extractor in older code — it is now a thin alias for
`gg_partial_varpro()`, so reach for the new name.

## Unsupervised variable dependence

Everything above needs an outcome. `varPro::uvarpro()` drops that requirement:
it fits an *unsupervised* forest and scores how strongly each variable depends
on the others, with no response to predict. `gg_udependent()` reads those
cross-variable scores and draws them as a dependency graph — nodes are
variables, edges the dependencies between them — so you can see which predictors
travel together before any modelling. Reach for it when you want to understand
the structure among the predictors themselves, for example to spot a cluster of
collinear measurements.

```{r}
#| label: udependent-fit
set.seed(42)
uv <- uvarpro(iris[, -5], ntree = 100)
plot(gg_udependent(uv)) + theme_hv_manuscript()
```

We fit it on the four `iris` measurements with the species label dropped, since
the method takes no outcome. A thick edge between two variables means the forest
finds them strongly interdependent; an isolated node is a variable that carries
information the others do not.

## Pitfalls

- **`gg_isopro()` needs an `isopro` object first.** There is no `.varpro` method,
  so calling `gg_isopro()` on the `varpro` object directly fails. Build
  `iso <- isopro(object = o, ...)` and then plot `iso`.
- **`gg_beta_varpro()` is regression-only.** The lasso β refinement is defined
  for a regression fit; do not expect it off a classification `varpro` object.
- **`conditional = TRUE` needs a classification forest.** `gg_varpro(conditional
  = TRUE)` facets priority by class, so it errors on a regression fit like the
  `mtcars` model. Use it on a classification fit such as the `iris` one.
- **`gg_ivarpro()` needs real signal.** Individual priority decomposes importance
  to each observation, and on a weak regression problem (mtcars) that
  decomposition can return zero rows and an empty plot. It works on a
  strong-signal dataset like the `iris` classification fit, which is why the
  individual-priority examples use that forest.
