# Variable priority with varPro

`varPro` [@R-varPro] ranks variables by *priority* using rule-based importance: rather than permuting predictors as in classical VIMP, it scores each variable by how much a tree's release rules depend on it. `ggRandomForests` [@R-ggRandomForests] provides a family of `gg_*` extractors that pull these scores out of a fitted `varpro` object and render them as tidy `ggplot2` graphics. This chapter works through the main extractors on a regression fit and a classification fit.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(varPro)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## Variable priority

We fit a regression forest with `varpro()`. A small `ntree` keeps the example fast; in practice you would use the default. We model fuel economy (`mpg`) on the remaining `mtcars` predictors.

```{r}
#| label: varpro-fit
set.seed(42)
o <- varpro(mpg ~ ., data = mtcars, ntree = 100)
plot(gg_varpro(o)) + theme_hv_manuscript()
```

`gg_varpro()` returns an *honest* 15/85 boxplot for each variable: the box spans the importance values obtained when the variable is released across the rule ensemble, and the dashed reference line at the `cutoff` (default 0.79 of the top score on the local-standardized scale) separates selected from unselected variables. Variables whose boxes sit to the right of the cutoff are retained.

Two useful variants. Setting `faithful = TRUE` shows the per-tree importance draws as jittered points behind the boxplot, exposing the spread that the summary hides:

```{r}
#| label: varpro-faithful
plot(gg_varpro(o, faithful = TRUE)) + theme_hv_manuscript()
```

Setting `nvar` restricts the display to the top-ranked variables, which is handy for wide problems:

```{r}
#| label: varpro-nvar
plot(gg_varpro(o, nvar = 5)) + theme_hv_manuscript()
```

## Beta-refined importance (regression)

`gg_beta_varpro()` is **regression-only**. Within each release region it fits a lasso and reports the refined slope (β) coefficients rather than a VIMP-style importance. This tells you not just *whether* a variable matters but the direction and magnitude of its local linear effect.

```{r}
#| label: beta-varpro
plot(gg_beta_varpro(o)) + theme_hv_manuscript()
```

The per-rule lasso refinement (`beta.varpro`) is the expensive step; under Quarto's `freeze` the result is cached after the first render. Note these are lasso-refined β coefficients, not importance scores, so they can be negative.

## Individual variable priority

`gg_ivarpro()` computes *individual* (observation-level) variable priority, decomposing the forest importance down to each case. On a regression fit with weak per-observation signal the decomposition can collapse to nothing, so we demonstrate it on a classification forest, which also lets us show the `which_class` argument. We fit `Species ~ .` on `iris`:

```{r}
#| label: ivarpro-fit
set.seed(42)
oc <- varpro(Species ~ ., data = iris, ntree = 100)
plot(gg_ivarpro(oc)) + theme_hv_manuscript()
```

Each panel summarises how strongly individual observations rely on a given variable. For multiclass problems you can focus on a single outcome level with `which_class`:

```{r}
#| label: ivarpro-class
plot(gg_ivarpro(oc, which_class = "setosa")) + theme_hv_manuscript()
```

The classification fit also supports `conditional = TRUE` in `gg_varpro()`, which facets the priority bars by outcome class so you can read which predictors separate which classes:

```{r}
#| label: varpro-conditional
plot(gg_varpro(oc, conditional = TRUE)) + theme_hv_manuscript()
```

## Anomaly / isolation scores

`gg_isopro()` visualises in-sample anomaly (isolation) scores. These come from an `isopro` object, which is built from a fitted `varpro` forest with `varPro::isopro()` rather than from the `varpro` object directly:

```{r}
#| label: isopro
iso <- isopro(object = o, ntree = 100)
plot(gg_isopro(iso)) + theme_hv_manuscript()
```

Higher isolation scores flag observations that the forest separates with few splits — candidate outliers. Pass `newdata` to `gg_isopro()` to score held-out observations against the fitted isolation forest.
